### 2020-05-26 latex/paper.tex
seeded latex with sample reference

\documentclass[conference,compsoc]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.

\usepackage{cite,doi}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization
\thanks{Identify applicable funding agency here. If none, delete this.}
}

\author{\IEEEauthorblockN{Lawton Manning, John Farrell, Grey Ballard}
\IEEEauthorblockA{Wake Forest University \\
Winston-Salem, NC, USA \\
\{mannlg15,farrjt17,ballard\}@wfu.edu}
\and
\IEEEauthorblockN{Ramakrishnan Kannan}
\IEEEauthorblockA{Oak Ridge National Laboratory \\
Oak Ridge, TN, USA \\
kannanr@ornl.gov}
\and
\IEEEauthorblockN{Haesun Park}
\IEEEauthorblockA{Georgia Institute of Technology \\
Atlanta, GA, USA \\
hpark@cc.gatech.edu}
}

\maketitle

\begin{abstract}

\end{abstract}

%\begin{IEEEkeywords}
%component, formatting, style, styling, insert
%\end{IEEEkeywords}

\section{Introduction}

test reference: \cite{GKP15}

%\section*{Acknowledgment}

\bibliography{paper}
\bibliographystyle{plainurl}

\end{document}

### 2020-06-08 distnmf/aunmf.hpp
tolerance fix

@@ -7,6 +7,7 @@
 #include <armadillo>
 #include <string>
 #include <vector>
+#include <cmath>
 #include "distnmf/distnmf.hpp"
 #include "distnmf/mpicomm.hpp"
 
@@ -458,7 +459,8 @@ class DistAUNMF : public DistNMF<INPUTMATTYPE> {
 #ifdef __WITH__BARRIER__TIMING__
     MPI_Barrier(MPI_COMM_WORLD);
 #endif
-    double relerr, old_relerr;
+    double relerr;
+    double old_relerr = 0.0;
     for (unsigned int iter = 0; iter < this->num_iterations(); iter++) {
       // saving current instance for error computation.
       if (iter > 0 && this->is_compute_error()) {
@@ -555,7 +557,9 @@ class DistAUNMF : public DistNMF<INPUTMATTYPE> {
                         << this->k << "::err::" << sqrt(this->objective_err)
                         << "::relerr::"
                         << relerr);
-        if (abs(relerr-old_relerr) < this->m_tolerance) {
+        printf("tolerance:%f  actual:%f\n",this->m_tolerance,std::abs(relerr-old_relerr));
+        if (std::abs(relerr-old_relerr) < this->m_tolerance) {
+          printf("breaking\n");
           break;
         }
       }

### 2020-06-08 distnmf/aunmf.hpp
removed tolerance print

@@ -557,7 +557,6 @@ class DistAUNMF : public DistNMF<INPUTMATTYPE> {
                         << this->k << "::err::" << sqrt(this->objective_err)
                         << "::relerr::"
                         << relerr);
-        printf("tolerance:%f  actual:%f\n",this->m_tolerance,std::abs(relerr-old_relerr));
         if (std::abs(relerr-old_relerr) < this->m_tolerance) {
           printf("breaking\n");
           break;

### 2020-06-08 common/parsecommandline.hpp
tolerance on printconfig

@@ -196,7 +196,8 @@ class ParseCommandLine {
               << "::k::" << this->m_k << "::m::" << this->m_globalm
               << "::n::" << this->m_globaln << "::t::" << this->m_num_it
               << "::pr::" << this->m_pr << "::pc::" << this->m_pc
-              << "::error::" << this->m_compute_error << "::regW::"
+              << "::error::" << this->m_compute_error  << "::tol::" << this->m_tolerance
+              << "::regW::"
               << "l2::" << this->m_regW(0) << "::l1::" << this->m_regW(1)
               << "::regH::"
               << "l2::" << this->m_regH(0) << "::l1::" << this->m_regH(1)

### 2020-06-08 hiernmf/hiernmf.cpp
number of leaves fix

@@ -126,7 +126,10 @@ class HierNMFDriver {
       this->root->enqueue(nodes);
 
 
-      while (!frontiers.empty() && nodes.size() <= pc->nodes()) {
+      for (int i = 2; i < pc->nodes(); i++) {
+        if (frontiers.empty()) {
+          break;
+        }
         frontier = frontiers.top();
         frontiers.pop();
         frontier->accept();

### 2020-06-08 hiernmf/node.hpp
random seed gen

@@ -143,6 +143,7 @@ namespace planc {
         void split() {
           this->accepted = true;
 
+          arma::arma_rng::set_seed_random();
           MAT W = arma::randu<MAT>(itersplit(A.n_rows,pc->pc(),mpicomm->col_rank()),2);
           MAT H = arma::randu<MAT>(itersplit(A.n_cols,pc->pr(),mpicomm->row_rank()),2);
 

### 2020-06-08 distnmf/aunmf.hpp
removed breaking message

@@ -558,7 +558,6 @@ class DistAUNMF : public DistNMF<INPUTMATTYPE> {
                         << "::relerr::"
                         << relerr);
         if (std::abs(relerr-old_relerr) < this->m_tolerance) {
-          printf("breaking\n");
           break;
         }
       }

### 2020-06-08 latex/paper.tex
set up outline for paper

@@
 
+\newcommand{\LM}[1]{\textcolor{blue}{\textbf{Lawton}: #1}}    
+\newcommand{\GB}[1]{\textcolor{red}{\textbf{Grey}: #1}}
+\newcommand{\RK}[1]{\textcolor{blue}{\textbf{Ramki}: #1}}
+
 \begin{document}
@@
 \title{Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization
-\thanks{Identify applicable funding agency here. If none, delete this.}
+\thanks{This material is based upon work supported by the National Science Foundation under Grant No. OAC-1642385 and OAC-1642410.
+This manuscript has been co-authored by UT-Battelle, LLC under Contract No. DE-AC05-00OR22725 with the U.S. Department of
+Energy. This project was partially funded by the Laboratory Director's Research and Development fund. This research used resources
+of the Oak Ridge Leadership Computing Facility at the Oak Ridge National Laboratory, which is supported by the Office of Science of
+the U.S. Department of Energy.}
 }
 
-\author{\IEEEauthorblockN{Lawton Manning, John Farrell, Grey Ballard}
+\author{\IEEEauthorblockN{Lawton Manning and Grey Ballard}
 \IEEEauthorblockA{Wake Forest University \\
 Winston-Salem, NC, USA \\
-\{mannlg15,farrjt17,ballard\}@wfu.edu}
+\{mannlg15,ballard\}@wfu.edu}
 \and
@@
 \begin{abstract}
-
+\GB{to be done...}
 \end{abstract}
@@
 
-%\section*{Acknowledgment}
+\section{Preliminaries}
+
+\section{Algorithm}
+
+\subsection{Sequential Algorithm}
+
+\subsubsection{Rank-2 NMF}
+
+\subsubsection{Hierarchical Clustering}
+
+\subsection{Parallelization}
+
+\subsubsection{Algorithm}
+
+\subsubsection{Analysis}
+
+\section{Experimental Results}
+
+\subsection{Experimental Platform}
+
+\subsection{Datasets}
+
+\subsection{Performance}
+
+\subsubsection{Weak Scaling}
+
+\subsubsection{Strong Scaling}
+
+\subsection{Clustering}
+
+\subsubsection{Hyperspectral Imaging}
+
+\subsubsection{Document Corpus}
+
+\section{Conclusion}
+
+\section*{Acknowledgment}
+
+The authors would like to thank Simin Ma for contributions to the algorithmic analysis and John Farrell for his contributions to the implementation of the parallel algorithm.
 

### 2020-06-09 hiernmf/hiernmf.cpp
root node to nodes queue

@@ -120,6 +120,8 @@ class HierNMFDriver {
       this->root = new RootNode<MAT>(A, this->m_globalm, this->m_globaln, cols, this->mpicomm, this->pc);
 #endif
 
+      nodes.push(root);
+
       this->root->split();
       this->root->accept();
       this->root->enqueue(frontiers);
@@ -149,14 +151,19 @@ class HierNMFDriver {
         printf("idx NMF sigma top_words\n");
         while (!nodes.empty()) {
           node = nodes.front();
+          nodes.pop();
+          
           printf("%d %f %f %f\n",node->index,node->timings.NMF,node->timings.sigma,node->timings.top_words);
+          
+          if (node->index == 0) {
+            continue;
+          }
           ofs << node->index << " ";
 #ifdef BUILD_SPARSE
           ofs << node->top_words.t();
 #else
           ofs << node->cols.t();
 #endif
-          nodes.pop();
         }
 
         ofs.close();

### 2020-06-09 hiernmf/node.hpp
delete local A after NMF

@@ -181,6 +181,8 @@ namespace planc {
           this->lvalid = !lcols.is_empty();
           this->rvalid = !rcols.is_empty();
           
+          A.clear();
+          
           if (this->lvalid) {
             this->lchild = new Node(this->A0, W.col(0), lcols, this, 2*this->index+1);
           }
@@ -240,7 +242,7 @@ namespace planc {
           this->parent = NULL;
           this->mpicomm = mpicomm;
           this->pc = pc;
-          this->A = this->A0;
+          this->A = INPUTMATTYPE(A);
           this->sigma = 0.0;
           this->index = 0;
           this->lvalid = false;

### 2020-06-11 hiernmf/node.hpp
reset and clear A

@@ -182,6 +182,7 @@ namespace planc {
           this->rvalid = !rcols.is_empty();
           
           A.clear();
+          A.reset();
           
           if (this->lvalid) {
             this->lchild = new Node(this->A0, W.col(0), lcols, this, 2*this->index+1);

### 2020-06-22 utilities/shuffle.cpp
removed shuffle

@@ -1,308 +1,0 @@
-#include <cstdio>
-#include <cstdlib>
-#include <vector>
-#include <ctime>
-#include <string>
-#include <iostream>
-#include <cinttypes>
-#include <cmath>
-
-/**
- * The dimension a particular rank holds out of 
- * the global dimension n across p processes.
- * @param[in] n is the global size of the tensor on that dimension
- * @param[in] p is the number of splits of n. 
- *            Typically number of processes on a particular mode
- * @param[in] r is the rank of the mpi process in that mode. fiber rank
- */
-inline uint64_t itersplit(uint64_t n, int p, int r) {
-  int split = (r < n % p) ? n / p + 1 : n / p;
-  return split;
-}
-
-/**
- * Returns the start idx of the current rank r
- * for a global dimension n across p processes. 
- * @param[in] n is the size of the tensor on that dimension
- * @param[in] p is the number of splits of n.
- * @param[in] r is the rank of the mpi process in that mode. fiber rank
- */
-inline uint64_t startidx(uint64_t n, int p, int r) {
-  int rem = n % p;
-  int idx =
-      (r < rem) ? r * (n / p + 1) : (rem * (n / p + 1) + ((r - rem) * (n / p)));
-  return idx;
-}
-
-/**
- * Returns the rank of the processor group
- * that owns the value at entry idx i for a
- * global dimension n across p processes.
- * @param[in] n is the size of the tensor on that dimension
- * @param[in] p is the number of splits of n
- * @param[in] i is the idx of the entry along that dimension
- */
-inline int idxproc(uint64_t n, int p, uint64_t i) {
- int rem = n % p;
- int idx = 
-     (i < rem * (n / p + 1)) ? i / (n / p + 1) : rem + (i - rem * (n / p + 1))/(n / p);
- return idx;
-}
-
-uint64_t range(uint64_t x[], int n) {
-  uint64_t max = x[0];
-  uint64_t min = x[0];
-  for (int i = 1; i < n; i++) {
-    if (x[i] > max) {
-      max = x[i];
-    }
-    if (x[i] < min) {
-      min = x[i];
-    }
-  }
-  return max - min;
[…]

### 2020-06-22 utilities/partition-dense.cpp
partitionining hyperspectral files

@@ -1,0 +1,84 @@
+#include <cstdio>
+#include <cstdlib>
+#include <vector>
+#include <ctime>
+#include <string>
+#include <iostream>
+#include <cinttypes>
+#include <armadillo>
+
+/**
+ * The dimension a particular rank holds out of 
+ * the global dimension n across p processes.
+ * @param[in] n is the global size of the tensor on that dimension
+ * @param[in] p is the number of splits of n. 
+ *            Typically number of processes on a particular mode
+ * @param[in] r is the rank of the mpi process in that mode. fiber rank
+ */
+inline uint64_t itersplit(uint64_t n, int p, int r) {
+  int split = (r < n % p) ? n / p + 1 : n / p;
+  return split;
+}
+
+/**
+ * Returns the start idx of the current rank r
+ * for a global dimension n across p processes. 
+ * @param[in] n is the size of the tensor on that dimension
+ * @param[in] p is the number of splits of n.
+ * @param[in] r is the rank of the mpi process in that mode. fiber rank
+ */
+inline uint64_t startidx(uint64_t n, int p, int r) {
+  int rem = n % p;
+  int idx =
+      (r < rem) ? r * (n / p + 1) : (rem * (n / p + 1) + ((r - rem) * (n / p)));
+  return idx;
+}
+
+/**
+ * Returns the rank of the processor group
+ * that owns the value at entry idx i for a
+ * global dimension n across p processes.
+ * @param[in] n is the size of the tensor on that dimension
+ * @param[in] p is the number of splits of n
+ * @param[in] i is the idx of the entry along that dimension
+ */
+inline int idxproc(int n, int p, int i) {
+ int rem = n % p;
+ int idx = 
+     (i < rem * (n / p + 1)) ? i / (n / p + 1) : rem + (i - rem * (n / p + 1))/(n / p);
+ return idx;
+}
+
+
+
+int main(int argc, char **argv) {
+  printf("partition-dense began.\n");
+  if (argc < 4) {
+    printf("Usage: partition-dense [matrix-file-name] [row-proc-count] [col-proc-count]\n");
+    return 0;
+  }
+
+  arma::mat X;
+  X.load(argv[1]);
+  int pr = atoi(argv[2]);
+  int pc = atoi(argv[3]);
+  uint64_t m = X.n_rows;
+  uint64_t n = X.n_cols;
+
[…]

### 2020-06-29 build-summit/hier.py
reuters scaling data

@@ -10,8 +10,6 @@ totals = {};
 
 count = 0
 
-ms = []
-ns = []
 begin = True
 end = False
 nodes = []
@@ -27,27 +25,20 @@ for line in open(name):
       totals[cat].append(times)
     begin = True
   
-  size = match(r".*globalm::(\d+)::globaln::(\d+)",line)
-  if size and begin:
-    ms.append(size.group(1))
-    ns.append(size.group(2))
-    count += 1
-    begin = False
-
   if end:
     nodes.append(line.split())
+    count = count+1
 
-  idx = match(r"^idx*",line)
+  idx = match(r"^idx.*",line)
   if idx:
     nodeline = line.split()
     end = True
 
 
-
 with open(name + '-nodes', 'w') as n, open(name + '-levels', 'w') as l:
   nwrite = writer(n,delimiter=',')
   lwrite = writer(l,delimiter=',')
-  first = nodeline + ['m','n']
+  first = nodeline
   for total in totals.keys():
     first.append(total + '-min')
     first.append(total + '-avg')
@@ -56,12 +47,12 @@ with open(name + '-nodes', 'w') as n, open(name + '-levels', 'w') as l:
   first[0] = "level"
   lwrite.writerow(first)
   
-  levels = [floor(log2(int(node[0])+1)) for node in nodes]
+  levels = [int(node[1]) for node in nodes]
   nodes_per_level = [levels.count(level) for level in list(set(levels))]
   full_levels = [count == 2**idx for idx,count in enumerate(nodes_per_level)]
   lrows = [[0.0]*len(first)]*(max(levels)+1)
   for idx in range(count):
-    line = nodes[idx] + [ms[idx],ns[idx]]
+    line = nodes[idx]
     for total,times in totals.items():
       line = line + times[idx]
     nwrite.writerow(line)

### 2020-06-29 hiernmf/hiernmf.cpp

@@ -158,12 +158,12 @@ class HierNMFDriver {
       if (this->mpicomm->rank() == 0) {
         std::ofstream ofs(this->m_outputfile_name, std::ofstream::out);
         
-        printf("idx NMF MATVEC(SIGMA) VECMAT(SIGMA) COMM(SIGMA) NORM(SIGMA)\n");
+        printf("idx level m n NMF MATVEC(SIGMA) VECMAT(SIGMA) COMM(SIGMA) NORM(SIGMA)\n");
         while (!nodes.empty()) {
           node = nodes.front();
           nodes.pop();
           
-          printf("%d %f %f %f %f %f\n",node->index,node->timings.NMF,node->timings.sigma.matvec,node->timings.sigma.vecmat,node->timings.sigma.communication,node->timings.sigma.normalisation);
+          printf("%d %d %llu %llu %f %f %f %f %f\n",node->index,node->level,node->global_m,node->global_n,node->timings.NMF,node->timings.sigma.matvec,node->timings.sigma.vecmat,node->timings.sigma.communication,node->timings.sigma.normalisation);
           
           if (node->index == 0) {
             continue;

### 2020-06-29 hiernmf/node.hpp

@@ -25,8 +25,9 @@ namespace planc {
         UVEC cols;
         bool accepted = false;
         int index;
+        int level;
 
-        int global_m, global_n;
+        unsigned long long global_m, global_n;
 
         UVEC top_words;
 
@@ -122,10 +123,11 @@ namespace planc {
           this->cols = cols;
           this->A0 = A;
           this->global_m = parent->global_m;
-          this->global_n = parent->global_n;
+          this->global_n = cols.n_elem;
           this->W = W;
           this->parent = parent;
           this->index = index;
+          this->level = parent->level + 1;
           this->mpicomm = parent->mpicomm;
           this->pc = parent->pc;
           this->allocate();
@@ -229,7 +231,7 @@ namespace planc {
   template <class INPUTMATTYPE>
     class RootNode : public Node<INPUTMATTYPE> {
       public:
-        RootNode(INPUTMATTYPE * A, int global_m, int global_n, UVEC & cols, MPICommunicator * mpicomm, ParseCommandLine * pc) : Node<INPUTMATTYPE>() {
+        RootNode(INPUTMATTYPE * A, unsigned long long global_m, unsigned long long global_n, UVEC & cols, MPICommunicator * mpicomm, ParseCommandLine * pc) : Node<INPUTMATTYPE>() {
           this->cols = cols;
           this->A0 = A;
           this->global_m = global_m;
@@ -240,6 +242,7 @@ namespace planc {
           this->A = INPUTMATTYPE(*A);
           this->sigma = 0.0;
           this->index = 0;
+          this->level = 0;
           this->lvalid = false;
           this->rvalid = false;
         }

### 2020-06-29 latex/paper.tex
adding our macros
added some algorithms to paper

@@
 \usepackage{amsmath,amssymb,amsfonts}
-\usepackage{algorithmic}
+\usepackage{algorithm,algpseudocode}
 \usepackage{graphicx}
@@
 \usepackage{xcolor}
+\usepackage{cleveref}
+\usepackage{ourmacros}
 \def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
@@
 
+\begin{algorithm}
+\caption{Rank-2 Nonnegative Least Squares Solve}
+\label{alg:r2nmf}
+\begin{algorithmic}[1]
+	\Require{$\M{C}$ is $n\times 2$ and $\M{G}$ is $2\times 2$ and s.p.d.}
+	\Function{$\M{H} =$ Rank2-NLS-Solve}{$\M{C},\M{G}$}
+		\State $\M{H} = \M{C}\M{G}^{-1}$
+		\For{$i=1$ to $n$}
+			\If{$\ME{H}{i1} < 0$ or $\ME{H}{i2} < 0$}
+				\If{$\ME{H}{i1} \sqrt{\ME{G}{11}}  < \ME{H}{i2} \sqrt{\ME{G}{22}} $}
+					\State $\ME{H}{i1} = 0$
+					\State $\ME{H}{i2} = \ME{C}{i2} / \ME{G}{22}$
+				\Else
+					\State $\ME{H}{i1} = \ME{C}{i1} / \ME{G}{11}$
+					\State $\ME{H}{i2} = 0$
+				\EndIf
+			\EndIf
+		\EndFor
+	\EndFunction
+	\Ensure{$\displaystyle \M{H} = \argmin_{\M[\bar]{H} \geq \M{0}} \|\M{A}-\M{W}\M[\bar]{H}^\Tra\|$ is $n\times 2$ with $\M{C}=\M{A}^\Tra \M{W}$ and $\M{G}=\M{W}^\Tra\M{W}$}
+\end{algorithmic}
+\end{algorithm}
+
 \subsubsection{Hierarchical Clustering}
@@
 \subsubsection{Algorithm}
+
+\begin{algorithm}
+\caption{Parallel Rank-2 NMF}
+\label{alg:parrank2nmf}
+\begin{algorithmic}[1]
+	\Require{$\M{A}$ is $m\times n$ and row-distributed across processors so that $\M[\hat]{A}$ is local $(m/p)\times n$ submatrix}
+	\Function{$[\M{W},\M{H}] =$ Parallel-Rank2-NMF}{$\M{A}$}
+		\State Initialize local $\M[\hat]{W}$ randomly
+		\While{not converged}
+			\State \Comment{Compute $\M{H}$}
+			\State $\M[\hat]{G}_W = \M[\hat]{W}^\Tra \M[\hat]{W}$
+			\State $\M{G}_W = \textsc{All-Reduce}(\M[\hat]{G})$
+			\State $\M[\hat]{B} = \M[\hat]{W}^\Tra \M[\hat]{A}$ %\hfill \Comment{local matrix multiplication}
+			\State $\M[\hat]{C} = \textsc{Reduce-Scatter}(\M[\hat]{B})$
+			\State $\M[\hat]{H} = \textsc{Rank2-NLS-Solve}(\M[\hat]{C},\M{G}_W)$
+			\State \Comment{Compute $\M{W}$}
+			\State $\M[\hat]{G}_H = \M[\hat]{H}^\Tra \M[\hat]{H}$
+			\State $\M{G}_H = \textsc{All-Reduce}(\M[\hat]{G}_H)$
+			\State $\M{H} = \textsc{All-Gather}(\M[\hat]{H})$
+			\State $\M[\hat]{D} = \M[\hat]{A} \M{H}$ %\hfill \Comment{local matrix multiplication}
+			\State $\M[\hat]{W} = \textsc{Rank2-NLS-Solve}(\M[\hat]{D},\M{G}_H)$
+		\EndWhile
+	\EndFunction
+	\Ensure{$\M{A} \approx \M{W}\M{H}^\Tra$ with $\M{W}$, $\M{H}$ row-distributed}
+\end{algorithmic}
+\end{algorithm}
 
@@
 
+\GB{Lawton, can you add a description of Summit?  use the recent SC submission as an example}
+
 \subsection{Datasets}
+
+\GB{Lawton, can you add these descriptions?}
+\begin{itemize}
+	\item synthetic (dense and sparse)
+	\item HSI: DC and Aviris
+	\item Doc-clustering: Reuters and PubMed
+\end{itemize}
 
@@
 
+\subsubsection{Rank-2 NMF Strong Scaling}
+
+\GB{Lawton, can you gather this data, Ramki can you visualize it?}
+\begin{itemize}
+	\item compare 1D distribution scaling with optimal 2D grid scaling
+	\item use synthetic short-and-fat dense and squarish sparse
+\end{itemize}
+
+\subsubsection{Level Scaling}
+
+\GB{Ramki, can you add a bar plot to show time breakdown across complete levels?  Let's do Aviris and PubMed; just use levels data file}
+
 \subsubsection{Weak Scaling}
 
+\GB{maybe this section is lower priority}
+\begin{itemize}
+	\item dense and sparse
+\end{itemize}
+
 \subsubsection{Strong Scaling}
+
+\begin{itemize}
+	\item big dense synthetic
+	\item big sparse synthetic
+	\item Aviris
+	\item PubMed
+\end{itemize}
 
@@
 
+\GB{Lawton, show some visualization of the tree}
+
 \subsubsection{Document Corpus}
+
+\GB{Lawton, show a visualization of top words in tree}
 

### 2020-06-30 latex/paper.tex
drafted abstract

@@
 \begin{abstract}
-\GB{to be done...}
+Nonnegative Matrix Factorization (NMF) is an effective tool for clustering nonnegative data, either for computing a flat partitioning of a dataset or for determining a hierarchy of similarity.
+In this paper, we propose a parallel algorithm for hierarchical clustering that uses a divide-and-conquer approach based on rank-two NMF to split a data set into two cohesive parts.
+Not only does this approach uncover more structure in the data than a flat NMF clustering, but also rank-two NMF can be computed more quickly than for general ranks, providing comparable overall time to solution.
+Furthermore, the rank-two NMF algorithm can be parallelized efficiently using a computational kernel similar to matrix-vector multiplication.
+Our data distribution and parallelization strategies are designed to maintain computational load balance throughout the data-dependent hierarchy of computation while limiting interprocess communication, allowing the algorithm to scale to large dense and sparse data sets.
+We demonstrate the scalability of our parallel algorithm in terms of data size and number of processors, and apply the hierarchical clustering approach to large (dense) hyperspectral imaging and (sparse) text document data.
 \end{abstract}

### 2020-06-30 latex/paper.tex
added sequential alg for hiernmf2

@@
 \begin{algorithm}
-\caption{Rank-2 Nonnegative Least Squares Solve}
+\caption{Rank-2 Nonnegative Least Squares Solve \cite{KP13}}
 \label{alg:r2nmf}
@@
 \subsubsection{Hierarchical Clustering}
+
+\begin{algorithm}
+\caption{Hierarchical Clustering via Rank-2 NMF}
+\label{alg:hiernmf2}
+\begin{algorithmic}[1]
+	\Require{$\M{A}$ is $m\times n$, $k$ is target number of leaves}
+	\Function{${\cal T} =$ Hier-R2-NMF}{$\M{A}$}
+		\State ${\cal R} = \text{node}(\M{A})$ \hfill \Comment{create root node}
+		\State inject$({\cal Q},{\cal R})$ \hfill \Comment{create priority queue of leaves}
+		\While{leaves$({\cal T}) < k$}
+			\State ${\cal N} = \text{deleteMax}({\cal Q})$ \hfill \Comment{get leaf with max score}
+			\State $[\M{W},\M{H}] = \textsc{Rank2-NMF}({\cal N}.\M{A})$ \hfill \Comment{split $\cal N$}
+			\State partition ${\cal N}.\M{A}$ into $\M{A}_1$ and $\M{A}_2$ using $\M{H}$
+			\State ${\cal N}_1 = \text{node}(\M{A}_1,\text{score}(\M{A}_1),\V{W}_1)$ \hfill \Comment{left child}
+			\State ${\cal N}_2 = \text{node}(\M{A}_2,\text{score}(\M{A}_2),\V{W}_2)$ \hfill \Comment{right child}
+			\State ${\cal N}.\text{left} = {\cal N}_1$, ${\cal N}.\text{right} = {\cal N}_2$
+			\State inject$({\cal Q},{\cal N}_1)$, inject$({\cal Q},{\cal N}_2)$ 
+		\EndWhile
+	\EndFunction
+	\Ensure{$\cal{T}$ is binary tree with $k$ leaves, each node has subset of columns of $\M{A}$ and feature vector $\V{w}$}
+\end{algorithmic}
+\end{algorithm}
 

### 2020-06-30 utilities/partition-uneven.cpp
redone 50 node experiment

@@ -71,13 +71,12 @@ int main(int argc, char **argv) {
   std::vector< std::vector<uint64_t> > procColIdxs(procCount);
   std::vector< std::vector<double> > procVals(procCount);
   printf("Partitioning nonzeros and outputting to files...\n");
-  int order;
   uint64_t nnz;
   uint64_t rowCount, colCount;
   fscanf(file, "%" SCNu64, &rowCount);
   fscanf(file, "%" SCNu64, &colCount);
   fscanf(file, "%" SCNu64, &nnz);
-  printf("order::%d::nnz::%ju::rowcount::%ju::colcount::%ju\n", order, nnz, rowCount, colCount);
+  printf("nnz::%ju::rowcount::%ju::colcount::%ju\n", nnz, rowCount, colCount);
   for (uint64_t i = 0; i < nnz; i++) {
     uint64_t rowIdx, colIdx;
     double val;
@@ -114,6 +113,7 @@ int main(int argc, char **argv) {
     for (uint64_t j = 0; j < curRowIdxs.size(); j++) {
       fprintf(file, "%" PRIu64 " %" PRIu64 " %e\n", curRowIdxs[j], curColIdxs[j], curVals[j]);
     }
+    fprintf(file, "%" PRIu64 " %" PRIu64 " 0\n",m-1,n-1); 
     fclose(file);
   }
 

### 2020-06-30 utilities/partition-uniform.cpp

@@ -28,18 +28,18 @@ int main(int argc, char **argv) {
   std::vector< std::vector<uint64_t> > procColIdxs(procCount);
   std::vector< std::vector<double> > procVals(procCount);
   printf("Partitioning nonzeros and outputting to files...\n");
-  int order;
+  int order = 0;
   uint64_t nnz;
   uint64_t rowCount, colCount;
-  fscanf(file, " %d %" SCNu64, &order, &nnz);
   fscanf(file, "%" SCNu64, &rowCount);
   fscanf(file, "%" SCNu64, &colCount);
+  fscanf(file, "%" SCNu64, &nnz);
   printf("order::%d::nnz::%ju::rowcount::%ju::colcount::%ju\n", order, nnz, rowCount, colCount);
   uint64_t rowsPerProc = rowCount / rowProcCount;
   uint64_t colsPerProc = colCount / colProcCount;
   printf("Assigning %ju rows and %ju columns per part.\n", rowsPerProc, colsPerProc);
   for (uint64_t i = 0; i < nnz; i++) {
-    if (i % 100000 == 0 && i > 0) { printf("Processing %ju.th nonzero...\n", i); }
+    //if (i % 100000 == 0 && i > 0) { printf("Processing %ju.th nonzero...\n", i); }
     uint64_t rowIdx, colIdx;
     double val;
     fscanf(file, "%" SCNu64, &rowIdx);

### 2020-06-30 distnmf/distio.hpp
Revert "redone 50 node experiment"
This reverts commit 69a4d59ead5ba74267a9b4796d27d92253236451.

@@ -466,8 +466,8 @@ namespace planc {
                 SP_MAT temp_spmat(idxs, vals);
                 m_A = temp_spmat;
               }
-              m_A.load(sr.str(), arma::coord_ascii);
-              // uniform_dist_matrix(m_A);
+              // m_A.load(sr.str(), arma::coord_ascii);
+              uniform_dist_matrix(m_A);
 #else
               m_A.load(sr.str());
 #endif

### 2020-06-30 utilities/partition-uneven.cpp

@@ -71,12 +71,13 @@ int main(int argc, char **argv) {
   std::vector< std::vector<uint64_t> > procColIdxs(procCount);
   std::vector< std::vector<double> > procVals(procCount);
   printf("Partitioning nonzeros and outputting to files...\n");
+  int order;
   uint64_t nnz;
   uint64_t rowCount, colCount;
   fscanf(file, "%" SCNu64, &rowCount);
   fscanf(file, "%" SCNu64, &colCount);
   fscanf(file, "%" SCNu64, &nnz);
-  printf("nnz::%ju::rowcount::%ju::colcount::%ju\n", nnz, rowCount, colCount);
+  printf("order::%d::nnz::%ju::rowcount::%ju::colcount::%ju\n", order, nnz, rowCount, colCount);
   for (uint64_t i = 0; i < nnz; i++) {
     uint64_t rowIdx, colIdx;
     double val;
@@ -113,7 +114,6 @@ int main(int argc, char **argv) {
     for (uint64_t j = 0; j < curRowIdxs.size(); j++) {
       fprintf(file, "%" PRIu64 " %" PRIu64 " %e\n", curRowIdxs[j], curColIdxs[j], curVals[j]);
     }
-    fprintf(file, "%" PRIu64 " %" PRIu64 " 0\n",m-1,n-1); 
     fclose(file);
   }
 

### 2020-06-30 utilities/partition-uniform.cpp

@@ -28,18 +28,18 @@ int main(int argc, char **argv) {
   std::vector< std::vector<uint64_t> > procColIdxs(procCount);
   std::vector< std::vector<double> > procVals(procCount);
   printf("Partitioning nonzeros and outputting to files...\n");
-  int order = 0;
+  int order;
   uint64_t nnz;
   uint64_t rowCount, colCount;
+  fscanf(file, " %d %" SCNu64, &order, &nnz);
   fscanf(file, "%" SCNu64, &rowCount);
   fscanf(file, "%" SCNu64, &colCount);
-  fscanf(file, "%" SCNu64, &nnz);
   printf("order::%d::nnz::%ju::rowcount::%ju::colcount::%ju\n", order, nnz, rowCount, colCount);
   uint64_t rowsPerProc = rowCount / rowProcCount;
   uint64_t colsPerProc = colCount / colProcCount;
   printf("Assigning %ju rows and %ju columns per part.\n", rowsPerProc, colsPerProc);
   for (uint64_t i = 0; i < nnz; i++) {
-    //if (i % 100000 == 0 && i > 0) { printf("Processing %ju.th nonzero...\n", i); }
+    if (i % 100000 == 0 && i > 0) { printf("Processing %ju.th nonzero...\n", i); }
     uint64_t rowIdx, colIdx;
     double val;
     fscanf(file, "%" SCNu64, &rowIdx);

### 2020-06-30 build-summit/parse.py
parse R2NMF

@@ -1,0 +1,42 @@
+from os import listdir
+from os.path import isfile, join
+import sys
+from re import search, match
+from csv import writer
+from math import log2, floor
+
+name = sys.argv[1]
+totals = {}
+
+for line in open(name):
+  total = match(r"total_(.+?)::.*min::(.+?)::avg::(.+?)::max::(.+?)$",line)
+  if total:
+    cat = total.group(1)
+    times = [total.group(2), total.group(3), total.group(4)]
+    totals[cat] = times
+  
+  size = match(r".*globalm::(\d+?)::globaln::(\d+?)::.*",line)
+
+  if size:
+    m = size.group(1)
+    n = size.group(2)
+  
+  procs = match(r".*pr::(\d+?)::pc::(\d+?)::.*",line)
+
+  if procs:
+    pr = int(procs.group(1))
+    pc = int(procs.group(2))
+    p = pr*pc
+
+
+with open(name + '-output', 'w') as f:
+  write = writer(f,delimiter=',')
+  header = ['m','n','p','pr','pc']
+  data = [m,n,p,pr,pc]
+  for total in totals.keys():
+    header.append(total + '-min')
+    header.append(total + '-avg')
+    header.append(total + '-max')
+    data = data + totals[total]
+  write.writerow(header)
+  write.writerow(data)

### 2020-07-01 latex/paper.tex
reformulated hiernmf2 code to use split function (more accurate to actual implementation)

--- 2020-06-30 latex/paper.tex
@@
 \usepackage{cleveref}
+\usepackage{tikz}
 \usepackage{ourmacros}
@@
 
-test reference: \cite{GKP15}
+\section{Preliminaries}
 
-\section{Preliminaries}
+\begin{itemize}
+	\item NMF, BCD, rank-2 NMF
+\end{itemize}
 
@@
 \subsubsection{Hierarchical Clustering}
+
+\begin{algorithm}
+\caption{Hierarchical Clustering \cite{KP13}}
+\label{alg:hiernmf2}
+\begin{algorithmic}[1]
+	\Require{$\M{A}$ is $m\times n$, $k$ is target number of leaves}
+	\Function{${\cal T} =$ Hier-R2-NMF}{$\M{A}$}
+		\State ${\cal R} = \text{node}(\M{A})$ \hfill \Comment{create root node}
+		\State \textsc{Split}$({\cal R})$
+		\State inject$({\cal Q},{\cal R}.\text{left})$ \hfill \Comment{create priority queue}
+		\State inject$({\cal Q},{\cal R}.\text{right})$ \hfill \Comment{of frontier nodes}
+		\While{size$({\cal Q}) < k$}
+			\State ${\cal N} = \text{eject}({\cal Q})$ \hfill \Comment{get leaf with max score}
+			\State \textsc{Split}$({\cal R}.\text{left})$ \hfill \Comment{split left child}
+			\State inject$({\cal Q},{\cal R}.\text{left})$ \hfill \Comment{and add to $\cal Q$}
+			\State \textsc{Split}$({\cal R}.\text{right})$ \hfill \Comment{split right child}
+			\State inject$({\cal Q},{\cal R}.\text{right})$ \hfill \Comment{and add to $\cal Q$}
+		\EndWhile
+	\EndFunction
+	\Ensure{$\cal{T}$ is binary tree rooted at $\cal R$ with $k$ frontier nodes, each node has subset of cols of $\M{A}$ and feature vector $\V{w}$}
+\end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}
+\caption{Node Splitting via Rank-Two NMF}
+\label{alg:split}
+\begin{algorithmic}[1]
+	\Require{$\cal N$ has a subset of columns given by field $\M{A}$}
+	\Function{Split}{$\cal N$}
+		\State $[\M{W},\M{H}] = \textsc{Rank2-NMF}({\cal N}.\M{A})$ \hfill \Comment{split $\cal N$}
+		\State partition ${\cal N}.\M{A}$ into $\M{A}_1$ and $\M{A}_2$ using $\M{H}$
+		\State ${\cal N}.\text{left} = \text{node}(\M{A}_1,\V{W}_1)$ \hfill \Comment{create left child}
+		\State ${\cal N}.\text{right} = \text{node}(\M{A}_2,\V{W}_2)$ \hfill \Comment{create right child}
+		\State ${\cal N}.\text{score} = \sigma_1(\M{A}_1) + \sigma_1(\M{A}_2) - \sigma_1({\cal N}.\M{A})$
+	\EndFunction
+	\Ensure{$\cal{N}$ has two children and a score}
+\end{algorithmic}
+\end{algorithm}
 
@@
 \end{algorithm}
+
+\begin{figure}
+\input{fig/split}
+\caption{Parallel splitting using Rank-2 NMF}
+\label{fig:split}
+\end{figure}
 

### 2020-07-01 latex/fig/tree.tex
added tree figure

%!TEX root = ../paper.tex

\centering
\begin{tikzpicture}[scale=.9,level/.style={sibling distance=50mm/#1}]
\node [circle,draw] (z) {}
  child {node [circle,draw,fill=gray] (a) {}
    child {node [circle,draw,dashed] (b) {}}
    child {node [circle,draw,dashed] (g) {}}
  }
  child {node [circle,draw] (j) {}
    child {node [circle,draw] (k) {}
      child {node [circle,draw,fill=gray] (o) {}
        child {node [circle,draw,dashed] (x) {}}
        child {node [circle,draw,dashed] (y) {}}
      }
      child {node [circle,draw,fill=gray] (p) {}
        child {node [circle,draw,dashed] (x) {}}
        child {node [circle,draw,dashed] (y) {}}
      }
    }
  child {node [circle,draw,fill=gray] (l) {}
    child {node [circle,draw,dashed] (c) {}}
    child {node [circle,draw,dashed] (d) {}}
  }
};

\node [shift={(-3.5,-4.25)},circle,draw,label={[xshift=1.5cm, yshift=-0.4cm] Internal Node}] (leg1) {};
\node [shift={(-3.5,-4.75)},circle,draw,fill=gray,label={[xshift=1.5cm, yshift=-0.4cm] Frontier Node}] (leg1) {};
\node [shift={(-3.5,-5.25)},circle,draw,dashed,label={[xshift=1.5cm, yshift=-0.4cm] Leaf Node}] (leg1) {};
\end{tikzpicture}

### 2020-07-02 build-summit/tree.py
tree utility

@@ -1,0 +1,32 @@
+import sys
+
+if __name__ == '__main__':
+    tree = sys.argv[1]
+    dictionary = sys.argv[2]
+    perm = sys.argv[3]
+    
+    nodes = {}
+    with open(tree) as f:
+        for line in f.readlines():
+            items = [int(item) for item in line.split()]
+            idx = items[0]
+            nodes[idx] = items[1:]
+    
+    idxs = []
+    with open(perm) as f:
+        oidxs = [int(line) for line in f.readlines()]
+        idxs = [0]*len(oidxs)
+        for idx,val in enumerate(oidxs):
+          idxs[val] = idx
+
+    with open(dictionary) as f:
+        words = f.read().splitlines()
+        for idx in nodes:
+            nodes[idx] = [words[idxs[i]] for i in nodes[idx]]
+
+    for idx in nodes:
+        print(idx, nodes[idx])
+
+
+    
+            

### 2020-07-03 distnmf/distio.hpp
sparse matrix read

@@ -466,8 +466,8 @@ namespace planc {
                 SP_MAT temp_spmat(idxs, vals);
                 m_A = temp_spmat;
               }
-              // m_A.load(sr.str(), arma::coord_ascii);
-              uniform_dist_matrix(m_A);
+              m_A.load(sr.str(), arma::coord_ascii);
+              // uniform_dist_matrix(m_A);
 #else
               m_A.load(sr.str());
 #endif

### 2020-07-03 utilities/partition-uneven.cpp
padding partitioning

@@ -114,6 +114,7 @@ int main(int argc, char **argv) {
     for (uint64_t j = 0; j < curRowIdxs.size(); j++) {
       fprintf(file, "%" PRIu64 " %" PRIu64 " %e\n", curRowIdxs[j], curColIdxs[j], curVals[j]);
     }
+    fprintf(file, "%" PRIu64 " %" PRIu64 " 0\n", m-1,n-1);
     fclose(file);
   }
 

### 2020-07-03 latex/fig/tree.tex

@@
   child {node [circle,draw,fill=gray] (a) {}
-    child {node [circle,draw,dashed] (b) {}}
-    child {node [circle,draw,dashed] (g) {}}
+    child[dashed] {node [circle,draw,dashed] (b) {}}
+    child[dashed] {node [circle,draw,dashed] (g) {}}
   }
@@
       child {node [circle,draw,fill=gray] (o) {}
-        child {node [circle,draw,dashed] (x) {}}
-        child {node [circle,draw,dashed] (y) {}}
+        child[dashed] {node [circle,draw,dashed] (x) {}}
+        child[dashed] {node [circle,draw,dashed] (y) {}}
       }
       child {node [circle,draw,fill=gray] (p) {}
-        child {node [circle,draw,dashed] (x) {}}
-        child {node [circle,draw,dashed] (y) {}}
+        child[dashed] {node [circle,draw,dashed] (x) {}}
+        child[dashed] {node [circle,draw,dashed] (y) {}}
       }
@@
   child {node [circle,draw,fill=gray] (l) {}
-    child {node [circle,draw,dashed] (c) {}}
-    child {node [circle,draw,dashed] (d) {}}
+    child[dashed] {node [circle,draw,dashed] (c) {}}
+    child[dashed] {node [circle,draw,dashed] (d) {}}
   }

### 2020-07-03 latex/paper.tex
added a few notes

@@
 
+\GB{ask Haesun to write intro paragraph motivating hierarchical clustering but noting the limitations of time and data size}
+
 \section{Preliminaries}
@@
 	\item NMF, BCD, rank-2 NMF
+	\item PLANC
 \end{itemize}
@@
 
+\begin{itemize}
+	\item subroutine for splitting node is rank-2 NMF
+	\item using BCD, subproblem is set of independent NNLS problems with variable vector of length 2
+	\item 4 possibilities
+	\begin{itemize}
+		\item all-zero option never optimum if input is nonnegative
+		\item if all-pos option is optimum, it is solution to unconstrained problem
+		\item can choose between one-zero options using line 6 of algorithm
+		\item describe \cref{alg:r2nmf}
+	\end{itemize}
+\end{itemize}
+
 \begin{algorithm}
@@
 	\Function{$\M{H} =$ Rank2-NLS-Solve}{$\M{C},\M{G}$}
-		\State $\M{H} = \M{C}\M{G}^{-1}$
+		\State $\M{H} = \M{C}\M{G}^{-1}$ \hfill \Comment{Solve unconstrained system}
 		\For{$i=1$ to $n$}
 			\If{$\ME{H}{i1} < 0$ or $\ME{H}{i2} < 0$}
+				\State \Comment{Choose between single-variable solutions}
 				\If{$\ME{H}{i1} \sqrt{\ME{G}{11}}  < \ME{H}{i2} \sqrt{\ME{G}{22}} $}
@@
 		\While{size$({\cal Q}) < k$}
-			\State ${\cal N} = \text{eject}({\cal Q})$ \hfill \Comment{get leaf with max score}
-			\State \textsc{Split}$({\cal R}.\text{left})$ \hfill \Comment{split left child}
-			\State inject$({\cal Q},{\cal R}.\text{left})$ \hfill \Comment{and add to $\cal Q$}
-			\State \textsc{Split}$({\cal R}.\text{right})$ \hfill \Comment{split right child}
-			\State inject$({\cal Q},{\cal R}.\text{right})$ \hfill \Comment{and add to $\cal Q$}
+			\State ${\cal N} = \text{eject}({\cal Q})$ \hfill \Comment{frontier node with max score}
+			\State \textsc{Split}$({\cal N}.\text{left})$ \hfill \Comment{split left child}
+			\State inject$({\cal Q},{\cal N}.\text{left})$ \hfill \Comment{and add to $\cal Q$}
+			\State \textsc{Split}$({\cal N}.\text{right})$ \hfill \Comment{split right child}
+			\State inject$({\cal Q},{\cal N}.\text{right})$ \hfill \Comment{and add to $\cal Q$}
 		\EndWhile
@@
 \end{algorithm}
+
+\begin{figure}
+\input{fig/tree}
+\caption{Hierarchy node classification}
+\label{fig:tree}
+\end{figure}
 

### 2020-07-04 latex/paper.tex
added description of rank-2 nnls algorithm

@@
 \usepackage{tikz}
+\usepackage{mathtools}
 \usepackage{ourmacros}
@@
 \begin{itemize}
-	\item NMF, BCD, rank-2 NMF
+	\item NMF, BCD, NNLS
 	\item PLANC
@@
 
+Using the 2-block BCD approach for a rank-2 NMF yields NNLS subproblems of the form $\displaystyle \min_{\M[\bar]{H} \geq \M{0}} \|\M{W}\M[\bar]{H}^\Tra-\M{A}\|$ and $\min_{\M[\bar]{W} \geq \M{0}} \|\M{H}\M[\bar]{W}^\Tra-\M{A}^\Tra\|$.
+In each case, the columns of the transposed variable matrix can be computed independently.
+Considering the $i$th row of $\M[\bar]{H}$, for example, the NNLS problem to solve is
+\begin{align*}
+\begin{multlined}
+\min_{\ME[\bar]{H}{i,1},\ME[\bar]{H}{i,2} \geq 0} \left\|\begin{bmatrix} \V{w}_1 & \V{w}_2 \end{bmatrix} \begin{bmatrix}\ME[\bar]{H}{i,1} \\ \ME[\bar]{H}{i,2} \end{bmatrix} - \V{A}_i \right\| \\ = \min_{\ME[\bar]{H}{i,1},\ME[\bar]{H}{i,2} \geq 0} \left\|\ME[\bar]{H}{i,1} \V{w}_1 + \ME[\bar]{H}{i,2} \V{w}_2 - \V{a}_i \right\|
+\end{multlined}
+\end{align*}
+where $\V{w}_1$ and $\V{W}_2$ are the two columns of $\M{W}$ and $\V{a}_i$ is the $i$ column of $\M{A}$.
+We note that there are four possibilities of solutions, as each of the two scalar variables may be positive or zero.
+
+As shown by Kuang and Park \cite{KP13}, determining which of the four possible solutions is feasible and optimal can be done efficiently by exploiting the following properties:
+\begin{itemize}
+	\item if the solution to the unconstrained least squares problem admits two positive values, it is the optimal solution to the nonnegatively constrained problem,
+	\item if $\M{W}$ and $\M{A}$ are both nonnegative, then the candidate solution with two zero values is never (uniquely) optimal and can be discarded, and
+	\item if the solution to the unconstrained problem does not admit two positive values, the better of the two remaining solutions can be determined by comparing $\V{a}_{j}^\Tra\V{w}_1 / \|\V{w}_1\|$ and $\V{a}_{j}^\Tra\V{w}_2 / \|\V{w}_2\|$.
+\end{itemize}
+If the unconstrained problem is solved via the normal equations, then the temporary matrices computed for the normal equations ($\M{W}^\Tra\M{W}$ and $\M{A}^\Tra\M{W}$) can be re-used to determine the better of the two single-positive-variable solutions.
+
+\Cref{alg:r2nmf} implements this strategy for all rows of $\M{H}$ simultaneously.
+It takes as input the matrices $\M{C}=\M{A}^\Tra\M{W}$ and $\M{G}=\M{W}^\Tra\M{W}$, first solves the normal equations for the unconstrained problem, and then determines between the two alternate possibilities as necessary.
+\GB{To avoid two passes through the data, we could combine the solve and the update if necessary.
+It might also help to pre-compute the inverse of $\M{G}$...
+Also Lawton, please check the details of the pseudocode!}
+
 \begin{algorithm}
@@
 				\State \Comment{Choose between single-variable solutions}
-				\If{$\ME{H}{i1} \sqrt{\ME{G}{11}}  < \ME{H}{i2} \sqrt{\ME{G}{22}} $}
+				\If{$\ME{C}{i1} / \sqrt{\ME{G}{11}}  < \ME{C}{i2} / \sqrt{\ME{G}{22}} $}
 					\State $\ME{H}{i1} = 0$

### 2020-07-06 latex/paper.tex
slight progress on alg section

@@
 
-\begin{itemize}
-	\item subroutine for splitting node is rank-2 NMF
-	\item using BCD, subproblem is set of independent NNLS problems with variable vector of length 2
-	\item 4 possibilities
-	\begin{itemize}
-		\item all-zero option never optimum if input is nonnegative
-		\item if all-pos option is optimum, it is solution to unconstrained problem
-		\item can choose between one-zero options using line 6 of algorithm
-		\item describe \cref{alg:r2nmf}
-	\end{itemize}
-\end{itemize}
-
 Using the 2-block BCD approach for a rank-2 NMF yields NNLS subproblems of the form $\displaystyle \min_{\M[\bar]{H} \geq \M{0}} \|\M{W}\M[\bar]{H}^\Tra-\M{A}\|$ and $\min_{\M[\bar]{W} \geq \M{0}} \|\M{H}\M[\bar]{W}^\Tra-\M{A}^\Tra\|$.
@@
 \end{itemize}
-If the unconstrained problem is solved via the normal equations, then the temporary matrices computed for the normal equations ($\M{W}^\Tra\M{W}$ and $\M{A}^\Tra\M{W}$) can be re-used to determine the better of the two single-positive-variable solutions.
+If the unconstrained problem is solved via the normal equations, then the temporary matrices computed for the normal equations ($\M{W}^\Tra\M{W}$ and $\M{A}^\Tra\M{W}$) can be re-used to determine the better of the two solutions with a single positive variable.
 
-\Cref{alg:r2nmf} implements this strategy for all rows of $\M{H}$ simultaneously.
-It takes as input the matrices $\M{C}=\M{A}^\Tra\M{W}$ and $\M{G}=\M{W}^\Tra\M{W}$, first solves the normal equations for the unconstrained problem, and then determines between the two alternate possibilities as necessary.
+\Cref{alg:r2nnls} implements this strategy for all rows of $\M{H}$ simultaneously.
+It takes as input the matrices $\M{C}=\M{A}^\Tra\M{W}$ and $\M{G}=\M{W}^\Tra\M{W}$, first solves the normal equations for the unconstrained problem, and then chooses between the two alternate possibilities as necessary.
+We note that each row of $\M{H}$ is independent, and therefore this algorithm is easily parallelized.
+Solving for $\M{W}$ can be done using inputs $\M{C}=\M{A}\M{H}$ and $\M{G}=\M{H}^\Tra\M{H}$.
 \GB{To avoid two passes through the data, we could combine the solve and the update if necessary.
@@
 \caption{Rank-2 Nonnegative Least Squares Solve \cite{KP13}}
-\label{alg:r2nmf}
+\label{alg:r2nnls}
 \begin{algorithmic}[1]
@@
 
+Given that the computational complexity of \Cref{alg:r2nnls} is $O(n)$ (or $O(m)$ when computing $\M{W}$), and the complexity of computing $\M{W}^\Tra\M{W}$ and $\M{H}^\Tra\M{H}$ is $O(m+n)$, the typical dominant cost of each iteration of Rank-2 NMF is that of computing $\M{A}^\Tra\M{W}$ and $\M{A}\M{H}$, which is $O(mn)$ when $\M{A}$ is dense and $O(\text{nnz}(A))$ when $\M{A}$ is sparse.
+
 \subsubsection{Hierarchical Clustering}
+
+\begin{itemize}
+	\item summarize approach: expand binary tree until number of leaves hits desired target
+	\item choosing leaf node to split is done using priority queue based on that leaf's score
+	\begin{itemize}
+		\item score can be defined in multiple ways...
+	\end{itemize}
+	\item explain frontier nodes
+	\item explain \Cref{alg:hiernmf2} notation, fields of nodes, etc.
+\end{itemize}
 
@@
 \begin{algorithmic}[1]
-	\Require{$\M{A}$ is $m\times n$, $k$ is target number of leaves}
+	\Require{$\M{A}$ is $m\times n$, $k$ is target number of leaf clusters}
 	\Function{${\cal T} =$ Hier-R2-NMF}{$\M{A}$}

### 2020-07-07 latex/paper.tex

@@
 
-\begin{itemize}
-	\item summarize approach: expand binary tree until number of leaves hits desired target
-	\item choosing leaf node to split is done using priority queue based on that leaf's score
-	\begin{itemize}
-		\item score can be defined in multiple ways...
-	\end{itemize}
-	\item explain frontier nodes
-	\item explain \Cref{alg:hiernmf2} notation, fields of nodes, etc.
-\end{itemize}
+A Rank-2 NMF can be used to partition the columns of the matrix into two parts.
+In this case, the columns of the $\M{W}$ factor represent feature weights for each of the two latent components, and the strength of membership in the two components for each column of $\M{A}$ is given by the two values in the corresponding row of $\M{H}$.
+We can determine part membership by comparing those values: if $\ME{H}{i1} > \ME{H}{i2}$, then column $i$ of $\M{A}$ is assigned to the first part, which is associated with feature vector $\V{w}_1$.
+Membership can be determined by other metrics that also take into account balance across parts or attempt to detect outliers.
+
+Given Rank-2 NMF as a splitting procedure, hierarchical clustering builds a binary tree such that each node corresponds to a subset of samples from the original data set and each node's children correspond to a 2-way partition of the node's samples.
+In this way, the leaves form a partition of the original data, and the internal nodes specify the hierarchical relationship among clusters.
+As the tree is built, nodes are split in order of their score, or relative value to the overall clustering of the data.
+The process can be continued until a target number of leaves is produced or until all remaining leaves have a score below a given threshold.
+
+A node's score can be computed in different ways.
+For document clustering, Kuang and Park \cite{KP13} propose using modified normalized discounted cumulative gain, which measures how distinct a node's children are from each other using the feature weights associated with the node and its children.
+For hyperspectral imaging data, Gillis et al.
+\cite{GKP15} propose using the possible reduction in overall NMF error if the node is split -- the difference in error between using the node itself or using its children.
+We use the latter in our implementation.
+
+In any case, a node's score depends on properties of its children, so the computation for a split must be done before the split is actually accepted.
+To this end, we define a \emph{frontier} node to be a parent of leaves; these are nodes whose children have been computed but whose splits have not been accepted.
+\Cref{fig:tree} depicts the classification of nodes into internal, frontier, and leaf nodes.
+As the tree is built, the algorithm selects the frontier node with the highest score to split, though no computation is required to split the node.
+When a frontier node split is accepted, it becomes an internal node and its children are split (so that their scores can be computed) and added to the set of frontier nodes.
+When the algorithm terminates, the leaves are discarded and the frontier nodes become the leaves of the output tree.
+
+\begin{figure}
+\input{fig/tree}
+\caption{Hierarchy node classification}
+\label{fig:tree}
+\end{figure}
+
+Our hierarchical clustering algorithm is presented in \Cref{alg:hiernmf2} and follows that of Kuang and Park \cite{KP13}.
+Each node includes a field $\M{A}$, which is a subset of columns (samples) of the original data, a feature vector $\V{w}$, which is its corresponding column of the $\M{W}$ matrix from its parent's Rank-2 NMF, a score, and pointers to its left and right children.
+A priority queue $\cal Q$ tracks the frontier nodes so that the node with the highest score is split at each step of the algorithm.
+We use a target number of leaf clusters $k$ as the termination condition.
+When a node is selected from the priority queue, it is removed from the set of frontier nodes and its children are added.
 
@@
 
-\begin{figure}
-\input{fig/tree}
-\caption{Hierarchy node classification}
-\label{fig:tree}
-\end{figure}
+The splitting procedure is specified in \Cref{alg:split}.
+After the Rank-2 NMF is performed, the $\M{H}$ factor is used to determine part membership, and the columns of the $\M{W}$ factor are assigned to the child nodes.
+The score of the node is computed as the reduction in overall NMF error if the node is split, which can be computed from the principal singular values of the subsets of columns of the node and its children.
+The principal singular values of the children are computed via the power method (the principal singular value of the node was computed when its parent's score was computed).
 
@@
 		\State ${\cal N}.\text{right} = \text{node}(\M{A}_2,\V{W}_2)$ \hfill \Comment{create right child}
-		\State ${\cal N}.\text{score} = \sigma_1(\M{A}_1) + \sigma_1(\M{A}_2) - \sigma_1({\cal N}.\M{A})$
+		\State ${\cal N}.\text{score} = \sigma_1^2(\M{A}_1) + \sigma_1^2(\M{A}_2) - \sigma_1^2({\cal N}.\M{A})$
 	\EndFunction
@@
 \input{fig/split}
-\caption{Parallel splitting using Rank-2 NMF}
+\caption{Parallel splitting using Rank-2 NMF \GB{add more to caption and possibly color code W and H}}
 \label{fig:split}

### 2020-07-11 latex/paper.tex
Summit description

@@
 \subsection{Experimental Platform}
+All the experiments in this section were conducted on Summit. Summit is a supercomputer created by IBM for the Oak Ridge National Laboratory. 
+There are approximately 4,600 nodes on Summit. Each node contains two IBM POWER9 processors on separate sockets with dual NVLINK bricks to facilitate a transfer rate of 
+25 GB/s between the processors. Each node contains 512 GB of DDR4 memory for use by both processors. Each POWER9 processor utilizes 22 IBM SIMD Multi-Cores (SMCs), 
+although one of these SMCs on each processor is dedicated to memory transfer and is therefore not available for computation. 
+For node scaling experiments, all 42 available SMCs were utilized in each node so that every node computed with 42 separate MPI processes.
+Additionally, every node also supports six NVIDIA Volta V100 accelerators but these were unused by our algorithm.
 
-\GB{Lawton, can you add a description of Summit?  use the recent SC submission as an example}
+H2NMF uses the Armadillo library (version 9.900.1) for all matrix operations. 
+In Armadillo, sparse matrices are stored in Compressed Sparse Column (CSC) format and dense matrices are stored in column-major ordering.
+On Summit, we linked this version of Armadillo with OpenBLAS (version 0.3.9) and IBM's Spectrum MPI (version 10.3.1.2-20200121).
 
@@
 \begin{itemize}
+
 	\item synthetic (dense and sparse)

### 2020-07-11 latex/paper.tex
hyperspectral and document descriptions

@@
 
-\GB{Lawton, can you add these descriptions?}
 \begin{itemize}
+	\item \textit{Document Clustering}:
+	These datasets are represented by sparse matrices in term frequency-inverdse document frequency (TFIDF) format.
+	The rows of these matrices represent words and the columns represent individual documents. Each value in the matrix is a TFIDF statistic for
+	a specific word in a document.
+
+	For H2NMF, we used two popular document clustering datasets: Reuters Newswire Topic Classification (Reuters-21578) and \dots
+	\LM{I can't find the official name for PubMed or a link to it}
+
+	The Reuters matrix is $12411\times 7984$ and is a preprocessed version of Reuters-21578 used in the SmallK implementation of H2NMF.
+	The preprocessed version is the 20 largest prelabeled classes of documents with only single class labels. 
+	\LM{Prof Park, is this explanation of Reuters correct?}
+	As Reuters is small, it is mainly used as a comparison tool between the ours and the SmallK implmentation.
+
+	The PubMed matrix is $141043\times 8200000$. It is not preprocessed and represents the full size dataset in TFIDF format. This version is
+	better suited for scaling as is it is much larger than Reuters and is a good representation of large text classification datasets.
+
+	\item \textit{Hyperspectral Imaging}:
+	The images chosen for this dataset follow a similar pattern to Document Clustering. We chose two datasets: the Hyperspectral digital imagery
+	collection experiment (HYDICE) image of the Washington DC Mall and the Airborne Visible/Infrared Image Spectrometer (AVIRIS) image 
+	of the DC area.
+
+	The images are formatted into 3-way tensors representing individual images in each spectral band captured. For H2NMF, these tensors 
+	are flattened so that the rows represent spectral bands (191 for HYDICE and 224 for AVIRIS) and the columns represent pixels.
+
+	The HYDICE image is flattened from $1280\times 307\times 191$ to $191\times 392960$ and the AVIRIS image is likewise
+	flattened from $8940\times 747\times 224$ to $224\times 6678180$. Unlike with document clustering, hyperspectral image sizes grow in pixels
+	and not spectral bands. So, for both images, large scaling is impossible given our current data distribution.
+	\LM{I think this is awkward and I'm not sure if we need to include it}
+
+	\LM{DC: https://engineering.purdue.edu/~biehl/MultiSpec/hyperspectral.html}
+
+	\LM{AVIRIS: https://aviris.jpl.nasa.gov/dataportal/}
 

### 2020-07-14 latex/paper.tex
added first part of parallelization section

@@
 \section{Preliminaries}
+\label{sec:prelim}
 
@@
 
-\subsubsection{Algorithm}
+In this section, we consider the options for parallelizing Hierarchical Rank-2 NMF Clustering (\Cref{alg:hiernmf2}) and provide an analysis for the approach we take.
+The running time of an algorithm is data dependent because not only does each Rank-2 NMF computation require a variable number of iterations, but also the shape of the tree can vary from a balanced binary tree with $O(\log k)$ levels to a tall, flat tree with $O(k)$ levels.
+For the sake of analysis, we will assume a fixed number of NMF iterations for every node of the tree and we will analyze the cost of complete levels.
+
+The first possibility for parallelization is across the nodes of the tree, as each Rank-2 NMF split is independent.
+We choose not to parallelize across nodes in the tree for two reasons.
+The first reason is that while the NMF computations are independent, choosing which nodes to split may depend on global information.
+In particular, when the global target is to determine $k$ leaf clusters, the nodes must be split in order of their scores, which leads to a serialization of the node splits.
+This serialization might be relaxed using speculative execution, but it risks performing unnecessary computation.
+If the global target is to split all nodes with sufficiently high scores, then this serialization is also avoided and node splits become truly independent.
+We choose not to parallelize in this way to remain agnostic to the global stopping criterion.
+
+The second reason is that parallelizing across nodes requires redistribution of the input data.
+Given a node split by $\hat p$ processors, in order to assign disjoint sets of processors to each child node, each of the $\hat p$ processors would have to redistribute their local data, sending data for samples not in their child's set and receiving data for those in their child's set.
+The communication would be data dependent, but on average, each processor would communicate half of its data in the redistribution set, which could have an all-to-all communication pattern among the $\hat p$ processors.
+For a node with $\hat n$ columns, the communication cost would be at least $O(m\hat n/ \hat p)$ words, which is much larger than the communication cost per iteration of Parallel Rank-2 NMF, as we will see in \Cref{sec:analysis}.
+
+By choosing not to parallelize across nodes in the tree, we employ all $p$ processors on each node, and split nodes in sequence.
+The primary computations used to split a node are the Rank-2 NMF and the score computation, which is based on an approximation of the largest singular value.
+We use an alternating-updating algorithm for Rank-2 NMF as described in \Cref{sec:prelim}, and we parallelize it following the methodology proposed in \cite{EH+19-TR} and presented in \Cref{alg:parrank2nmf}.
+
+The communication cost of the algorithm depends on the parallel distribution of the input matrix data $\M{A}$.
+In order to avoid redistribution of the matrix data, we choose a 1D row distribution so that each processor owns a subset of the rows of $\M{A}$.
+Because the clustering partition splits the columns of $\M{A}$, each processor can partition its local data into left and right children to perform the split without any interprocessor communication.
+If we use a 2D distribution for a given node, then because the partition is data dependent, a data redistribution is required in order to obtain a load balanced distribution of both children.
+In the following subsections, we describe the parallel algorithms for Rank-2 NMF and approximating the principle singular value given this 1D data distribution and analyze their complexity in the context of the hierarchical clustering algorithm.
+
+\subsubsection{Algorithms}
+
+\paragraph{\emph{Parallel Rank-2 NMF}}
 
@@
 			\State $\M{G}_W = \textsc{All-Reduce}(\M[\hat]{G})$
-			\State $\M[\hat]{B} = \M[\hat]{W}^\Tra \M[\hat]{A}$ %\hfill \Comment{local matrix multiplication}
+			\State $\M[\hat]{B} = \M[\hat]{A}^\Tra \M[\hat]{W}$ %\hfill \Comment{local matrix multiplication}
 			\State $\M[\hat]{C} = \textsc{Reduce-Scatter}(\M[\hat]{B})$
@@
 
+\paragraph{\emph{Parallel Power Method}}
+
+\begin{algorithm}
+\caption{Parallel Power Method}
+\label{alg:parpowmeth}
+\begin{algorithmic}[1]
+	\Require{$\M{A}$ is $m\times n$ and row-distributed across processors so that $\M[\hat]{A}$ is local $(m/p)\times n$ submatrix}
+	\Function{$\sigma =$ Parallel-Power-Method}{$\M{A}$}
+		\State Initialize $\V{v}$ randomly and redundantly
+		\While{not converged}
+			\State $\V[\hat]{u} = \M[\hat]{A} \V{v}$
+			\State $\V[\hat]{z} = \M[\hat]{A}^\Tra \V[\hat]{u}$
+			\State $\V{v} = \textsc{All-Reduce}(\V[\hat]{z})$
+			\State $\sigma = \|\V{v}\|$
+			\State $\V{v} = \V{v}/\sigma$
+		\EndWhile
+	\EndFunction
+	\Ensure{$\sigma \approx \sigma_1(\M{A})$ is  redundantly owned by all procs}
+\end{algorithmic}
+\end{algorithm}
+
 \begin{figure}
@@
 \subsubsection{Analysis}
+\label{sec:analysis}
+
+\paragraph{\emph{Parallel Rank-2 NMF}}
+
+\paragraph{\emph{Parallel Power Method}}
+
+\paragraph{\emph{Hierarchical Clustering}}
 

### 2020-07-14 latex/paper.tex
added description of parallel algorithsm

@@
 If we use a 2D distribution for a given node, then because the partition is data dependent, a data redistribution is required in order to obtain a load balanced distribution of both children.
-In the following subsections, we describe the parallel algorithms for Rank-2 NMF and approximating the principle singular value given this 1D data distribution and analyze their complexity in the context of the hierarchical clustering algorithm.
+\Cref{fig:split} presents a visualization of the node-splitting process using a 1D processor distribution.
+In the following subsections, we describe the parallel algorithms for Rank-2 NMF and approximating the principal singular value given this 1D data distribution and analyze their complexity in the context of the hierarchical clustering algorithm.
+
+\begin{figure}
+\input{fig/split}
+\caption{Parallel splitting using Rank-2 NMF and 1D processor distribution.
+A Rank-2 NMF computes factor matrices $\M{W}$ and $\M{H}$ to approximate $\M{A}$, the values of $\M{H}$ are used to determine child membership of each column (either red or blue), and the corresponding column of the $\M{W}$ matrix represents the part's feature weighting.
+The 1D distribution is depicted for 3 processors to show that the splitting requires no interprocessor redistribution as the children are evenly distributed identically to the parent.}
+\label{fig:split}
+\end{figure}
 
@@
 \paragraph{\emph{Parallel Rank-2 NMF}}
+
+\Cref{alg:parrank2nmf} presents the parallelization of an alternating-updating scheme for NMF that uses the exact rank-2 solve algorithm presented in \Cref{alg:r2nnls} to update each factor matrix.
+The algorithm computes the inputs to the rank-2 solves in parallel and then exploits the parallelism across rows of the factor matrix so that each processor solves for a subset of rows simultaneously.
+The distribution of all matrices is 1D row distribution, so that each processor owns a subset of the rows of $\M{A}$, $\M{W}$, and $\M{H}$.
+We use the notation $\M[\hat]{A}$ to refer to the $(m/p) \times n$ local data matrix and $\M[\hat]{W}$ and $\M[\hat]{H}$ to refer to the $(m/p) \times 2$ and $(n/p)\times 2$ local factor matrices.
+With this distribution, the computation of $\M{W}^\Tra\M{W}$ and $\M{H}^\Tra\M{H}$ each is done via local multiplication followed by a single all-reduce collective.
+All processors own the data they need to compute their contribution to $\M{A}^\Tra \M{W}$; in order to distribute the result to compute the rows $\M{H}$ independently, a reduce-scatter collective is used to sum and simultaneously distribute across processors.
+To obtain the data needed to compute $\M[\hat]{W}$, each processor must access all of $\M{H}$, which is performed via an all-gather collective.
+The iteration progresses until a convergence criterion is satisfied.
+For performance benchmarking we use a fixed number of iterations, and in practice we use relative change in objective function value (residual norm).
 
@@
 
+In order to compute the score for a frontier node, we use the difference between the principal singular value of the matrix columns of the node and the sum of those of its children.
+Thus, we must determine the principal singular value of every node in the tree once, including leaf nodes.
+We use the power method to approximate it, repeatedly applying $\M{A}\M{A}^\Tra$ to a vector until it converges to the leading right singular vector.
+We present the power method in \Cref{alg:parpowmeth}.
+Note that we do not normalize the approximate left singular vector so that the computed value approximates the square of the largest singular value.
+
+Given the 1D distribution, only one communication collective is required for the pair of matrix-vector multiplications.
+That is, the approximate right singular vector $\V{v}$ is redundantly owned on each processor, and the approximate left singular vector $\V{u}$ is distributed across processors.
+Each processor can compute its local $\V[\hat]{u}$ from $\V{v}$ without communication and use the result for its contribution to $\V{v}=\M{A}^\Tra \V{u}$.
+An all-reduce collective is used to obtain a copy of $\V{v}$ on every processor for the next iteration, and the norm is redundantly computed without further communication.
+Again, we use a fixed number of iterations for benchmarking, but the more practical stopping criterion is based on the relative change in $\sigma$.
+
 \begin{algorithm}
@@
 	\EndFunction
-	\Ensure{$\sigma \approx \sigma_1(\M{A})$ is  redundantly owned by all procs}
+	\Ensure{$\sigma \approx \sigma_1^2(\M{A})$ is  redundantly owned by all procs}
 \end{algorithmic}
 \end{algorithm}
-
-\begin{figure}
-\input{fig/split}
-\caption{Parallel splitting using Rank-2 NMF \GB{add more to caption and possibly color code W and H}}
-\label{fig:split}
-\end{figure}
 

### 2020-07-14 latex/paper.tex
finished first draft of analysis

@@
 	\item PLANC
+	\item communication model and collectives
 \end{itemize}
@@
 
+Each iteration of \Cref{alg:parrank2nmf} incurs the same cost, so we analyze per-iteration computation and communication costs.
+We first consider the cost of the Rank-2 NNLS solves, which are local computations.
+In the notation of \Cref{alg:r2nnls}, matrix $\M{G}$ is $2\times 2$, so solving the unconstrained system (via Cholesky decomposition) and then choosing between single-positive-variables solutions if necessary requires constant time per row of $\M{C}$.
+Thus, the cost of \cref{alg:r2nnls} is proportional to the number of rows of the first input matrix.
+In the context of \Cref{alg:parrank2nmf}, the per-iteration computational cost of rank-2 solves is then $O((m+n)/p)$.
+The other local computations are the matrix multiplications $\M[\hat]{W}^\Tra \M[\hat]{W}$ and $\M[\hat]{H}^\Tra \M[\hat]{H}$, which also amount to $O((m+n)/p)$ flops, and $\M[\hat]{A}^\Tra \M[\hat]{W}$ and $\M[\hat]{A} \M{H}$, which require $O(mn/p)$ flops because they involve the data matrix.
+Thus, the computation cost is $\gamma \cdot O((mn+m+n)/p)$ and typically dominated by the multiplications involving $\M{A}$.
+
+There are four communication collectives each iteration, and each involves all $p$ processors.
+The two all-reduce collectives to compute the Gram matrices of the factor matrices involve $2\times 2$ matrices and incur a communication cost of $(\beta + \alpha) \cdot O(\log p)$.
+The reduce-scatter and all-gather collectives involve $n\times 2$ matrices (the size of $\M{H}$) and require $\beta \cdot O(n) + \alpha \cdot O(\log p)$ in communication cost (we ignore the computation cost of the reduce-scatter because it is typically dominated by the bandwidth cost).
+If the algorithm performs $\imath$ iterations, the overall cost of \Cref{alg:parrank2nmf} is
+\begin{equation}
+\label{eq:r2nmfcost}
+\gamma \cdot O\left( \frac{\imath mn}{p} \right) + \beta \cdot O(\imath n) + \alpha \cdot O(\imath \log p).
+\end{equation}
+
 \paragraph{\emph{Parallel Power Method}}
 
+Similar to the previous analysis, we consider a single iteration of the power method.
+The local computation is dominated by two matrix-vector products involving the local data matrix of size $O(mn/p)$ words, incurring $O(mn/p)$ flops.
+The single communication collective is an all-reduce of the approximate right singular vector, which is of size $n$, incurring $\beta \cdot O(n) + \alpha \cdot O(\log p)$ communication.
+We ignore the $O(n)$ computation cost of normalizing the vector, as it will typically be dominated by the communication cost of the all-reduce.
+Over $\jmath$ iterations, \Cref{alg:parpowmeth} has an overall cost of
+\begin{equation}
+\label{eq:powmethcost}
+\gamma \cdot O\left( \frac{\jmath mn}{p} \right) + \beta \cdot O(\jmath n) + \alpha \cdot O(\jmath \log p).
+\end{equation}
+Note the per-iteration cost of the power method differs by only a constant from the per-iteration cost of Rank-2 NMF.
+Because the power method involves single vectors rather than factor matrices with two columns, its constants are smaller than half the size of their counterparts.
+
 \paragraph{\emph{Hierarchical Clustering}}
+
+To analyze the overall cost of the hierarchical clustering algorithm, we sum the costs over all nodes in the tree.
+Because the shape of the tree is data dependent and affects the overall costs, for the sake of analysis we will analyze only complete levels.
+The number of rows in any node is $m$, the same as the root node, as each splitting corresponds to a partition of the columns.
+Furthermore, because each split is a partition, every column of $\M{A}$ is represented exactly once in every complete level of the tree.
+If we assume that all nodes perform the same number of NMF iterations ($\imath$) and power method iterations $(\jmath)$, then the dominating costs of a node with $\overline{n}$ columns is
+\begin{equation*}
+\label{eq:nodecost}
+\gamma \cdot O\left( \frac{(\imath+\jmath) m\overline{n}}{p} \right) + \beta \cdot O((\imath+\jmath) \overline{n}) + \alpha \cdot O((\imath+\jmath) \log p).
+\end{equation*}
+Because the sum of the number of columns across any level of the tree is $n$, the cost of the $\ell$th level of the tree is
+\begin{equation*}
+\label{eq:levelcost}
+\gamma \cdot O\left( \frac{(\imath+\jmath) mn}{p} \right) + \beta \cdot O((\imath+\jmath) n) + \alpha \cdot O((\imath+\jmath) 2^\ell \log p).
+\end{equation*}
+Note that the only cost that depends on the level index $\ell$ is the latency cost.
+Summing over levels and assuming the tree is nearly balanced and has height $O(\log k)$ where $k$ is the number of frontier nodes, we obtain an overall cost of \Cref{alg:hiernmf2} of
+\begin{equation*}
+\label{eq:treecost}
+\begin{split}
+\gamma \cdot O\left( \frac{(\imath+\jmath) mn}{p} \log k \right) + \beta \cdot O((\imath+\jmath) n \log k) \\ + \alpha \cdot O((\imath+\jmath) k \log p).
+\end{split}
+\end{equation*}
+
+We see that the computational cost is logarithmic in $k$ and perfectly load balanced.
+If the overall running time is dominated by the computation (and in particular the matrix multiplications involving $\M{A}$), we expect near-perfect strong scaling.
+The bandwidth cost is also logarithmic in $k$ but does not scale with the number of processors.
+The latency cost grows most quickly with the target number of clusters $k$ but is also independent of the matrix dimensions $m$ and $n$.
 

### 2020-07-14 latex/fig/dc.tex
DC Mall Figure

\begin{tikzpicture}
    \node[inner sep=0pt] (original) at (0,0)
        {\includegraphics[height=0.20\textwidth]{data/DC/0.png}};
        \node[inner sep=0pt] (1) at (-3,-2)
            {\includegraphics[height=0.20\textwidth]{data/DC/1.png}};
            \node[inner sep=0pt] (3) at (-1.6,-4)
                {\includegraphics[height=0.20\textwidth]{data/DC/3.png}};
                \node[inner sep=0pt] (7) at (-2.5,-6)
                    {\includegraphics[height=0.20\textwidth]{data/DC/7.png}};
                \node[inner sep=0pt] (8) at (-0.7,-6)
                    {\includegraphics[height=0.20\textwidth]{data/DC/8.png}};
            \node[inner sep=0pt] (4) at (-4.4,-4)
                {\includegraphics[height=0.20\textwidth]{data/DC/4.png}};
        \node[inner sep=0pt] (2) at (3,-2)
            {\includegraphics[height=0.20\textwidth]{data/DC/2.png}};
            \node[inner sep=0pt] (5) at (4.4,-4)
                {\includegraphics[height=0.20\textwidth]{data/DC/3.png}};
            \node[inner sep=0pt] (6) at (1.6,-4)
                {\includegraphics[height=0.20\textwidth]{data/DC/4.png}};
                \node[inner sep=0pt] (13) at (0.7,-6)
                    {\includegraphics[height=0.20\textwidth]{data/DC/3.png}};
                \node[inner sep=0pt] (14) at (2.5,-6)
                    {\includegraphics[height=0.20\textwidth]{data/DC/4.png}};
\end{tikzpicture}

### 2020-07-14 latex/fig/dc.tex
dc arrows

@@
     \node[inner sep=0pt] (original) at (0,0)
-        {\includegraphics[height=0.20\textwidth]{data/DC/0.png}};
+        {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/0.png}};
+        \draw[thick,->] (-0.4,0.8) .. controls (-3,0.8) .. (-3,-0.4);
         \node[inner sep=0pt] (1) at (-3,-2)
-            {\includegraphics[height=0.20\textwidth]{data/DC/1.png}};
+            {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/1.png}};
+            \draw[thick,->] (-2.6,-1.2) .. controls (-1.6,-1.2) .. (-1.6,-2.4);
             \node[inner sep=0pt] (3) at (-1.6,-4)
-                {\includegraphics[height=0.20\textwidth]{data/DC/3.png}};
+                {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/3.png}};
+                \draw[thick,->] (-2,-3.2) .. controls (-2.5,-3.2) .. (-2.5,-4.4);
                 \node[inner sep=0pt] (7) at (-2.5,-6)
-                    {\includegraphics[height=0.20\textwidth]{data/DC/7.png}};
+                    {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/7.png}};
+                \draw[thick,->] (-1.2,-3.2) .. controls (-0.7,-3.2) .. (-0.7,-4.4);
                 \node[inner sep=0pt] (8) at (-0.7,-6)
-                    {\includegraphics[height=0.20\textwidth]{data/DC/8.png}};
+                    {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/8.png}};
+            \draw[thick,->] (-3.4,-1.2) .. controls (-4.4,-1.2) .. (-4.4,-2.4);
             \node[inner sep=0pt] (4) at (-4.4,-4)
-                {\includegraphics[height=0.20\textwidth]{data/DC/4.png}};
+                {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/4.png}};
+        \draw[thick,->] (0.4,0.8) .. controls (3,0.8) .. (3,-0.4);
         \node[inner sep=0pt] (2) at (3,-2)
-            {\includegraphics[height=0.20\textwidth]{data/DC/2.png}};
+            {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/2.png}};
+            \draw[thick,->] (3.4,-1.2) .. controls (4.4,-1.2) .. (4.4,-2.4);
             \node[inner sep=0pt] (5) at (4.4,-4)
-                {\includegraphics[height=0.20\textwidth]{data/DC/3.png}};
+                {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/3.png}};
+            \draw[thick,->] (2.6,-1.2) .. controls (1.6,-1.2) .. (1.6,-2.4);
             \node[inner sep=0pt] (6) at (1.6,-4)
-                {\includegraphics[height=0.20\textwidth]{data/DC/4.png}};
+                {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/4.png}};
+                \draw[thick,->] (1.2,-3.2) .. controls (0.7,-3.2) .. (0.7,-4.4);
                 \node[inner sep=0pt] (13) at (0.7,-6)
-                    {\includegraphics[height=0.20\textwidth]{data/DC/3.png}};
+                    {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/3.png}};
+                    \draw[thick,->] (2,-3.2) .. controls (2.5,-3.2) .. (2.5,-4.4);
                 \node[inner sep=0pt] (14) at (2.5,-6)
-                    {\includegraphics[height=0.20\textwidth]{data/DC/4.png}};
+                    {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/4.png}};
 \end{tikzpicture}

### 2020-07-14 distnmf/distio.hpp
mpi-io

@@ -44,6 +44,64 @@ namespace planc {
         int * ccounts;
 
         const iodistributions m_distio;
+
+        void read_dist_tensor(const std::string filename) {
+          // all processes reading the file_name.info file.
+          std::string filename_no_extension =
+              filename.substr(0, filename.find_last_of("."));
+          int modes = local_tensor.modes();
+          int global_rows = 3145728;
+          int global_cols = 33126;
+          int gsizes[] = {global_rows, global_cols};
+          int lsizes[] = {global_rows/p, global_cols};
+          int *starts = new int[modes];
+          UVEC tmp_proc_grids = this->m_mpicomm.proc_grids();
+          for (int i = 0; i < 2; i++) {
+            starts[i] =
+                startidx(gsizes[i], tmp_proc_grids[i], this->m_mpicomm.fiber_rank(i));
+          }
+          MPI_Barrier(MPI_COMM_WORLD);
+          // Now each of write to the file.
+          // Create the datatype associated with this layout
+
+          MPI_Datatype view;
+          MPI_Type_create_subarray(modes, gsizes, lsizes, starts, MPI_ORDER_FORTRAN,
+                                  MPI_DOUBLE, &view);
+          MPI_Type_commit(&view);
+
+          // Open the file
+          MPI_File fh;
+          int ret =
+              MPI_File_open(MPI_COMM_WORLD, filename.c_str(),
+                            MPI_MODE_CREATE | MPI_MODE_WRONLY, MPI_INFO_NULL, &fh);
+          if (ISROOT && ret != MPI_SUCCESS) {
+            DISTPRINTINFO("Error: Could not open file " << filename << std::endl);
+          }
+
+          // Set the view
+          MPI_Offset disp = 0;
+          MPI_File_set_view(fh, disp, MPI_DOUBLE, view, "native", MPI_INFO_NULL);
+
+          // Write the file
+          int count = lsizes[0]*lsizes[1];
+          double *input_array = (double *)malloc(sizeof(float)*count);
+          assert(count <= std::numeric_limits<int>::max());
+          MPI_Status status;
+          ret = MPI_File_read_all(fh, input, count, MPI_DOUBLE,
+                                  &status);
+          if (ret != MPI_SUCCESS) {
+            DISTPRINTINFO("Error: Could not write file " << filename << std::endl);
+          }
+          // Close the file
+          MPI_File_close(&fh);
+          // Free the datatype
+          MPI_Type_free(&view);
+          // free the allocated memories
[…]

### 2020-07-14 distnmf/distio.hpp
fixed tensor to matrix

@@ -45,27 +45,25 @@ namespace planc {
 
         const iodistributions m_distio;
 
-        void read_dist_tensor(const std::string filename) {
+        void read_dist_matrix(const std::string filename) {
           // all processes reading the file_name.info file.
           std::string filename_no_extension =
               filename.substr(0, filename.find_last_of("."));
-          int modes = local_tensor.modes();
+          int pr = m_mpicomm.pr();
+          int pc = m_mpicomm.pc();
+          int rrank = m_mpicomm.row_rank();
+          int crank = m_mpicomm.col_rank();
           int global_rows = 3145728;
           int global_cols = 33126;
           int gsizes[] = {global_rows, global_cols};
-          int lsizes[] = {global_rows/p, global_cols};
-          int *starts = new int[modes];
-          UVEC tmp_proc_grids = this->m_mpicomm.proc_grids();
-          for (int i = 0; i < 2; i++) {
-            starts[i] =
-                startidx(gsizes[i], tmp_proc_grids[i], this->m_mpicomm.fiber_rank(i));
-          }
+          int lsizes[] = {itersplit(global_rows,pr,rrank), itersplit(global_cols,pc,crank)};
+          int starts[] = {startidx(global_rows,pr,rrank),  startidx(global_cols,pc,crank)};
           MPI_Barrier(MPI_COMM_WORLD);
           // Now each of write to the file.
           // Create the datatype associated with this layout
 
           MPI_Datatype view;
-          MPI_Type_create_subarray(modes, gsizes, lsizes, starts, MPI_ORDER_FORTRAN,
+          MPI_Type_create_subarray(2, gsizes, lsizes, starts, MPI_ORDER_FORTRAN,
                                   MPI_DOUBLE, &view);
           MPI_Type_commit(&view);
 
@@ -87,7 +85,7 @@ namespace planc {
           double *input_array = (double *)malloc(sizeof(float)*count);
           assert(count <= std::numeric_limits<int>::max());
           MPI_Status status;
-          ret = MPI_File_read_all(fh, input, count, MPI_DOUBLE,
+          ret = MPI_File_read_all(fh, &m_A, count, MPI_DOUBLE,
                                   &status);
           if (ret != MPI_SUCCESS) {
             DISTPRINTINFO("Error: Could not write file " << filename << std::endl);
@@ -96,10 +94,6 @@ namespace planc {
           MPI_File_close(&fh);
           // Free the datatype
           MPI_Type_free(&view);
[…]

### 2020-07-14 distnmf/distio.hpp
read instead of write

@@ -71,7 +71,7 @@ namespace planc {
           MPI_File fh;
           int ret =
               MPI_File_open(MPI_COMM_WORLD, filename.c_str(),
-                            MPI_MODE_CREATE | MPI_MODE_WRONLY, MPI_INFO_NULL, &fh);
+                            MPI_MODE_RDONLY, MPI_INFO_NULL, &fh);
           if (ISROOT && ret != MPI_SUCCESS) {
             DISTPRINTINFO("Error: Could not open file " << filename << std::endl);
           }

### 2020-07-15 distnmf/distio.hpp
matrix preallocation

@@ -53,6 +53,7 @@ namespace planc {
           int pc = m_mpicomm.pc();
           int rrank = m_mpicomm.row_rank();
           int crank = m_mpicomm.col_rank();
+          int rank = m_mpicomm.rank();
           int global_rows = 3145728;
           int global_cols = 33126;
           int gsizes[] = {global_rows, global_cols};
@@ -80,15 +81,15 @@ namespace planc {
           MPI_Offset disp = 0;
           MPI_File_set_view(fh, disp, MPI_DOUBLE, view, "native", MPI_INFO_NULL);
 
-          // Write the file
+          // Read the file
+          m_A.zeros(lsizes[0],lsizes[1]);
           int count = lsizes[0]*lsizes[1];
-          double *input_array = (double *)malloc(sizeof(float)*count);
           assert(count <= std::numeric_limits<int>::max());
           MPI_Status status;
-          ret = MPI_File_read_all(fh, &m_A, count, MPI_DOUBLE,
+          ret = MPI_File_read_all(fh, m_A.memptr(), count, MPI_DOUBLE,
                                   &status);
           if (ret != MPI_SUCCESS) {
-            DISTPRINTINFO("Error: Could not write file " << filename << std::endl);
+            DISTPRINTINFO("Error: Could not read file " << filename << std::endl);
           }
           // Close the file
           MPI_File_close(&fh);

### 2020-07-15 latex/plots.tex
included strong scaling plots for synthetic data

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_strongscaling.pdf}
\caption{Relative Time Breakdown for Rank-2 NMF \cref{alg:parrank2nmf} Strong Scaling at Level-0}
\label{fig:synrank2strongscaling}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_speedup.pdf}
\caption{Strong Scaling Speedup for Rank-2 NMF \cref{alg:parrank2nmf} at Level-0}
\label{fig:synrank2speedup}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hierarchical_speedup.pdf}
\caption{Strong Scaling Speedup for Hierarchical Clustering. The total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
\label{fig:synhierspeedup}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hier_strongscaling.pdf}
\caption{Relative Time Breakdown for Hierarchical Clustering. The plot uses the total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
\label{fig:synhierstrongscaling}
\end{center}
\end{figure}

### 2020-07-15 latex/paper.tex
DC Mall Figure
included strong scaling plots for synthetic data
added description of communication model

@@
 
-\section{Preliminaries}
+\section{Preliminaries and Related Work}
 \label{sec:prelim}
 
-\begin{itemize}
-	\item NMF, BCD, NNLS
-	\item PLANC
-	\item communication model and collectives
-\end{itemize}
+\GB{use Kuang's related work for NMF (2016 paper for divide-and-conquer NMF in JGO: \url{https://link.springer.com/article/10.1007/s10898-017-0515-z}), use PPoPP/TKDE/arXiv of ours and also Liavas, pyramid NMF: \doi{10.1109/IJCNN.2003.1224004}}
+
+\subsection{NMF}
+
+\GB{general NMF, BCD/AO algorithms, NNLS}
+
+\subsection{Parallel NMF}
+
+\GB{PLANC and other related work}
+
+\subsection{Communication Model}
+
+We use the $\alpha$-$\beta$-$\gamma$ model \cite{TRG05,CH+07,BCDH+14} for analysis of distributed-memory parallel algorithms. 
+In this model, the cost of sending a single message of $n$ words of data between two processors is $\alpha + \beta \cdot n$, so that $\alpha$ represents the latency cost of the message and $\beta$ represents the bandwidth cost of each word in the message.
+The $\gamma$ parameter represents the computational cost of a single floating point operation (flop).
+In this simplified communication model, we ignore contention in the network, assuming in effect a fully connected network, and other limiting factors in practice such as the number of hops between nodes and the network injection rate \cite{GOS16}.
+We let $p$ represent the number of processors available on the machine.
+
+All of the interprocessor communication in the algorithms presented in this work are encapsulated in collective communication operations that involve the full set of processors.
+Algorithms for implementing the collective operations are built out of pairwise send and receive operations, and we assume the most efficient algorithms are used in our analysis \cite{TRG05,CH+07}.
+The collectives used in our algorithms are all-reduce, all-gather, and reduce-scatter.
+In an all-reduce, all processors start out with the same amount of data and all end with a copy of the same result, which is in our case a sum of all the inputs (and the same size as a single input).
+The cost of an all-reduce of size $n$ words is $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n)$ for $n>p$ and $(\alpha+\beta) \cdot O(\log p)$ for $n<p$.
+In an all-gather, all processors start out with separate data and all end with a copy of the same result, which is the union of all the input data.
+If each processor starts with $n/p$ data and ends with $n$ data, the cost of the all-gather is $\alpha \cdot O(\log p) + \beta \cdot O(n)$.
+In a reduce-scatter, all processors start out with the same amount of data and all end with a subset of the result, which is in our case a sum of all the inputs (and is smaller than its input).
+If each processor starts with $n$ data and ends with $n/p$ data, the cost of the reduce-scatter is $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n)$.
+In the case of all-reduce and reduce-scatter, the computational cost is typically dominated by the bandwidth cost because $\beta \gg \gamma$.
 
@@
 
+\GB{track NNLS of $\M{W}$ term even though it's lower order}
+
 To analyze the overall cost of the hierarchical clustering algorithm, we sum the costs over all nodes in the tree.
@@
 \subsubsection{Hyperspectral Imaging}
+\begin{figure}
+\input{fig/dc}
+\caption{Hierarchical Cluster from DC Mall Image}
+\label{fig:dc}
+\end{figure}
 

### 2020-07-15 latex/plots.tex

@@
 \end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_sequential_level_breakdown.pdf}
+\caption{Sequential Breakdown plot for levels 0 to 5. y-axis is absolute wall clock time in seconds}
+\label{fig:seqlevelbreakdown}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_parallel_level_breakdown.pdf}
+\caption{Parallel Breakdown plot for levels 0 to 5 on 40 nodes. y-axis is absolute wall clock time in seconds}
+\label{fig:parallellevelbreakdown}
+\end{center}
+\end{figure}

### 2020-07-15 latex/paper.tex
added nnls cost lower order term to computation analysis

@@
 
-\GB{use Kuang's related work for NMF (2016 paper for divide-and-conquer NMF in JGO: \url{https://link.springer.com/article/10.1007/s10898-017-0515-z}), use PPoPP/TKDE/arXiv of ours and also Liavas, pyramid NMF: \doi{10.1109/IJCNN.2003.1224004}}
+\GB{use Kuang's related work for NMF (2016 paper for divide-and-conquer NMF in JGO \cite{DKDP17}, use PPoPP/TKDE/arXiv of ours and also Liavas, pyramid NMF: \doi{10.1109/IJCNN.2003.1224004}}
 
@@
 
-\GB{general NMF, BCD/AO algorithms, NNLS}
+\GB{general NMF, BCD/AO algorithms, NNLS, R2-HierNMF \cite{KP13,GKP15}, DC-NMF \cite{DKDP17}}
 
@@
 
-\GB{PLANC and other related work}
+\GB{PLANC (PPoPP/TKDE/arXiv) and other related work like Liavas}
 
@@
 Thus, the computation cost is $\gamma \cdot O((mn+m+n)/p)$ and typically dominated by the multiplications involving $\M{A}$.
+We track the lower order terms corresponding to NNLS solves because their hidden constants are larger than that of the dominating term.
 
@@
 \label{eq:r2nmfcost}
-\gamma \cdot O\left( \frac{\imath mn}{p} \right) + \beta \cdot O(\imath n) + \alpha \cdot O(\imath \log p).
+\gamma \cdot O\left( \frac{\imath (mn+m+n)}{p} \right) + \beta \cdot O(\imath n) + \alpha \cdot O(\imath \log p).
 \end{equation}
@@
 
-\GB{track NNLS of $\M{W}$ term even though it's lower order}
-
 To analyze the overall cost of the hierarchical clustering algorithm, we sum the costs over all nodes in the tree.
@@
 \label{eq:nodecost}
-\gamma \cdot O\left( \frac{(\imath+\jmath) m\overline{n}}{p} \right) + \beta \cdot O((\imath+\jmath) \overline{n}) + \alpha \cdot O((\imath+\jmath) \log p).
+\begin{split}
+\gamma \cdot O\left( \frac{(\imath+\jmath) m\overline{n} + \imath (m+\overline{n}) }{p} \right) + \beta \cdot O((\imath+\jmath) \overline{n}) \\ + \alpha \cdot O((\imath+\jmath) \log p).
+\end{split}
 \end{equation*}
@@
 \label{eq:levelcost}
-\gamma \cdot O\left( \frac{(\imath+\jmath) mn}{p} \right) + \beta \cdot O((\imath+\jmath) n) + \alpha \cdot O((\imath+\jmath) 2^\ell \log p).
+\begin{split}
+\gamma \cdot O\left( \frac{(\imath+\jmath) mn + \imath m 2^\ell}{p} \right) + \beta \cdot O((\imath+\jmath) n) \\ + \alpha \cdot O((\imath+\jmath) 2^\ell \log p).
+\end{split}
 \end{equation*}
@@
 \begin{split}
-\gamma \cdot O\left( \frac{(\imath+\jmath) mn}{p} \log k \right) + \beta \cdot O((\imath+\jmath) n \log k) \\ + \alpha \cdot O((\imath+\jmath) k \log p).
+\gamma \cdot O\left( \frac{(\imath+\jmath) mn}{p}  \log k + \frac{\imath mk}{p} \right) + \beta \cdot O((\imath+\jmath) n \log k) \\ + \alpha \cdot O((\imath+\jmath) k \log p).
 \end{split}
@@
 
-We see that the computational cost is logarithmic in $k$ and perfectly load balanced.
+We see that the leading order computational cost is logarithmic in $k$ and perfectly load balanced.
 If the overall running time is dominated by the computation (and in particular the matrix multiplications involving $\M{A}$), we expect near-perfect strong scaling.
@@
 
+\include{plots}
+
 \begin{itemize}

### 2020-07-15 latex/paper.tex

@@
 
-\GB{use Kuang's related work for NMF (2016 paper for divide-and-conquer NMF in JGO \cite{DKDP17}, use PPoPP/TKDE/arXiv of ours and also Liavas, pyramid NMF: \doi{10.1109/IJCNN.2003.1224004}}
+\GB{use Kuang's related work for NMF (2016 paper for divide-and-conquer NMF in JGO \cite{DKDP17}, use PPoPP \cite{KBP16}/TKDE \cite{KBP17}/arXiv \cite{EH+19-TR} of ours and also Liavas \cite{LK+17b}, pyramid NMF: \doi{10.1109/IJCNN.2003.1224004}}
 
@@
 
-\GB{general NMF, BCD/AO algorithms, NNLS, R2-HierNMF \cite{KP13,GKP15}, DC-NMF \cite{DKDP17}}
+\GB{general NMF, BCD/AO algorithms \cite{KHP14}, NNLS \cite{KP11}, R2-HierNMF \cite{KP13,GKP15}, DC-NMF \cite{DKDP17}}
 
@@
 Solving for $\M{W}$ can be done using inputs $\M{C}=\M{A}\M{H}$ and $\M{G}=\M{H}^\Tra\M{H}$.
-\GB{To avoid two passes through the data, we could combine the solve and the update if necessary.
-It might also help to pre-compute the inverse of $\M{G}$...
-Also Lawton, please check the details of the pseudocode!}
 

### 2020-07-15 latex/paper.tex
finished draft of preliminaries

@@
 
-\GB{use Kuang's related work for NMF (2016 paper for divide-and-conquer NMF in JGO \cite{DKDP17}, use PPoPP \cite{KBP16}/TKDE \cite{KBP17}/arXiv \cite{EH+19-TR} of ours and also Liavas \cite{LK+17b}, pyramid NMF: \doi{10.1109/IJCNN.2003.1224004}}
-
 \subsection{NMF}
 
-\GB{general NMF, BCD/AO algorithms \cite{KHP14}, NNLS \cite{KP11}, R2-HierNMF \cite{KP13,GKP15}, DC-NMF \cite{DKDP17}}
+The NMF constrained optimization problem
+$$\min_{\M{W},\M{H}\geq \M{0}} \|\M{A} - \M{W}\M{H}^\Tra\|_2$$
+is nonlinear and nonconvex, and various optimization techniques can be used to approximately solve it.
+A popular approach is to use alternating optimization of the two factor matrices because each subproblem is a nonnegative least squares (NNLS) problem, which is convex and can be solved exactly.
+Many block coordinate descent (BCD) approaches are possible \cite{KHP14}, and one 2-block BCD algorithm that solves the NNLS subproblems exactly is block principal pivoting \cite{KP11}.
+This NNLS algorithm is an active-set-like method that determines the sets of entries in the solution vectors that are zero and those that are positive through an iterative but finite process.
+
+When the rank of the factorization (the number of columns of $\M{W}$ and $\M{H}$) is 2, the NNLS subproblems can be solved much more quickly because the number of possible active sets is only 4.
+As explained in more detail in \Cref{sec:r2nmf}, the optimal solution across the 4 possibilities can be determined efficiently to solve the NNLS subproblem more quickly than general-rank approaches like block principal pivoting.
+Because of the relative ease of solving the NMF problem for the rank-2 case, Kuang and Park \cite{KP13} propose a recursive method to use a rank-2 NMF to partition the input data into 2 parts, whereby each part can be further partitioned via rank-2 NMF of the corresponding original data.
+This approach yields a hierarchical factorization, potentially uncovering more global structure of the input data and allowing for better scalability of the algorithm to large NMF ranks.
+
+The hierarchical rank-2 NMF method has been applied to document clustering \cite{KP13} and hyperspectral image segmentation \cite{GKP15}.
+The leaves of the tree also yield a set of column vectors that can be aggregated into an approximate $\M{W}$ factor (ignoring their hierarchical structure).
+Using this factor matrix to initialize a higher-rank NMF computation leads to quick convergence and overall faster performance than initializing NMF with random data; this approach is known as Divide-and-Conquer NMF \cite{DKDP17}.
+We focus in this paper on parallelizing the hierarchical algorithms proposed by Kuang and Park \cite{KP13} and Gillis et al. \cite{GKP15}.
 
@@
 
-\GB{PLANC (PPoPP/TKDE/arXiv) and other related work like Liavas}
+Scaling algorithms for NMF to large data often requires parallelization in order to fit the data across the memories of multiple compute nodes or speed up the computation to complete in reasonable time.
+Parallelizations of multiple optimization approaches have been proposed for general NMF.
+In particular, we build upon the work of Kannan et al.
+\cite{KBP16,KBP17,EH+19-TR} and the open-source library PLANC, designed for nonnegative matrix and tensor factorizations of dense and sparse data.
+In this parallelization, the alternating optimization approach is employed with various options for the algorithm used to (approximately) solve the NNLS subproblems.
+The efficiency of the parallelization is based on scalable algorithms for the parallel matrix multiplications involved in all NNLS algorithms; these algorithms are based on Cartesian distributions of the input matrix across 1D or 2D processor grids.
 
@@
 
-\section{Algorithm}
+\section{Algorithms}
 
-\subsection{Sequential Algorithm}
+\subsection{Sequential Algorithms}
 
 \subsubsection{Rank-2 NMF}
+\label{sec:r2nmf}
 

### 2020-07-15 latex/plots.tex
weak scaling plot
realworld rank 2 plots

@@
 \end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_weakscaling.pdf}
+\caption{Relative Time Breakdown for Rank-2 NMF \cref{alg:parrank2nmf} Weak Scaling at Level-0}
+\label{fig:synrank2weakscaling}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_strongscaling.pdf}
+\caption{Relative Time Breakdown for Rank-2 NMF \cref{alg:parrank2nmf} Strong Scaling at Level-0 on Realworld Data}
+\label{fig:rwrank2strongscaling}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_speedup.pdf}
+\caption{Strong Scaling Speedup for Rank-2 NMF \cref{alg:parrank2nmf} at Level-0 on Realworld Data}
+\label{fig:rwrank2speedup}
+\end{center}
+\end{figure}

### 2020-07-15 latex/plots.tex
realworld hierarchical and level breakdown plots

@@
 \end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/realworld_hierarchical_speedup.pdf}
+\caption{Strong Scaling Speedup for Hierarchical Clustering on Realworld Data. The total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
+\label{fig:rwhierspeedup}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/realworld_hier_strongscaling.pdf}
+\caption{Relative Time Breakdown for Hierarchical Clustering on Realworld. The plot uses the total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
+\label{fig:rwhierstrongscaling}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/realworld_sequential_level_breakdown.pdf}
+\caption{Parallel Breakdown plot for levels 0 to 4 on 10 nodes. y-axis is absolute wall clock time in seconds}
+\label{fig:rwseqlevelbreakdown}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/realworld_parallel_level_breakdown.pdf}
+\caption{Parallel Breakdown plot for levels 0 to 5 on 80 nodes. y-axis is absolute wall clock time in seconds}
+\label{fig:rwparallellevelbreakdown}
+\end{center}
+\end{figure}

### 2020-07-15 latex/paper.tex
organized plots

@@
 
-\GB{ask Haesun to write intro paragraph motivating hierarchical clustering but noting the limitations of time and data size}
+\GB{Prof Park, can you draft an introduction?
+We are focusing on the dense case and using HSI and image clustering as applications.
+Section 4 is still unfinished, but nearly all of the performance plots are there.
+Feel free to leave placeholders for details that aren't obvious, we can clean them up Wednesday night.}
 
@@
 	flattened from $8940\times 747\times 224$ to $224\times 6678180$. Unlike with document clustering, hyperspectral image sizes grow in pixels
-	and not spectral bands. So, for both images, large scaling is impossible given our current data distribution.
+	and not spectral bands. So, for both images, large scaling is impossible given our current data distribtion.
 	\LM{I think this is awkward and I'm not sure if we need to include it}
@@
 
-\GB{Lawton, can you gather this data, Ramki can you visualize it?}
-\begin{itemize}
-	\item compare 1D distribution scaling with optimal 2D grid scaling
-	\item use synthetic short-and-fat dense and squarish sparse
-\end{itemize}
+\GB{4 plots: 2 data sets with speedup/breakdown}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_strongscaling.pdf}
+\caption{Relative Time Breakdown for Rank-2 NMF \cref{alg:parrank2nmf} Strong Scaling at Level-0}
+\label{fig:synrank2strongscaling}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_speedup.pdf}
+\caption{Strong Scaling Speedup for Rank-2 NMF \cref{alg:parrank2nmf} at Level-0}
+\label{fig:synrank2speedup}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_strongscaling.pdf}
+\caption{Relative Time Breakdown for Rank-2 NMF \cref{alg:parrank2nmf} Strong Scaling at Level-0 on Realworld Data}
+\label{fig:rwrank2strongscaling}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_speedup.pdf}
+\caption{Strong Scaling Speedup for Rank-2 NMF \cref{alg:parrank2nmf} at Level-0 on Realworld Data}
+\label{fig:rwrank2speedup}
+\end{center}
+\end{figure}
+
+\subsubsection{Hierarchical Clustering Strong Scaling}
+
+\GB{4 plots: 2 data sets with speedup/breakdown}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hierarchical_speedup.pdf}
+\caption{Strong Scaling Speedup for Hierarchical Clustering. The total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
+\label{fig:synhierspeedup}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hier_strongscaling.pdf}
+\caption{Relative Time Breakdown for Hierarchical Clustering. The plot uses the total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
+\label{fig:synhierstrongscaling}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/realworld_hierarchical_speedup.pdf}
+\caption{Strong Scaling Speedup for Hierarchical Clustering on Realworld Data. The total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
+\label{fig:rwhierspeedup}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/realworld_hier_strongscaling.pdf}
+\caption{Relative Time Breakdown for Hierarchical Clustering on Realworld. The plot uses the total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
+\label{fig:rwhierstrongscaling}
+\end{center}
+\end{figure}
 
@@
 
-\GB{Ramki, can you add a bar plot to show time breakdown across complete levels?  Let's do Aviris and PubMed; just use levels data file}
+\GB{3 plots: for synthetic level breakdown for 1 node and for 40 nodes; for realworld just for 80 nodes
 
-\subsubsection{Weak Scaling}
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_sequential_level_breakdown.pdf}
+\caption{Sequential Breakdown plot for levels 0 to 5. y-axis is absolute wall clock time in seconds}
+\label{fig:seqlevelbreakdown}
+\end{center}
+\end{figure}
 
-\GB{maybe this section is lower priority}
-\begin{itemize}
-	\item dense and sparse
-\end{itemize}
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_parallel_level_breakdown.pdf}
+\caption{Parallel Breakdown plot for levels 0 to 5 on 40 nodes. y-axis is absolute wall clock time in seconds}
+\label{fig:parallellevelbreakdown}
+\end{center}
+\end{figure}
 
-\subsubsection{Strong Scaling}
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/realworld_parallel_level_breakdown.pdf}
+\caption{Parallel Breakdown plot for levels 0 to 5 on 80 nodes. y-axis is absolute wall clock time in seconds}
+\label{fig:rwparallellevelbreakdown}
+\end{center}
+\end{figure}
 
-\include{plots}
-
-\begin{itemize}
-	\item big dense synthetic
-	\item big sparse synthetic
-	\item Aviris
-	\item PubMed
-\end{itemize}
-
-\subsection{Clustering}
+\subsection{DC Clustering Results}
 
@@
 \end{figure}
-
-\GB{Lawton, show some visualization of the tree}
-
-\subsubsection{Document Corpus}
-
-\GB{Lawton, show a visualization of top words in tree}
 

### 2020-07-15 latex/paper.tex
scaled the dc figure and fixed latex bug
adding notes for experimental analysis

@@
 
+\GB{specify that $\imath=\jmath=100$ in all experiments, also what do we set $k$ to be?  fixed number for each dataset?  any other experimental parameters we should specify?}
+
 \subsection{Datasets}
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_strongscaling.pdf}
-\caption{Relative Time Breakdown for Rank-2 NMF \cref{alg:parrank2nmf} Strong Scaling at Level-0}
-\label{fig:synrank2strongscaling}
-\end{center}
-\end{figure}
+\begin{itemize}
+	\item theory says...
+	\item Figs 3 and 5 show performance relative to the smallest number of nodes required to store data and factor matrices
+	\item strong scaling is nearly linear, see 49X (check number) on 50 nodes for synthetic data and 7.3X (check number) on 80 nodes for image data
+	\item Figs 4 and 6 show time breakdown, explain scaling: MM scales perfectly because it's mem-bw bound and the local problem size decreases linearly, bandwidth cost is fixed relative to $p$, so relative cost grows as we scale; comm is bandwidth bound because we see all-gather and reduce-scatter dominate all-reduce
+\end{itemize}
 
@@
 \begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_strongscaling.pdf}
-\caption{Relative Time Breakdown for Rank-2 NMF \cref{alg:parrank2nmf} Strong Scaling at Level-0 on Realworld Data}
-\label{fig:rwrank2strongscaling}
+\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_strongscaling.pdf}
+\caption{Relative Time Breakdown for Rank-2 NMF \cref{alg:parrank2nmf} Strong Scaling at Level-0}
+\label{fig:synrank2strongscaling}
 \end{center}
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_strongscaling.pdf}
+\caption{Relative Time Breakdown for Rank-2 NMF \cref{alg:parrank2nmf} Strong Scaling at Level-0 on Realworld Data}
+\label{fig:rwrank2strongscaling}
+\end{center}
+\end{figure}
+
 \subsubsection{Hierarchical Clustering Strong Scaling}
 
-\GB{4 plots: 2 data sets with speedup/breakdown}
+\GB{Do these breakdowns include the score function?  actually this may explain why all-reduce costs more, it also has $O(n)$ data for score...}
+
+\GB{4 plots: 2 data sets with speedup/breakdown, note that we include time for complete levels, ignoring nodes in incomplete levels for easier interpretation}
+
+\begin{itemize}
+	\item theory says...t
+	\item Fig 7 shows limit of strong scaling for synthetic data (time increases at 50 nodes compared to 40), Fig 8 gives breakdown and explains
+	\begin{itemize}
+		\item drastic performance dropoff may be artifact of all-gather, which should take about the same time as reduce-scatter, but speedup from 30 to 40 nodes is also slight
+		\item by 50 nodes, about 50\% of time in communication; still mostly dominated by bandwidth cost because ag and rs dominate ar
+		\item note the difference between the scaling of single rank-2 and entire tree: poor scaling must be due to lower levels, which we explore in the next section
+	\end{itemize}
+	\item Fig 9 shows reasonable scaling for entire tree -- speedup drops from 7x to 6x over 80 nodes; Fig 10 shows that MM is still 70\% of run time on 80 nodes, which allows for near-linear scaling
+\end{itemize}
 
@@
 
-\GB{3 plots: for synthetic level breakdown for 1 node and for 40 nodes; for realworld just for 80 nodes
+\GB{3 plots: for synthetic level breakdown for 1 node and for 40 nodes; for realworld just for 80 nodes}
+
+\begin{itemize}
+	\item theory says 
+\end{itemize}
 
@@
 
+\begin{itemize}
+	\item consider sparse case
+	\item consider parallelizing across tree
+	\item consider redistribution for 2D
+\end{itemize}
+
 \section*{Acknowledgment}

### 2020-07-15 latex/paper.tex
added discussion of r2nmf scaling

@@
 In the notation of \Cref{alg:r2nnls}, matrix $\M{G}$ is $2\times 2$, so solving the unconstrained system (via Cholesky decomposition) and then choosing between single-positive-variables solutions if necessary requires constant time per row of $\M{C}$.
-Thus, the cost of \cref{alg:r2nnls} is proportional to the number of rows of the first input matrix.
+Thus, the cost of \Cref{alg:r2nnls} is proportional to the number of rows of the first input matrix.
 In the context of \Cref{alg:parrank2nmf}, the per-iteration computational cost of rank-2 solves is then $O((m+n)/p)$.
@@
 Because the sum of the number of columns across any level of the tree is $n$, the cost of the $\ell$th level of the tree is
-\begin{equation*}
+\begin{equation}
 \label{eq:levelcost}
@@
 \end{split}
-\end{equation*}
+\end{equation}
 Note that the only cost that depends on the level index $\ell$ is the latency cost.
 Summing over levels and assuming the tree is nearly balanced and has height $O(\log k)$ where $k$ is the number of frontier nodes, we obtain an overall cost of \Cref{alg:hiernmf2} of
-\begin{equation*}
+\begin{equation}
 \label{eq:treecost}
@@
 \end{split}
-\end{equation*}
+\end{equation}
 
@@
 
-\GB{specify that $\imath=\jmath=100$ in all experiments, also what do we set $k$ to be?  fixed number for each dataset?  any other experimental parameters we should specify?}
+\GB{specify that $\imath=\jmath=100$ in all experiments, also what do we set $k$ to be?
+fixed number for each dataset?
+any other experimental parameters we should specify?
+could also specify these at the beginning of \Cref{sec:perf}}
 
 \subsection{Datasets}
+
+\GB{need shortcut names for datasets: synthetic, DC, and something better than IMAGE}
 
@@
 \subsection{Performance}
+\label{sec:perf}
 
@@
 
-\GB{4 plots: 2 data sets with speedup/breakdown}
+We perform strong scaling experiments for a single Rank-2 NMF (\Cref{alg:parrank2nmf}) on the synthetic and IMAGE datasets.
+The theory (\Cref{eq:r2nmfcost}) suggests that perfect strong scaling is possible as long as the execution time is dominated by local computation.
+Both the matrix multiplications and NNLS solves scale linearly with $1/p$ (we expect MatMul to dominate), but the bandwidth cost is independent of $p$ and the latency cost increases slightly with $p$.
 
-\begin{itemize}
-	\item theory says...
-	\item Figs 3 and 5 show performance relative to the smallest number of nodes required to store data and factor matrices
-	\item strong scaling is nearly linear, see 49X (check number) on 50 nodes for synthetic data and 7.3X (check number) on 80 nodes for image data
-	\item Figs 4 and 6 show time breakdown, explain scaling: MM scales perfectly because it's mem-bw bound and the local problem size decreases linearly, bandwidth cost is fixed relative to $p$, so relative cost grows as we scale; comm is bandwidth bound because we see all-gather and reduce-scatter dominate all-reduce
-\end{itemize}
+\Cref{fig:synrank2speedup,fig:rwrank2speedup} show performance relative to the smallest number of nodes required to store data and factor matrices.
+The synthetic data is chosen to fit on a single node, while the IMAGE data requires 10 nodes to store the input matrix as well as temporary and output matrices.
+For these data sets, we observe nearly perfect strong scaling, with 47$\times$ speedup on 50 nodes (over 1 node) for synthetic data and 7.1$\times$ speedup on 80 nodes (over 10 nodes) for IMAGE data.
 
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_speedup.pdf}
-\caption{Strong Scaling Speedup for Rank-2 NMF \cref{alg:parrank2nmf} at Level-0}
+\caption{Strong Scaling Speedup for Rank-2 NMF on synthetic data}
 \label{fig:synrank2speedup}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_strongscaling.pdf}
-\caption{Relative Time Breakdown for Rank-2 NMF \cref{alg:parrank2nmf} Strong Scaling at Level-0}
+\caption{Relative Time Breakdown for Rank-2 NMF on synthetic data}
 \label{fig:synrank2strongscaling}
@@
 
+The relative time breakdowns are presented in \Cref{fig:synrank2strongscaling,fig:rwrank2strongscaling} and explain the strong scaling performance.
+Each experiment is normalized to 100\% time, so comparisons cannot be readily made across numbers of compute nodes. 
+For both data sets, we see that the time is dominated by MatMul, which is the primary reason for the scalability.
+The dominant matrix multiplications are between a large matrix and a matrix with 2 columns, so it is locally memory bandwidth bound, with performance proportional to the size of the large matrix.
+In each plot, we also see the relative time of all-gather and reduce-scatter increasing, which is because the local computation is decreasing while the communication cost is slightly increasing with $p$.
+This pattern will continue as $p$ increases, which will eventually limit scalability, but for these data sets the MatMul takes around 80\% of the time at over 2000 cores.
+
 \begin{figure}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_speedup.pdf}
-\caption{Strong Scaling Speedup for Rank-2 NMF \cref{alg:parrank2nmf} at Level-0 on Realworld Data}
+\caption{Strong Scaling Speedup for Rank-2 NMF on IMAGE data}
 \label{fig:rwrank2speedup}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_strongscaling.pdf}
-\caption{Relative Time Breakdown for Rank-2 NMF \cref{alg:parrank2nmf} Strong Scaling at Level-0 on Realworld Data}
+\caption{Relative Time Breakdown for Rank-2 NMF on IMAGE data}
 \label{fig:rwrank2strongscaling}
@@
 \begin{itemize}
-	\item theory says...t
+	\item theory (\Cref{eq:treecost}) says...
 	\item Fig 7 shows limit of strong scaling for synthetic data (time increases at 50 nodes compared to 40), Fig 8 gives breakdown and explains
@@
 \begin{itemize}
-	\item theory says 
+	\item theory says per-level cost is \Cref{eq:levelcost}
 \end{itemize}

### 2020-07-15 latex/fig/dc.tex
scaled the dc figure and fixed latex bug

@@
+% !TEX root = ../paper.tex
+
+\resizebox{\columnwidth}{!}{
 \begin{tikzpicture}
@@
             {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/1.png}};
+            \draw[thick,->] (-3.4,-1.2) .. controls (-4.4,-1.2) .. (-4.4,-2.4);
+            \node[inner sep=0pt] (3) at (-4.4,-4)
+                {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/4.png}};
+            \node[below=0.1] at (3.south) {\textbf{3}};
             \draw[thick,->] (-2.6,-1.2) .. controls (-1.6,-1.2) .. (-1.6,-2.4);
-            \node[inner sep=0pt] (3) at (-1.6,-4)
+            \node[inner sep=0pt] (4) at (-1.6,-4)
                 {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/3.png}};
                 \draw[thick,->] (-2,-3.2) .. controls (-2.5,-3.2) .. (-2.5,-4.4);
-                \node[inner sep=0pt] (7) at (-2.5,-6)
+                \node[inner sep=0pt] (9) at (-2.5,-6)
                     {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/7.png}};
+                \node[below=0.1] at (9.south) {\textbf{9}};
                 \draw[thick,->] (-1.2,-3.2) .. controls (-0.7,-3.2) .. (-0.7,-4.4);
-                \node[inner sep=0pt] (8) at (-0.7,-6)
+                \node[inner sep=0pt] (10) at (-0.7,-6)
                     {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/8.png}};
-            \draw[thick,->] (-3.4,-1.2) .. controls (-4.4,-1.2) .. (-4.4,-2.4);
-            \node[inner sep=0pt] (4) at (-4.4,-4)
-                {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/4.png}};
+                \node[below=0.1] at (10.south) {\textbf{10}};
+             \node[below=0.1] at (4.south) {\textbf{4}};
+        \node[below=0.1] at (1.south) {\textbf{1}};
         \draw[thick,->] (0.4,0.8) .. controls (3,0.8) .. (3,-0.4);
@@
             {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/2.png}};
+            \draw[thick,->] (2.6,-1.2) .. controls (1.6,-1.2) .. (1.6,-2.4);
+            \node[inner sep=0pt] (5) at (1.6,-4)
+                {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/6.png}};
+                \draw[thick,->] (1.2,-3.2) .. controls (0.7,-3.2) .. (0.7,-4.4);
+                \node[inner sep=0pt] (11) at (0.7,-6)
+                    {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/13.png}};
+                \node[below=0.1] at (11.south) {\textbf{11}};
+                \draw[thick,->] (2,-3.2) .. controls (2.5,-3.2) .. (2.5,-4.4);
+                \node[inner sep=0pt] (12) at (2.5,-6)
+                    {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/14.png}};
+                \node[below=0.1] at (12.south) {\textbf{12}};
+            \node[below=0.1] at (5.south) {\textbf{5}};
             \draw[thick,->] (3.4,-1.2) .. controls (4.4,-1.2) .. (4.4,-2.4);
-            \node[inner sep=0pt] (5) at (4.4,-4)
-                {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/3.png}};
-            \draw[thick,->] (2.6,-1.2) .. controls (1.6,-1.2) .. (1.6,-2.4);
-            \node[inner sep=0pt] (6) at (1.6,-4)
-                {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/4.png}};
-                \draw[thick,->] (1.2,-3.2) .. controls (0.7,-3.2) .. (0.7,-4.4);
-                \node[inner sep=0pt] (13) at (0.7,-6)
-                    {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/3.png}};
-                    \draw[thick,->] (2,-3.2) .. controls (2.5,-3.2) .. (2.5,-4.4);
-                \node[inner sep=0pt] (14) at (2.5,-6)
-                    {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/4.png}};
+            \node[inner sep=0pt] (6) at (4.4,-4)
+                {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/5.png}};
+            \node[below=0.1] at (6.south) {\textbf{6}};
+        \node[below=0.1] at (2.south) {\textbf{2}};
+    \node[below=0.1] at (original.south) {\textbf{0}};
 \end{tikzpicture}
+}

### 2020-07-16 latex/paper.tex
added placeholder perf subsection for DC speedup

@@
 
+\subsubsection{Single-Node Scaling for DC Dataset}
+
+\GB{add only speedup plot for entire tree (at least all complete levels) from 1 core to 42 cores}
+
+\begin{itemize}
+	\item emphasize we start with a small data set that fits on one node and has only 191 rows
+	\item using 1 node (2 sockets), because we are memory bandwidth bound we expect limited speedup but we can save time with parallelism even for this small a data set
+\end{itemize}
+
 \subsubsection{Rank-2 NMF Strong Scaling}

### 2020-07-16 latex/paper.tex
updated datasets

--- 2020-07-15 latex/paper.tex
@@
 \newcommand{\RK}[1]{\textcolor{blue}{\textbf{Ramki}: #1}}
+
+\newcommand{\hyper}{DC-HYDICE}
+\newcommand{\image}{SIIM-ISIC}
 
@@
 \begin{itemize}
-	\item \textit{Document Clustering}:
-	These datasets are represented by sparse matrices in term frequency-inverdse document frequency (TFIDF) format.
-	The rows of these matrices represent words and the columns represent individual documents. Each value in the matrix is a TFIDF statistic for
-	a specific word in a document.
-
-	For H2NMF, we used two popular document clustering datasets: Reuters Newswire Topic Classification (Reuters-21578) and \dots
-	\LM{I can't find the official name for PubMed or a link to it}
-
-	The Reuters matrix is $12411\times 7984$ and is a preprocessed version of Reuters-21578 used in the SmallK implementation of H2NMF.
-	The preprocessed version is the 20 largest prelabeled classes of documents with only single class labels. 
-	\LM{Prof Park, is this explanation of Reuters correct?}
-	As Reuters is small, it is mainly used as a comparison tool between the ours and the SmallK implmentation.
-
-	The PubMed matrix is $141043\times 8200000$. It is not preprocessed and represents the full size dataset in TFIDF format. This version is
-	better suited for scaling as is it is much larger than Reuters and is a good representation of large text classification datasets.
-
 	\item \textit{Hyperspectral Imaging}:
-	The images chosen for this dataset follow a similar pattern to Document Clustering. We chose two datasets: the Hyperspectral digital imagery
-	collection experiment (HYDICE) image of the Washington DC Mall and the Airborne Visible/Infrared Image Spectrometer (AVIRIS) image 
-	of the DC area.
-
-	The images are formatted into 3-way tensors representing individual images in each spectral band captured. For H2NMF, these tensors 
-	are flattened so that the rows represent spectral bands (191 for HYDICE and 224 for AVIRIS) and the columns represent pixels.
-
-	The HYDICE image is flattened from $1280\times 307\times 191$ to $191\times 392960$ and the AVIRIS image is likewise
-	flattened from $8940\times 747\times 224$ to $224\times 6678180$. Unlike with document clustering, hyperspectral image sizes grow in pixels
-	and not spectral bands. So, for both images, large scaling is impossible given our current data distribtion.
-	\LM{I think this is awkward and I'm not sure if we need to include it}
-
 	\LM{DC: https://engineering.purdue.edu/~biehl/MultiSpec/hyperspectral.html}
-
-	\LM{AVIRIS: https://aviris.jpl.nasa.gov/dataportal/}
-
-	\item synthetic (dense and sparse)
-	\item HSI: DC and Aviris
-	\item Doc-clustering: Reuters and PubMed
+	For Hyperspectral imaging, we chose the Hyperspectral digital imagery collection experiment (HYDICE) image of the Washington DC Mall. We will refer
+	to this dataset as \hyper{}.
+	\hyper{} is formatted into a 3-way tensor representing two spatial dimensions of pixels and one dimension of spectral bands. So, a slice along
+	the spectral band dimnesion would be the full \hyper{} image in that spectral band. For H2NMF, these tensors are flattened so that the rows represent the
+	$191$ spectral bands and the columns represent the $392960$ pixels. 
+	Larger hyperspectral datasets tend to not grow larger than around 200 spectral bands and have millions or billions of pixels. So, scaling is 
+	limited for this dataset given our data distribution scheme.
+	\item \textit{Image Classification}:
+	\LM{https://www.kaggle.com/c/siim-isic-melanoma-classification/data}
+	Image Classification datasets contain tens of thousands of images of equal size. The dataset we chose is the SIIM-ISIC Melanoma classification
+	dataset, which we will refer to as \image{}. \image{} consists of $33126$ RGB training images equally sized at $1024 \times 1024$. Unlike with hyperspectral imaging, the resulting
+	matrix used in H2NMF consists of image pixels along the rows and individual images along the columns. So, the resulting sized matrix is
+	$3145728 \times 33126$, which is approximately 800GB in size. 
+	Given its size, \image{} can only scale from $10$ Summit nodes.
+	\item \textit{Synthetic Dataset}:
+	Our synthetic dataset is the same shape of \image{} but consists of fewer rows and columns by a factor of $3$. The resulting matrix is $1048576 \times 11042$. 
+	We chose a smaller size dataset to better see the limits of scaling and to scale from fewer Summit nodes.
 \end{itemize}

### 2020-07-16 latex/paper.tex
dataset spacing
started drafting introduction

--- 2020-07-16 latex/paper.tex
@@
 Furthermore, the rank-two NMF algorithm can be parallelized efficiently using a computational kernel similar to matrix-vector multiplication.
-Our data distribution and parallelization strategies are designed to maintain computational load balance throughout the data-dependent hierarchy of computation while limiting interprocess communication, allowing the algorithm to scale to large dense and sparse data sets.
-We demonstrate the scalability of our parallel algorithm in terms of data size and number of processors, and apply the hierarchical clustering approach to large (dense) hyperspectral imaging and (sparse) text document data.
+Our data distribution and parallelization strategies are designed to maintain computational load balance throughout the data-dependent hierarchy of computation while limiting interprocess communication, allowing the algorithm to scale to large dense data sets.
+We demonstrate the scalability of our parallel algorithm in terms of data size and number of processors, and apply the hierarchical clustering approach to hyperspectral imaging and image classification data.
 \end{abstract}
@@
 
-\GB{Prof Park, can you draft an introduction?
-We are focusing on the dense case and using HSI and image clustering as applications.
-Section 4 is still unfinished, but nearly all of the performance plots are there.
-Feel free to leave placeholders for details that aren't obvious, we can clean them up Wednesday night.}
+Nonnegative Matrix Factorization (NMF) has been demonstrated to be an effective tool for unsupervised learning problems including clustering \cite{}. \GB{add citations}
+An NMF consists of two tall-and-skinny non-negative matrices whose product approximates a nonnegative data matrix.
+That is, given an $m\times n$ data matrix $\M{A}$, we seek nonnegative matrices $\M{W}$ and $\M{H}$ that each have $k$ columns so that $\M{A} \approx \M{W}\M{H}^\Tra$.
+Each pair of corresponding columns of $\M{W}$ and $\M{H}$ form a latent component of the NMF.
+If the rows of $\M{A}$ correspond to features and the columns to samples, the $i$th row of the $\M{H}$ matrix represents the loading of sample $i$ onto each latent component and provides a soft clustering.
+Because the $\M{W}$ factor is also nonnegative, each column can typically be interpreted as a latent feature vector for each cluster.
+
+While standard NMF provides a flat clustering, NMF can also be used as a tool for hierarchical clustering.
+Kuang and Park \cite{KP13} propose a method that uses rank-2 NMF to recursively bipartition the samples.
+An NMF with $k=2$ determines a row of $\M{H}$ of length 2 for each sample that determines its part membership.
+By recursively bipartitioning the samples in each part, the method determines a binary tree such that each sample belongs to a unique leaf and internal nodes of the tree determine higher-level clusters of the samples.
+A single latent feature vector for each node can also be used for interpreting clusters.
+
+We illustrate the output of the hierarchical clustering method with an example data set and output tree.
+Following Gillis et al.
+\cite{GKP15}, we apply the method to a hyperspectral imaging (HSI) data set of the Washington, D.C~national mall, which has pixel dimensions $1280 \times 307$ and 191 spectral bands.
+\Cref{fig:dc} visualizes the output tree with 6 leaves along with their hierarchical relationships.
+The root node, labeled 0, is a flattening of the HSI data to a 2D grayscale image.
+Each other node is represented by an overlay of the member pixels of the clusters (in blue) on the original grayscale image.
+The first bipartitioning separates vegetation (cluster 1) from non-vegetation (cluster 2), the bipartitioning of cluster 1 seems to separate grass (cluster 3) from trees (cluster 4), the bipartitioning of cluster 2 seems to separate buildings (cluster 5) from sidewalks/water (cluster 6), and so on.
+
+\begin{itemize}
+	\item discuss limitation of current implementations of the hierarchical method in MATLAB and shared-memory parallelization in terms of scalability to data size and bandwidth limitations
+	\item brief summary of parallelization approach, point to analysis
+	\item summarize results, highlight 600 GB data set and scaling to 2000-2500 cores, speedups of up to X for rank-2 NMF and up to Y for full hierarchical clustering algorithm
+\end{itemize}
+
+\GB{we should have perhaps compared against SmallK and also against flat NMF...}
+
+\begin{figure}
+\input{fig/dc}
+\caption{Hierarchical Clustering of DC Mall Hyperspectral Image}
+\label{fig:dc}
+\end{figure}
 
@@
 
-\subsection{DC Clustering Results}
-
-\subsubsection{Hyperspectral Imaging}
-\begin{figure}
-\input{fig/dc}
-\caption{Hierarchical Cluster from DC Mall Image}
-\label{fig:dc}
-\end{figure}
-
 \section{Conclusion}

### 2020-07-16 latex/paper.tex

@@
 \begin{itemize}
-	\item \textit{Document Clustering}:
-	These datasets are represented by sparse matrices in term frequency-inverdse document frequency (TFIDF) format.
-	The rows of these matrices represent words and the columns represent individual documents. Each value in the matrix is a TFIDF statistic for
-	a specific word in a document.
+	\item \textit{Hyperspectral Imaging}:
+	\LM{DC: https://engineering.purdue.edu/~biehl/MultiSpec/hyperspectral.html}
+	For Hyperspectral imaging, we chose the Hyperspectral digital imagery collection experiment (HYDICE) image of the Washington DC Mall. We will refer
+	to this dataset as \hyper{}.
 
-	For H2NMF, we used two popular document clustering datasets: Reuters Newswire Topic Classification (Reuters-21578) and \dots
-	\LM{I can't find the official name for PubMed or a link to it}
+	\hyper{} is formatted into a 3-way tensor representing two spatial dimensions of pixels and one dimension of spectral bands. So, a slice along
+	the spectral band dimnesion would be the full \hyper{} image in that spectral band. For H2NMF, these tensors are flattened so that the rows represent the
+	$191$ spectral bands and the columns represent the $392960$ pixels.
 
-	The Reuters matrix is $12411\times 7984$ and is a preprocessed version of Reuters-21578 used in the SmallK implementation of H2NMF.
-	The preprocessed version is the 20 largest prelabeled classes of documents with only single class labels. 
-	\LM{Prof Park, is this explanation of Reuters correct?}
-	As Reuters is small, it is mainly used as a comparison tool between the ours and the SmallK implmentation.
+	Larger hyperspectral datasets tend to not grow larger than around 200 spectral bands and have millions or billions of pixels. So, scaling is 
+	limited for this dataset given our data distribution scheme.
+	\item \textit{Image Classification}:
+	\LM{https://www.kaggle.com/c/siim-isic-melanoma-classification/data}
+	Image Classification datasets contain tens of thousands of images of equal size. The dataset we chose is the SIIM-ISIC Melanoma classification
+	dataset, which we will refer to as \image{}. \image{} consists of $33126$ RGB training images equally sized at $1024 \times 1024$. Unlike with hyperspectral imaging, the resulting
+	matrix used in H2NMF consists of image pixels along the rows and individual images along the columns. So, the resulting sized matrix is
+	$3145728 \times 33126$, which is approximately 800GB in size.
 
-	The PubMed matrix is $141043\times 8200000$. It is not preprocessed and represents the full size dataset in TFIDF format. This version is
-	better suited for scaling as is it is much larger than Reuters and is a good representation of large text classification datasets.
-
-	\item \textit{Hyperspectral Imaging}:
-	The images chosen for this dataset follow a similar pattern to Document Clustering. We chose two datasets: the Hyperspectral digital imagery
-	collection experiment (HYDICE) image of the Washington DC Mall and the Airborne Visible/Infrared Image Spectrometer (AVIRIS) image 
-	of the DC area.
-
-	The images are formatted into 3-way tensors representing individual images in each spectral band captured. For H2NMF, these tensors 
-	are flattened so that the rows represent spectral bands (191 for HYDICE and 224 for AVIRIS) and the columns represent pixels.
-
-	The HYDICE image is flattened from $1280\times 307\times 191$ to $191\times 392960$ and the AVIRIS image is likewise
-	flattened from $8940\times 747\times 224$ to $224\times 6678180$. Unlike with document clustering, hyperspectral image sizes grow in pixels
-	and not spectral bands. So, for both images, large scaling is impossible given our current data distribtion.
-	\LM{I think this is awkward and I'm not sure if we need to include it}
-
-	\LM{DC: https://engineering.purdue.edu/~biehl/MultiSpec/hyperspectral.html}
-
-	\LM{AVIRIS: https://aviris.jpl.nasa.gov/dataportal/}
-
-	\item synthetic (dense and sparse)
-	\item HSI: DC and Aviris
-	\item Doc-clustering: Reuters and PubMed
+	Given its size, \image{} can only scale from $10$ Summit nodes.
+	\item \textit{Synthetic Dataset}:
+	Our synthetic dataset is the same shape of \image{} but consists of fewer rows and columns by a factor of $3$. The resulting matrix is $1048576 \times 11042$. 
+	We chose a smaller size dataset to better see the limits of scaling and to scale from fewer Summit nodes.
 \end{itemize}

### 2020-07-16 latex/paper.tex
replaced IMAGE with macro

--- 2020-07-16 latex/paper.tex
@@
 
-\GB{need shortcut names for datasets: synthetic, DC, and something better than IMAGE}
+\GB{need shortcut names for datasets: synthetic, DC, and something better than \image{}}
 
@@
 	to this dataset as \hyper{}.
+
 	\hyper{} is formatted into a 3-way tensor representing two spatial dimensions of pixels and one dimension of spectral bands. So, a slice along
@@
 	$191$ spectral bands and the columns represent the $392960$ pixels.
+
 	Larger hyperspectral datasets tend to not grow larger than around 200 spectral bands and have millions or billions of pixels. So, scaling is 
@@
 	$3145728 \times 33126$, which is approximately 800GB in size.
+
 	Given its size, \image{} can only scale from $10$ Summit nodes.
@@
 
-We perform strong scaling experiments for a single Rank-2 NMF (\Cref{alg:parrank2nmf}) on the synthetic and IMAGE datasets.
+We perform strong scaling experiments for a single Rank-2 NMF (\Cref{alg:parrank2nmf}) on the synthetic and \image{} datasets.
 The theory (\Cref{eq:r2nmfcost}) suggests that perfect strong scaling is possible as long as the execution time is dominated by local computation.
@@
 \Cref{fig:synrank2speedup,fig:rwrank2speedup} show performance relative to the smallest number of nodes required to store data and factor matrices.
-The synthetic data is chosen to fit on a single node, while the IMAGE data requires 10 nodes to store the input matrix as well as temporary and output matrices.
-For these data sets, we observe nearly perfect strong scaling, with 47$\times$ speedup on 50 nodes (over 1 node) for synthetic data and 7.1$\times$ speedup on 80 nodes (over 10 nodes) for IMAGE data.
+The synthetic data is chosen to fit on a single node, while the \image{} data requires 10 nodes to store the input matrix as well as temporary and output matrices.
+For these data sets, we observe nearly perfect strong scaling, with 47$\times$ speedup on 50 nodes (over 1 node) for synthetic data and 7.1$\times$ speedup on 80 nodes (over 10 nodes) for \image{} data.
 
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_speedup.pdf}
-\caption{Strong Scaling Speedup for Rank-2 NMF on IMAGE data}
+\caption{Strong Scaling Speedup for Rank-2 NMF on \image{} data}
 \label{fig:rwrank2speedup}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_strongscaling.pdf}
-\caption{Relative Time Breakdown for Rank-2 NMF on IMAGE data}
+\caption{Relative Time Breakdown for Rank-2 NMF on \image{} data}
 \label{fig:rwrank2strongscaling}

### 2020-07-16 latex/paper.tex
replaced real world with \image

@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_hierarchical_speedup.pdf}
-\caption{Strong Scaling Speedup for Hierarchical Clustering on Realworld Data. The total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
+\caption{Strong Scaling Speedup for Hierarchical Clustering on \image{} Data. The total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
 \label{fig:rwhierspeedup}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_hier_strongscaling.pdf}
-\caption{Relative Time Breakdown for Hierarchical Clustering on Realworld. The plot uses the total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
+\caption{Relative Time Breakdown for Hierarchical Clustering on \image{}. The plot uses the total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
 \label{fig:rwhierstrongscaling}

### 2020-07-16 latex/paper.tex
cited datasets

@@
 	\item \textit{Hyperspectral Imaging}:
-	\LM{DC: https://engineering.purdue.edu/~biehl/MultiSpec/hyperspectral.html}
 	For Hyperspectral imaging, we chose the Hyperspectral digital imagery collection experiment (HYDICE) image of the Washington DC Mall. We will refer
-	to this dataset as \hyper{}.
+	to this dataset as \hyper{}\cite{DC-HYDICE}.
 
@@
 	\item \textit{Image Classification}:
-	\LM{https://www.kaggle.com/c/siim-isic-melanoma-classification/data}
 	Image Classification datasets contain tens of thousands of images of equal size. The dataset we chose is the SIIM-ISIC Melanoma classification
-	dataset, which we will refer to as \image{}. \image{} consists of $33126$ RGB training images equally sized at $1024 \times 1024$. Unlike with hyperspectral imaging, the resulting
+	dataset, which we will refer to as \image{}\cite{SIIM-ISIC}. \image{} consists of $33126$ RGB training images equally sized at $1024 \times 1024$. Unlike with hyperspectral imaging, the resulting
 	matrix used in H2NMF consists of image pixels along the rows and individual images along the columns. So, the resulting sized matrix is

### 2020-07-16 latex/paper.tex
still working in intro

--- 2020-07-16 latex/paper.tex
@@
 
-\GB{Prof Park, can you draft an introduction?
-We are focusing on the dense case and using HSI and image clustering as applications.
-Section 4 is still unfinished, but nearly all of the performance plots are there.
-Feel free to leave placeholders for details that aren't obvious, we can clean them up Wednesday night.}
+Nonnegative Matrix Factorization (NMF) has been demonstrated to be an effective tool for unsupervised learning problems including clustering \cite{}. \GB{add citations}
+An NMF consists of two tall-and-skinny non-negative matrices whose product approximates a nonnegative data matrix.
+That is, given an $m\times n$ data matrix $\M{A}$, we seek nonnegative matrices $\M{W}$ and $\M{H}$ that each have $k$ columns so that $\M{A} \approx \M{W}\M{H}^\Tra$.
+Each pair of corresponding columns of $\M{W}$ and $\M{H}$ form a latent component of the NMF.
+If the rows of $\M{A}$ correspond to features and the columns to samples, the $i$th row of the $\M{H}$ matrix represents the loading of sample $i$ onto each latent component and provides a soft clustering.
+Because the $\M{W}$ factor is also nonnegative, each column can typically be interpreted as a latent feature vector for each cluster.
+
+While standard NMF provides a flat clustering, NMF can also be used as a tool for hierarchical clustering.
+Kuang and Park \cite{KP13} propose a method that uses rank-2 NMF to recursively bipartition the samples.
+An NMF with $k=2$ determines a row of $\M{H}$ of length 2 for each sample that determines its part membership.
+By recursively bipartitioning the samples in each part, the method determines a binary tree such that each sample belongs to a unique leaf and internal nodes of the tree determine higher-level clusters of the samples.
+A single latent feature vector for each node can also be used for interpreting clusters.
+Du et al.
+\cite{DKDP17} show that the latent features vectors of the leaves can also be used to initialize a $\M{W}$ within an algorithm for flat NMF, yielding fast convergence and overall speedup compared to initializing the algorithm with random data.
+
+We illustrate the output of the hierarchical clustering method with an example data set and output tree.
+Following Gillis et al.
+\cite{GKP15}, we apply the method to a hyperspectral imaging (HSI) data set of the Washington, D.C~national mall, which has pixel dimensions $1280 \times 307$ and 191 spectral bands.
+\Cref{fig:dc} visualizes the output tree with 6 leaves along with their hierarchical relationships.
+The root node, labeled 0, is a flattening of the HSI data to a 2D grayscale image.
+Each other node is represented by an overlay of the member pixels of the clusters (in blue) on the original grayscale image.
+The first bipartitioning separates vegetation (cluster 1) from non-vegetation (cluster 2), the bipartitioning of cluster 1 separates grass (cluster 3) from trees (cluster 4), the bipartitioning of cluster 2 separates buildings (cluster 5) from sidewalks/water (cluster 6), and so on.
+
+While the hierarchical clustering method offers advantages in terms of interpretation as well as execution time compared to flat NMF, implementations of the algorithm are limited to single workstations and the dataset must fit in the available memory.
+Currently available implementations can utilize multiple cores via MATLAB \cite{KP13} or explicit shared-memory parallelization \cite{SmallK}.
+The goal of this work is to use distributed-memory parallelism to scale the algorithm to large datasets that require the memory of multiple compute nodes and to high processor counts.
+
+\begin{itemize}
+	\item discuss limitation of current implementations of the hierarchical method in MATLAB and shared-memory parallelization in terms of scalability to data size and bandwidth limitations
+	\item brief summary of parallelization approach, point to analysis
+	\item summarize results, highlight 600 GB data set and scaling to 2000-2500 cores, speedups of up to X for rank-2 NMF and up to Y for full hierarchical clustering algorithm
+\end{itemize}
+
+\GB{we should have perhaps compared against SmallK and also against flat NMF...}
+
+\begin{figure}
+\input{fig/dc}
+\caption{Hierarchical Clustering of DC Mall Hyperspectral Image}
+\label{fig:dc}
+\end{figure}
 
@@
 
-\GB{need shortcut names for datasets: synthetic, DC, and something better than \image{}}
+\GB{need shortcut names for datasets: synthetic, DC, and something better than IMAGE}
 
@@
 
-We perform strong scaling experiments for a single Rank-2 NMF (\Cref{alg:parrank2nmf}) on the synthetic and \image{} datasets.
+We perform strong scaling experiments for a single Rank-2 NMF (\Cref{alg:parrank2nmf}) on the synthetic and IMAGE datasets.
 The theory (\Cref{eq:r2nmfcost}) suggests that perfect strong scaling is possible as long as the execution time is dominated by local computation.
@@
 \Cref{fig:synrank2speedup,fig:rwrank2speedup} show performance relative to the smallest number of nodes required to store data and factor matrices.
-The synthetic data is chosen to fit on a single node, while the \image{} data requires 10 nodes to store the input matrix as well as temporary and output matrices.
-For these data sets, we observe nearly perfect strong scaling, with 47$\times$ speedup on 50 nodes (over 1 node) for synthetic data and 7.1$\times$ speedup on 80 nodes (over 10 nodes) for \image{} data.
+The synthetic data is chosen to fit on a single node, while the IMAGE data requires 10 nodes to store the input matrix as well as temporary and output matrices.
+For these data sets, we observe nearly perfect strong scaling, with 47$\times$ speedup on 50 nodes (over 1 node) for synthetic data and 7.1$\times$ speedup on 80 nodes (over 10 nodes) for IMAGE data.
 
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_speedup.pdf}
-\caption{Strong Scaling Speedup for Rank-2 NMF on \image{} data}
+\caption{Strong Scaling Speedup for Rank-2 NMF on IMAGE data}
 \label{fig:rwrank2speedup}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_strongscaling.pdf}
-\caption{Relative Time Breakdown for Rank-2 NMF on \image{} data}
+\caption{Relative Time Breakdown for Rank-2 NMF on IMAGE data}
 \label{fig:rwrank2strongscaling}

### 2020-07-16 latex/paper.tex

@@
 Furthermore, the rank-two NMF algorithm can be parallelized efficiently using a computational kernel similar to matrix-vector multiplication.
-Our data distribution and parallelization strategies are designed to maintain computational load balance throughout the data-dependent hierarchy of computation while limiting interprocess communication, allowing the algorithm to scale to large dense data sets.
-We demonstrate the scalability of our parallel algorithm in terms of data size and number of processors, and apply the hierarchical clustering approach to hyperspectral imaging and image classification data.
+Our data distribution and parallelization strategies are designed to maintain computational load balance throughout the data-dependent hierarchy of computation while limiting interprocess communication, allowing the algorithm to scale to large dense and sparse data sets.
+We demonstrate the scalability of our parallel algorithm in terms of data size and number of processors, and apply the hierarchical clustering approach to large (dense) hyperspectral imaging and (sparse) text document data.
 \end{abstract}
@@
 
+\subsection{DC Clustering Results}
+
+\subsubsection{Hyperspectral Imaging}
+\begin{figure}
+\input{fig/dc}
+\caption{Hierarchical Cluster from DC Mall Image}
+\label{fig:dc}
+\end{figure}
+
 \section{Conclusion}

### 2020-07-16 latex/paper.tex
finished first draft of intro

@@
 \cite{DKDP17} show that the latent features vectors of the leaves can also be used to initialize a $\M{W}$ within an algorithm for flat NMF, yielding fast convergence and overall speedup compared to initializing the algorithm with random data.
+We discuss the hierarchical method in more detail in \Cref{sec:prelim} and \Cref{sec:seqalgs}.
 
@@
 The first bipartitioning separates vegetation (cluster 1) from non-vegetation (cluster 2), the bipartitioning of cluster 1 separates grass (cluster 3) from trees (cluster 4), the bipartitioning of cluster 2 separates buildings (cluster 5) from sidewalks/water (cluster 6), and so on.
+If the algorithm continues, it chooses to split the leaf node that provides the greatest benefit to the overall tree, which can be quantified as a node's ``score'' in various ways.
+The algorithm proceeds to split nodes until a target number of leaves is obtained, until no node is worth splitting, or some other stopping criterion is met.
 
 While the hierarchical clustering method offers advantages in terms of interpretation as well as execution time compared to flat NMF, implementations of the algorithm are limited to single workstations and the dataset must fit in the available memory.
-Currently available implementations can utilize multiple cores via MATLAB \cite{KP13} or explicit shared-memory parallelization \cite{SmallK}.
+Currently available implementations can utilize multiple cores via MATLAB \cite{KP13} or explicit shared-memory parallelization in the SmallK library \cite{SmallK}.
 The goal of this work is to use distributed-memory parallelism to scale the algorithm to large datasets that require the memory of multiple compute nodes and to high processor counts.
+While flat NMF algorithms have been scaled to HPC platforms \cite{} \GB{add citations beyond our own}, our implementation is the first to our knowledge to scale a hierarchical NMF method to 1000s of cores.
+As discussed in detail in \Cref{sec:paralgs}, we choose to parallelize the computations associated with each node in the tree, which involve a Rank-2 NMF and the computation of the node's score.
+We choose a data matrix distribution across processors that avoids any redistribution of the input matrix regardless of the data-dependent structure of the tree's splitting decisions so that the communication required involves only the small factor matrices.
+Analysis of the algorithm shows the dependence of execution time on computation and communication costs as well as the number of clusters computed.
 
-\begin{itemize}
-	\item discuss limitation of current implementations of the hierarchical method in MATLAB and shared-memory parallelization in terms of scalability to data size and bandwidth limitations
-	\item brief summary of parallelization approach, point to analysis
-	\item summarize results, highlight 600 GB data set and scaling to 2000-2500 cores, speedups of up to X for rank-2 NMF and up to Y for full hierarchical clustering algorithm
-\end{itemize}
+We demonstrate in \Cref{sec:perf} the efficiency and scalability of our parallel algorithm on three data sets, including the HSI data of the DC mall and an image classification data set involving skin melanoma.
+The experimental results show that our parallelization of Rank-2 NMF is highly scalable, maintaining computation bound performance on 1000s of cores.
+We also show the limits of strong scalability when scaling to large numbers of clusters (leaf nodes), as the execution time shifts to becoming interprocessor bandwidth bound and eventually latency bound.
+The image classification data set requires 800 GB of memory across multiple nodes to process, and in scaling from 10 nodes to 80 nodes of the Summit supercomputer, we demonstrate parallel speedups of 7.1$\times$ for a single Rank-2 NMF and ZZ$\times$ for a complete hierarchical clustering.
 
@@
 Scaling algorithms for NMF to large data often requires parallelization in order to fit the data across the memories of multiple compute nodes or speed up the computation to complete in reasonable time.
-Parallelizations of multiple optimization approaches have been proposed for general NMF.
+Parallelizations of multiple optimization approaches have been proposed for general NMF. \GB{add more citations here, including to SmallK}
 In particular, we build upon the work of Kannan et al.
@@
 \subsection{Sequential Algorithms}
+\label{sec:seqalgs}
 

### 2020-07-16 latex/paper.tex
sections 2 and 3 comments

--- 2020-07-16 latex/paper.tex
@@
 
-Nonnegative Matrix Factorization (NMF) has been demonstrated to be an effective tool for unsupervised learning problems including clustering \cite{}. \GB{add citations}
-An NMF consists of two tall-and-skinny non-negative matrices whose product approximates a nonnegative data matrix.
-That is, given an $m\times n$ data matrix $\M{A}$, we seek nonnegative matrices $\M{W}$ and $\M{H}$ that each have $k$ columns so that $\M{A} \approx \M{W}\M{H}^\Tra$.
-Each pair of corresponding columns of $\M{W}$ and $\M{H}$ form a latent component of the NMF.
-If the rows of $\M{A}$ correspond to features and the columns to samples, the $i$th row of the $\M{H}$ matrix represents the loading of sample $i$ onto each latent component and provides a soft clustering.
-Because the $\M{W}$ factor is also nonnegative, each column can typically be interpreted as a latent feature vector for each cluster.
-
-While standard NMF provides a flat clustering, NMF can also be used as a tool for hierarchical clustering.
-Kuang and Park \cite{KP13} propose a method that uses rank-2 NMF to recursively bipartition the samples.
-An NMF with $k=2$ determines a row of $\M{H}$ of length 2 for each sample that determines its part membership.
-By recursively bipartitioning the samples in each part, the method determines a binary tree such that each sample belongs to a unique leaf and internal nodes of the tree determine higher-level clusters of the samples.
-A single latent feature vector for each node can also be used for interpreting clusters.
-Du et al.
-\cite{DKDP17} show that the latent features vectors of the leaves can also be used to initialize a $\M{W}$ within an algorithm for flat NMF, yielding fast convergence and overall speedup compared to initializing the algorithm with random data.
-
-We illustrate the output of the hierarchical clustering method with an example data set and output tree.
-Following Gillis et al.
-\cite{GKP15}, we apply the method to a hyperspectral imaging (HSI) data set of the Washington, D.C~national mall, which has pixel dimensions $1280 \times 307$ and 191 spectral bands.
-\Cref{fig:dc} visualizes the output tree with 6 leaves along with their hierarchical relationships.
-The root node, labeled 0, is a flattening of the HSI data to a 2D grayscale image.
-Each other node is represented by an overlay of the member pixels of the clusters (in blue) on the original grayscale image.
-The first bipartitioning separates vegetation (cluster 1) from non-vegetation (cluster 2), the bipartitioning of cluster 1 separates grass (cluster 3) from trees (cluster 4), the bipartitioning of cluster 2 separates buildings (cluster 5) from sidewalks/water (cluster 6), and so on.
-
-While the hierarchical clustering method offers advantages in terms of interpretation as well as execution time compared to flat NMF, implementations of the algorithm are limited to single workstations and the dataset must fit in the available memory.
-Currently available implementations can utilize multiple cores via MATLAB \cite{KP13} or explicit shared-memory parallelization \cite{SmallK}.
-The goal of this work is to use distributed-memory parallelism to scale the algorithm to large datasets that require the memory of multiple compute nodes and to high processor counts.
-
-\begin{itemize}
-	\item discuss limitation of current implementations of the hierarchical method in MATLAB and shared-memory parallelization in terms of scalability to data size and bandwidth limitations
-	\item brief summary of parallelization approach, point to analysis
-	\item summarize results, highlight 600 GB data set and scaling to 2000-2500 cores, speedups of up to X for rank-2 NMF and up to Y for full hierarchical clustering algorithm
-\end{itemize}
-
-\GB{we should have perhaps compared against SmallK and also against flat NMF...}
-
-\begin{figure}
-\input{fig/dc}
-\caption{Hierarchical Clustering of DC Mall Hyperspectral Image}
-\label{fig:dc}
-\end{figure}
+\GB{Prof Park, can you draft an introduction?
+We are focusing on the dense case and using HSI and image clustering as applications.
+Section 4 is still unfinished, but nearly all of the performance plots are there.
+Feel free to leave placeholders for details that aren't obvious, we can clean them up Wednesday night.}
 
@@
 When the rank of the factorization (the number of columns of $\M{W}$ and $\M{H}$) is 2, the NNLS subproblems can be solved much more quickly because the number of possible active sets is only 4.
-As explained in more detail in \Cref{sec:r2nmf}, the optimal solution across the 4 possibilities can be determined efficiently to solve the NNLS subproblem more quickly than general-rank approaches like block principal pivoting.
+As explained in more detail in \Cref{sec:r2nmf}, the optimal solution across the 4 possibilities \LM{sets} can be determined efficiently to solve the NNLS subproblem more quickly than general-rank approaches like block principal pivoting.
 Because of the relative ease of solving the NMF problem for the rank-2 case, Kuang and Park \cite{KP13} propose a recursive method to use a rank-2 NMF to partition the input data into 2 parts, whereby each part can be further partitioned via rank-2 NMF of the corresponding original data.
@@
 The cost of an all-reduce of size $n$ words is $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n)$ for $n>p$ and $(\alpha+\beta) \cdot O(\log p)$ for $n<p$.
+\LM{is the n<p case right?
+where's the computation?}
 In an all-gather, all processors start out with separate data and all end with a copy of the same result, which is the union of all the input data.
@@
 	\item if the solution to the unconstrained least squares problem admits two positive values, it is the optimal solution to the nonnegatively constrained problem,
-	\item if $\M{W}$ and $\M{A}$ are both nonnegative, then the candidate solution with two zero values is never (uniquely) optimal and can be discarded, and
+	\item if $\M{W}$ and $\M{A}$ are both nonnegative, then the candidate solution with two zero values is never (uniquely) \LM{why uniquely?} optimal and can be discarded, and
 	\item if the solution to the unconstrained problem does not admit two positive values, the better of the two remaining solutions can be determined by comparing $\V{a}_{j}^\Tra\V{w}_1 / \|\V{w}_1\|$ and $\V{a}_{j}^\Tra\V{w}_2 / \|\V{w}_2\|$.
@@
 Given that the computational complexity of \Cref{alg:r2nnls} is $O(n)$ (or $O(m)$ when computing $\M{W}$), and the complexity of computing $\M{W}^\Tra\M{W}$ and $\M{H}^\Tra\M{H}$ is $O(m+n)$, the typical dominant cost of each iteration of Rank-2 NMF is that of computing $\M{A}^\Tra\M{W}$ and $\M{A}\M{H}$, which is $O(mn)$ when $\M{A}$ is dense and $O(\text{nnz}(A))$ when $\M{A}$ is sparse.
+\LM{should our analysis keep sparse even when we don't have a sparse dataset?}
 
@@
 
+\LM{should we have a little bit more of the description of the scoring method? also the language here is awkward}
 The splitting procedure is specified in \Cref{alg:split}.
@@
 Again, we use a fixed number of iterations for benchmarking, but the more practical stopping criterion is based on the relative change in $\sigma$.
+\LM{we did not use a fixed number of iterations for benchmarking}
 
@@
 Note that the only cost that depends on the level index $\ell$ is the latency cost.
+\LM{both the computation and the latency dpend on this}
 Summing over levels and assuming the tree is nearly balanced and has height $O(\log k)$ where $k$ is the number of frontier nodes, we obtain an overall cost of \Cref{alg:hiernmf2} of

### 2020-07-16 latex/paper.tex
edits to intro

--- 2020-07-16 latex/paper.tex
@@
 We choose a data matrix distribution across processors that avoids any redistribution of the input matrix regardless of the data-dependent structure of the tree's splitting decisions so that the communication required involves only the small factor matrices.
-Analysis of the algorithm shows the dependence of execution time on computation and communication costs as well as the number of clusters computed.
+Analysis of the algorithm shows the dependence of execution time on computation and communication costs as well as on $k$, the number of clusters computed.
+In particular, we confirm that many of the dominant costs are logarithmic in $k$, which is favorable to the linear or sometimes superlinear dependence of flat NMF algorithms.
 
-We demonstrate in \Cref{sec:perf} the efficiency and scalability of our parallel algorithm on three data sets, including the HSI data of the DC mall and an image classification data set involving skin melanoma.
+We demonstrate in \Cref{sec:results} the efficiency and scalability of our parallel algorithm on three data sets, including the HSI data of the DC mall and an image classification data set involving skin melanoma.
 The experimental results show that our parallelization of Rank-2 NMF is highly scalable, maintaining computation bound performance on 1000s of cores.
 We also show the limits of strong scalability when scaling to large numbers of clusters (leaf nodes), as the execution time shifts to becoming interprocessor bandwidth bound and eventually latency bound.
-The image classification data set requires 800 GB of memory across multiple nodes to process, and in scaling from 10 nodes to 80 nodes of the Summit supercomputer, we demonstrate parallel speedups of 7.1$\times$ for a single Rank-2 NMF and ZZ$\times$ for a complete hierarchical clustering.
+The image classification data set requires 800 GB of memory across multiple nodes to process, and in scaling from 10 nodes to 80 nodes of the Summit supercomputer (see \Cref{sec:summit}), we demonstrate parallel speedups of 7.1$\times$ for a single Rank-2 NMF and ZZ$\times$ for a complete hierarchical clustering.
 
@@
 \subsection{Parallelization}
+\label{sec:paralgs}
 
@@
 \section{Experimental Results}
+\label{sec:results}
 
 \subsection{Experimental Platform}
+\label{sec:summit}
+
 All the experiments in this section were conducted on Summit. Summit is a supercomputer created by IBM for the Oak Ridge National Laboratory. 

### 2020-07-16 latex/paper.tex
introduction clustering citations

@@
 
-Nonnegative Matrix Factorization (NMF) has been demonstrated to be an effective tool for unsupervised learning problems including clustering \cite{}. \GB{add citations}
+Nonnegative Matrix Factorization (NMF) has been demonstrated to be an effective tool for unsupervised learning problems including clustering \cite{XLG03, SBPP06, DHS05}. \GB{add citations}
 An NMF consists of two tall-and-skinny non-negative matrices whose product approximates a nonnegative data matrix.

### 2020-07-16 latex/paper.tex
remove gb add citations
fixed comments

--- 2020-07-16 latex/paper.tex
@@
 In an all-reduce, all processors start out with the same amount of data and all end with a copy of the same result, which is in our case a sum of all the inputs (and the same size as a single input).
-The cost of an all-reduce of size $n$ words is $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n)$ for $n>p$ and $(\alpha+\beta) \cdot O(\log p)$ for $n<p$.
-\LM{is the n<p case right?
-where's the computation?}
+The cost of an all-reduce of size $n$ words is $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n)$ for $n>p$ and $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n\log p)$ for $n<p$.
 In an all-gather, all processors start out with separate data and all end with a copy of the same result, which is the union of all the input data.
@@
 An all-reduce collective is used to obtain a copy of $\V{v}$ on every processor for the next iteration, and the norm is redundantly computed without further communication.
-Again, we use a fixed number of iterations for benchmarking, but the more practical stopping criterion is based on the relative change in $\sigma$.
-\LM{we did not use a fixed number of iterations for benchmarking}
+We used the relative change in $\sigma$ as the stopping criterion for benchmarking.
 

### 2020-07-16 latex/paper.tex

@@
 
-Nonnegative Matrix Factorization (NMF) has been demonstrated to be an effective tool for unsupervised learning problems including clustering \cite{}. \GB{add citations}
+Nonnegative Matrix Factorization (NMF) has been demonstrated to be an effective tool for unsupervised learning problems including clustering \cite{XLG03, SBPP06, DHS05}. 
 An NMF consists of two tall-and-skinny non-negative matrices whose product approximates a nonnegative data matrix.

### 2020-07-16 latex/paper.tex
edits to sections 2 and 3

--- 2020-07-16 latex/paper.tex
@@
 
-Nonnegative Matrix Factorization (NMF) has been demonstrated to be an effective tool for unsupervised learning problems including clustering \cite{XLG03, SBPP06, DHS05}. \GB{add citations}
+Nonnegative Matrix Factorization (NMF) has been demonstrated to be an effective tool for unsupervised learning problems including clustering \cite{}. \GB{add citations}
 An NMF consists of two tall-and-skinny non-negative matrices whose product approximates a nonnegative data matrix.
@@
 
-\LM{should we have a little bit more of the description of the scoring method? also the language here is awkward}
 The splitting procedure is specified in \Cref{alg:split}.
 After the Rank-2 NMF is performed, the $\M{H}$ factor is used to determine part membership, and the columns of the $\M{W}$ factor are assigned to the child nodes.
-The score of the node is computed as the reduction in overall NMF error if the node is split, which can be computed from the principal singular values of the subsets of columns of the node and its children.
-The principal singular values of the children are computed via the power method (the principal singular value of the node was computed when its parent's score was computed).
+The score of the node is computed as the reduction in overall NMF error if the node is split, which can be computed from the principal singular values of the subsets of columns of the node and its children, as given in \Cref{line:score}.
+The principal singular values of the children are computed via the power method.
+Note that the principal singular value of the node itself need not be recomputed as it was needed for its parent's score.
 
@@
 		\State ${\cal N}.\text{right} = \text{node}(\M{A}_2,\V{W}_2)$ \hfill \Comment{create right child}
-		\State ${\cal N}.\text{score} = \sigma_1^2(\M{A}_1) + \sigma_1^2(\M{A}_2) - \sigma_1^2({\cal N}.\M{A})$
+		\State ${\cal N}.\text{score} = \sigma_1^2(\M{A}_1) + \sigma_1^2(\M{A}_2) - \sigma_1^2({\cal N}.\M{A})$ \label{line:score}
 	\EndFunction
@@
 \end{equation}
-Note that the only cost that depends on the level index $\ell$ is the latency cost.
-\LM{both the computation and the latency dpend on this}
+Note that the only costs that depend on the level index $\ell$ are the latency cost and a lower-order computational cost.
+
 Summing over levels and assuming the tree is nearly balanced and has height $O(\log k)$ where $k$ is the number of frontier nodes, we obtain an overall cost of \Cref{alg:hiernmf2} of

### 2020-07-16 latex/paper.tex
completed abstract

--- 2020-07-16 latex/paper.tex
@@
 We demonstrate the scalability of our parallel algorithm in terms of data size and number of processors, and apply the hierarchical clustering approach to large (dense) hyperspectral imaging and (sparse) text document data.
+\GB{On 40 nodes of Summit supercomputer, our algorithm achieves (a) linear speedup for solving the Rank-2 Non-negative Least squares kernels and (b) achieves 14x speedup on synthetic matrices.
+We also demonstrate that in the case of real world matrices, the algorithm scales to 80 summit nodes.}
 \end{abstract}
@@
 In an all-reduce, all processors start out with the same amount of data and all end with a copy of the same result, which is in our case a sum of all the inputs (and the same size as a single input).
-The cost of an all-reduce of size $n$ words is $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n)$ for $n>p$ and $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n\log p)$ for $n<p$.
+The cost of an all-reduce of size $n$ words is $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n)$ for $n>p$ and $(\alpha+\beta) \cdot O(\log p)$ for $n<p$.
+\LM{is the n<p case right?
+where's the computation?}
 In an all-gather, all processors start out with separate data and all end with a copy of the same result, which is the union of all the input data.
@@
 An all-reduce collective is used to obtain a copy of $\V{v}$ on every processor for the next iteration, and the norm is redundantly computed without further communication.
-We used the relative change in $\sigma$ as the stopping criterion for benchmarking.
+Again, we use a fixed number of iterations for benchmarking, but the more practical stopping criterion is based on the relative change in $\sigma$.
+\LM{we did not use a fixed number of iterations for benchmarking}
 

### 2020-07-16 latex/paper.tex
resolved comments

--- 2020-07-16 latex/paper.tex
@@
 When the rank of the factorization (the number of columns of $\M{W}$ and $\M{H}$) is 2, the NNLS subproblems can be solved much more quickly because the number of possible active sets is only 4.
-As explained in more detail in \Cref{sec:r2nmf}, the optimal solution across the 4 possibilities \LM{sets} can be determined efficiently to solve the NNLS subproblem more quickly than general-rank approaches like block principal pivoting.
+As explained in more detail in \Cref{sec:r2nmf}, the optimal solution across the 4 sets can be determined efficiently to solve the NNLS subproblem more quickly than general-rank approaches like block principal pivoting.
 Because of the relative ease of solving the NMF problem for the rank-2 case, Kuang and Park \cite{KP13} propose a recursive method to use a rank-2 NMF to partition the input data into 2 parts, whereby each part can be further partitioned via rank-2 NMF of the corresponding original data.
@@
 	\item if the solution to the unconstrained least squares problem admits two positive values, it is the optimal solution to the nonnegatively constrained problem,
-	\item if $\M{W}$ and $\M{A}$ are both nonnegative, then the candidate solution with two zero values is never (uniquely) \LM{why uniquely?} optimal and can be discarded, and
+	\item if $\M{W}$ and $\M{A}$ are both nonnegative, then the candidate solution with two zero values is never (uniquely) optimal and can be discarded, and
 	\item if the solution to the unconstrained problem does not admit two positive values, the better of the two remaining solutions can be determined by comparing $\V{a}_{j}^\Tra\V{w}_1 / \|\V{w}_1\|$ and $\V{a}_{j}^\Tra\V{w}_2 / \|\V{w}_2\|$.
@@
 
-Given that the computational complexity of \Cref{alg:r2nnls} is $O(n)$ (or $O(m)$ when computing $\M{W}$), and the complexity of computing $\M{W}^\Tra\M{W}$ and $\M{H}^\Tra\M{H}$ is $O(m+n)$, the typical dominant cost of each iteration of Rank-2 NMF is that of computing $\M{A}^\Tra\M{W}$ and $\M{A}\M{H}$, which is $O(mn)$ when $\M{A}$ is dense and $O(\text{nnz}(A))$ when $\M{A}$ is sparse.
-\LM{should our analysis keep sparse even when we don't have a sparse dataset?}
+Given that the computational complexity of \Cref{alg:r2nnls} is $O(n)$ (or $O(m)$ when computing $\M{W}$), and the complexity of computing $\M{W}^\Tra\M{W}$ and $\M{H}^\Tra\M{H}$ is $O(m+n)$, the typical dominant cost of each iteration of Rank-2 NMF is that of computing $\M{A}^\Tra\M{W}$ and $\M{A}\M{H}$, which is $O(mn)$.
 

### 2020-07-16 latex/paper.tex
updated abstract

--- 2020-07-16 latex/paper.tex
@@
 Not only does this approach uncover more structure in the data than a flat NMF clustering, but also rank-two NMF can be computed more quickly than for general ranks, providing comparable overall time to solution.
-Furthermore, the rank-two NMF algorithm can be parallelized efficiently using a computational kernel similar to matrix-vector multiplication.
 Our data distribution and parallelization strategies are designed to maintain computational load balance throughout the data-dependent hierarchy of computation while limiting interprocess communication, allowing the algorithm to scale to large dense and sparse data sets.
-We demonstrate the scalability of our parallel algorithm in terms of data size and number of processors, and apply the hierarchical clustering approach to large (dense) hyperspectral imaging and (sparse) text document data.
+We demonstrate the scalability of our parallel algorithm in terms of data size (up to 800 GB) and number of processors (up to 80 nodes of the Summit supercomputer), applying the hierarchical clustering approach to hyperspectral imaging and image classification data.
+Our algorithm for Rank-2 NMF scales perfectly on up to 1000s of cores and the entire hierarchical clustering method achieves 5.9x speedup scaling from 10 to 80 nodes on an 800 GB dataset.
 \end{abstract}

### 2020-07-16 latex/paper.tex
stashing the change in paper.tex

--- 2020-07-16 latex/paper.tex
@@
 Not only does this approach uncover more structure in the data than a flat NMF clustering, but also rank-two NMF can be computed more quickly than for general ranks, providing comparable overall time to solution.
+Furthermore, the rank-two NMF algorithm can be parallelized efficiently using a computational kernel similar to matrix-vector multiplication.
 Our data distribution and parallelization strategies are designed to maintain computational load balance throughout the data-dependent hierarchy of computation while limiting interprocess communication, allowing the algorithm to scale to large dense and sparse data sets.
-We demonstrate the scalability of our parallel algorithm in terms of data size (up to 800 GB) and number of processors (up to 80 nodes of the Summit supercomputer), applying the hierarchical clustering approach to hyperspectral imaging and image classification data.
-Our algorithm for Rank-2 NMF scales perfectly on up to 1000s of cores and the entire hierarchical clustering method achieves 5.9x speedup scaling from 10 to 80 nodes on an 800 GB dataset.
+We demonstrate the scalability of our parallel algorithm in terms of data size and number of processors, and apply the hierarchical clustering approach to large (dense) hyperspectral imaging and (sparse) text document data.
+\GB{On 40 nodes of Summit supercomputer, our algorithm achieves (a) linear speedup for solving the Rank-2 Non-negative Least squares kernels and (b) achieves 14x speedup on synthetic matrices.
+We also demonstrate that in the case of real world matrices, the algorithm scales to 80 summit nodes.}
 \end{abstract}
@@
 
-While standard NMF provides a flat clustering, NMF can also be used as a tool for hierarchical clustering.
+Hierarchical clustering is the process of recursively breaking a group into two similar groups until every sample is a cluster by itself.
+That is., we consider all the data points as a single cluster and in each iteration, the data points in the current cluster are split into two separate clusters till there is only one sample in a cluster.
+While standard NMF is interpreted as a flat clustering, it can also be extended for hierarchical clustering.
 Kuang and Park \cite{KP13} propose a method that uses rank-2 NMF to recursively bipartition the samples.
@@
 A single latent feature vector for each node can also be used for interpreting clusters.
-Du et al.
+Similarly, Du et al.
 \cite{DKDP17} show that the latent features vectors of the leaves can also be used to initialize a $\M{W}$ within an algorithm for flat NMF, yielding fast convergence and overall speedup compared to initializing the algorithm with random data.

### 2020-07-16 latex/paper.tex
adding citations

--- 2020-07-16 latex/paper.tex
@@
 The goal of this work is to use distributed-memory parallelism to scale the algorithm to large datasets that require the memory of multiple compute nodes and to high processor counts.
-While flat NMF algorithms have been scaled to HPC platforms \cite{} \GB{add citations beyond our own}, our implementation is the first to our knowledge to scale a hierarchical NMF method to 1000s of cores.
+While flat NMF algorithms have been scaled to HPC platforms \cite{FKPB15} \GB{add citations beyond our own}, our implementation is the first to our knowledge to scale a hierarchical NMF method to 1000s of cores.
 As discussed in detail in \Cref{sec:paralgs}, we choose to parallelize the computations associated with each node in the tree, which involve a Rank-2 NMF and the computation of the node's score.

### 2020-07-16 latex/paper.tex
hierarchical explanation beginning

--- 2020-07-16 latex/paper.tex
@@
 The goal of this work is to use distributed-memory parallelism to scale the algorithm to large datasets that require the memory of multiple compute nodes and to high processor counts.
-While flat NMF algorithms have been scaled to HPC platforms \cite{FKPB15} \GB{add citations beyond our own}, our implementation is the first to our knowledge to scale a hierarchical NMF method to 1000s of cores.
+While flat NMF algorithms have been scaled to HPC platforms \cite{} \GB{add citations beyond our own}, our implementation is the first to our knowledge to scale a hierarchical NMF method to 1000s of cores.
 As discussed in detail in \Cref{sec:paralgs}, we choose to parallelize the computations associated with each node in the tree, which involve a Rank-2 NMF and the computation of the node's score.
@@
 \subsection{Datasets}
-
-\GB{need shortcut names for datasets: synthetic, DC, and something better than \image{}}
-
 \begin{itemize}
@@
 
+From equation \ref{eq:treecost}, we expect to see perfect strong scaling in a computationally bound H2NMF problem with fixed frontier node $k$ count.
+However, the problem should become latency bound with its $\log p$ cost as we scale. Figure \ref{fig:fig:synhierspeedup} shows this latency limit as
+we scale to 50 compute nodes on the synthetic data. Figure \ref{fig:synhierstrongscaling}
+
+From equation \ref{eq:levelcost}
 \begin{itemize}

### 2020-07-16 latex/paper.tex
split the paper.tex into multiple files

--- 2020-07-16 latex/paper.tex
@@
 
-\begin{abstract}
-Nonnegative Matrix Factorization (NMF) is an effective tool for clustering nonnegative data, either for computing a flat partitioning of a dataset or for determining a hierarchy of similarity.
-In this paper, we propose a parallel algorithm for hierarchical clustering that uses a divide-and-conquer approach based on rank-two NMF to split a data set into two cohesive parts.
-Not only does this approach uncover more structure in the data than a flat NMF clustering, but also rank-two NMF can be computed more quickly than for general ranks, providing comparable overall time to solution.
-Our data distribution and parallelization strategies are designed to maintain computational load balance throughout the data-dependent hierarchy of computation while limiting interprocess communication, allowing the algorithm to scale to large dense and sparse data sets.
-We demonstrate the scalability of our parallel algorithm in terms of data size (up to 800 GB) and number of processors (up to 80 nodes of the Summit supercomputer), applying the hierarchical clustering approach to hyperspectral imaging and image classification data.
-Our algorithm for Rank-2 NMF scales perfectly on up to 1000s of cores and the entire hierarchical clustering method achieves 5.9x speedup scaling from 10 to 80 nodes on an 800 GB dataset.
-\end{abstract}
+\input{tex/abstract.tex}
 
@@
 
-\section{Introduction}
+\input{tex/introduction.tex}
 
-Nonnegative Matrix Factorization (NMF) has been demonstrated to be an effective tool for unsupervised learning problems including clustering \cite{XLG03, SBPP06, DHS05}. 
-An NMF consists of two tall-and-skinny non-negative matrices whose product approximates a nonnegative data matrix.
-That is, given an $m\times n$ data matrix $\M{A}$, we seek nonnegative matrices $\M{W}$ and $\M{H}$ that each have $k$ columns so that $\M{A} \approx \M{W}\M{H}^\Tra$.
-Each pair of corresponding columns of $\M{W}$ and $\M{H}$ form a latent component of the NMF.
-If the rows of $\M{A}$ correspond to features and the columns to samples, the $i$th row of the $\M{H}$ matrix represents the loading of sample $i$ onto each latent component and provides a soft clustering.
-Because the $\M{W}$ factor is also nonnegative, each column can typically be interpreted as a latent feature vector for each cluster.
+\input{tex/prelims.tex}
 
-Hierarchical clustering is the process of recursively breaking a group into two similar groups until every sample is a cluster by itself.
-That is., we consider all the data points as a single cluster and in each iteration, the data points in the current cluster are split into two separate clusters till there is only one sample in a cluster.
-While standard NMF is interpreted as a flat clustering, it can also be extended for hierarchical clustering.
-Kuang and Park \cite{KP13} propose a method that uses rank-2 NMF to recursively bipartition the samples.
-An NMF with $k=2$ determines a row of $\M{H}$ of length 2 for each sample that determines its part membership.
-By recursively bipartitioning the samples in each part, the method determines a binary tree such that each sample belongs to a unique leaf and internal nodes of the tree determine higher-level clusters of the samples.
-A single latent feature vector for each node can also be used for interpreting clusters.
-Similarly, Du et al.
-\cite{DKDP17} show that the latent features vectors of the leaves can also be used to initialize a $\M{W}$ within an algorithm for flat NMF, yielding fast convergence and overall speedup compared to initializing the algorithm with random data.
-We discuss the hierarchical method in more detail in \Cref{sec:prelim} and \Cref{sec:seqalgs}.
+\input{tex/algorithm.tex}
 
-We illustrate the output of the hierarchical clustering method with an example data set and output tree.
-Following Gillis et al.
-\cite{GKP15}, we apply the method to a hyperspectral imaging (HSI) data set of the Washington, D.C~national mall, which has pixel dimensions $1280 \times 307$ and 191 spectral bands.
-\Cref{fig:dc} visualizes the output tree with 6 leaves along with their hierarchical relationships.
-The root node, labeled 0, is a flattening of the HSI data to a 2D grayscale image.
-Each other node is represented by an overlay of the member pixels of the clusters (in blue) on the original grayscale image.
-The first bipartitioning separates vegetation (cluster 1) from non-vegetation (cluster 2), the bipartitioning of cluster 1 separates grass (cluster 3) from trees (cluster 4), the bipartitioning of cluster 2 separates buildings (cluster 5) from sidewalks/water (cluster 6), and so on.
-If the algorithm continues, it chooses to split the leaf node that provides the greatest benefit to the overall tree, which can be quantified as a node's ``score'' in various ways.
-The algorithm proceeds to split nodes until a target number of leaves is obtained, until no node is worth splitting, or some other stopping criterion is met.
+\input{tex/analysis.tex}
 
-While the hierarchical clustering method offers advantages in terms of interpretation as well as execution time compared to flat NMF, implementations of the algorithm are limited to single workstations and the dataset must fit in the available memory.
-Currently available implementations can utilize multiple cores via MATLAB \cite{KP13} or explicit shared-memory parallelization in the SmallK library \cite{SmallK}.
-The goal of this work is to use distributed-memory parallelism to scale the algorithm to large datasets that require the memory of multiple compute nodes and to high processor counts.
-While flat NMF algorithms have been scaled to HPC platforms \cite{FKPB15} \GB{add citations beyond our own}, our implementation is the first to our knowledge to scale a hierarchical NMF method to 1000s of cores.
-As discussed in detail in \Cref{sec:paralgs}, we choose to parallelize the computations associated with each node in the tree, which involve a Rank-2 NMF and the computation of the node's score.
-We choose a data matrix distribution across processors that avoids any redistribution of the input matrix regardless of the data-dependent structure of the tree's splitting decisions so that the communication required involves only the small factor matrices.
-Analysis of the algorithm shows the dependence of execution time on computation and communication costs as well as on $k$, the number of clusters computed.
-In particular, we confirm that many of the dominant costs are logarithmic in $k$, which is favorable to the linear or sometimes superlinear dependence of flat NMF algorithms.
+\input{tex/experiments.tex}
 
-We demonstrate in \Cref{sec:results} the efficiency and scalability of our parallel algorithm on three data sets, including the HSI data of the DC mall and an image classification data set involving skin melanoma.
-The experimental results show that our parallelization of Rank-2 NMF is highly scalable, maintaining computation bound performance on 1000s of cores.
-We also show the limits of strong scalability when scaling to large numbers of clusters (leaf nodes), as the execution time shifts to becoming interprocessor bandwidth bound and eventually latency bound.
-The image classification data set requires 800 GB of memory across multiple nodes to process, and in scaling from 10 nodes to 80 nodes of the Summit supercomputer (see \Cref{sec:summit}), we demonstrate parallel speedups of 7.1$\times$ for a single Rank-2 NMF and ZZ$\times$ for a complete hierarchical clustering.
+\input{tex/conclusion.tex}
 
-\GB{we should have perhaps compared against SmallK and also against flat NMF...}
-
-\begin{figure}
-\input{fig/dc}
-\caption{Hierarchical Clustering of DC Mall Hyperspectral Image}
-\label{fig:dc}
-\end{figure}
-
-\section{Preliminaries and Related Work}
-\label{sec:prelim}
-
-\subsection{NMF}
-
-The NMF constrained optimization problem
-$$\min_{\M{W},\M{H}\geq \M{0}} \|\M{A} - \M{W}\M{H}^\Tra\|_2$$
-is nonlinear and nonconvex, and various optimization techniques can be used to approximately solve it.
-A popular approach is to use alternating optimization of the two factor matrices because each subproblem is a nonnegative least squares (NNLS) problem, which is convex and can be solved exactly.
-Many block coordinate descent (BCD) approaches are possible \cite{KHP14}, and one 2-block BCD algorithm that solves the NNLS subproblems exactly is block principal pivoting \cite{KP11}.
-This NNLS algorithm is an active-set-like method that determines the sets of entries in the solution vectors that are zero and those that are positive through an iterative but finite process.
-
-When the rank of the factorization (the number of columns of $\M{W}$ and $\M{H}$) is 2, the NNLS subproblems can be solved much more quickly because the number of possible active sets is only 4.
-As explained in more detail in \Cref{sec:r2nmf}, the optimal solution across the 4 sets can be determined efficiently to solve the NNLS subproblem more quickly than general-rank approaches like block principal pivoting.
-Because of the relative ease of solving the NMF problem for the rank-2 case, Kuang and Park \cite{KP13} propose a recursive method to use a rank-2 NMF to partition the input data into 2 parts, whereby each part can be further partitioned via rank-2 NMF of the corresponding original data.
-This approach yields a hierarchical factorization, potentially uncovering more global structure of the input data and allowing for better scalability of the algorithm to large NMF ranks.
-
-The hierarchical rank-2 NMF method has been applied to document clustering \cite{KP13} and hyperspectral image segmentation \cite{GKP15}.
-The leaves of the tree also yield a set of column vectors that can be aggregated into an approximate $\M{W}$ factor (ignoring their hierarchical structure).
-Using this factor matrix to initialize a higher-rank NMF computation leads to quick convergence and overall faster performance than initializing NMF with random data; this approach is known as Divide-and-Conquer NMF \cite{DKDP17}.
-We focus in this paper on parallelizing the hierarchical algorithms proposed by Kuang and Park \cite{KP13} and Gillis et al. \cite{GKP15}.
-
-\subsection{Parallel NMF}
-
-Scaling algorithms for NMF to large data often requires parallelization in order to fit the data across the memories of multiple compute nodes or speed up the computation to complete in reasonable time.
-Parallelizations of multiple optimization approaches have been proposed for general NMF. \GB{add more citations here, including to SmallK}
-In particular, we build upon the work of Kannan et al.
-\cite{KBP16,KBP17,EH+19-TR} and the open-source library PLANC, designed for nonnegative matrix and tensor factorizations of dense and sparse data.
-In this parallelization, the alternating optimization approach is employed with various options for the algorithm used to (approximately) solve the NNLS subproblems.
-The efficiency of the parallelization is based on scalable algorithms for the parallel matrix multiplications involved in all NNLS algorithms; these algorithms are based on Cartesian distributions of the input matrix across 1D or 2D processor grids.
-
-\subsection{Communication Model}
-
-We use the $\alpha$-$\beta$-$\gamma$ model \cite{TRG05,CH+07,BCDH+14} for analysis of distributed-memory parallel algorithms. 
-In this model, the cost of sending a single message of $n$ words of data between two processors is $\alpha + \beta \cdot n$, so that $\alpha$ represents the latency cost of the message and $\beta$ represents the bandwidth cost of each word in the message.
-The $\gamma$ parameter represents the computational cost of a single floating point operation (flop).
-In this simplified communication model, we ignore contention in the network, assuming in effect a fully connected network, and other limiting factors in practice such as the number of hops between nodes and the network injection rate \cite{GOS16}.
-We let $p$ represent the number of processors available on the machine.
-
-All of the interprocessor communication in the algorithms presented in this work are encapsulated in collective communication operations that involve the full set of processors.
-Algorithms for implementing the collective operations are built out of pairwise send and receive operations, and we assume the most efficient algorithms are used in our analysis \cite{TRG05,CH+07}.
-The collectives used in our algorithms are all-reduce, all-gather, and reduce-scatter.
-In an all-reduce, all processors start out with the same amount of data and all end with a copy of the same result, which is in our case a sum of all the inputs (and the same size as a single input).
-The cost of an all-reduce of size $n$ words is $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n)$ for $n>p$ and $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n\log p)$ for $n<p$.
-In an all-gather, all processors start out with separate data and all end with a copy of the same result, which is the union of all the input data.
-If each processor starts with $n/p$ data and ends with $n$ data, the cost of the all-gather is $\alpha \cdot O(\log p) + \beta \cdot O(n)$.
-In a reduce-scatter, all processors start out with the same amount of data and all end with a subset of the result, which is in our case a sum of all the inputs (and is smaller than its input).
-If each processor starts with $n$ data and ends with $n/p$ data, the cost of the reduce-scatter is $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n)$.
-In the case of all-reduce and reduce-scatter, the computational cost is typically dominated by the bandwidth cost because $\beta \gg \gamma$.
-
-\section{Algorithms}
-
-\subsection{Sequential Algorithms}
-\label{sec:seqalgs}
-
-\subsubsection{Rank-2 NMF}
-\label{sec:r2nmf}
-
-Using the 2-block BCD approach for a rank-2 NMF yields NNLS subproblems of the form $\displaystyle \min_{\M[\bar]{H} \geq \M{0}} \|\M{W}\M[\bar]{H}^\Tra-\M{A}\|$ and $\min_{\M[\bar]{W} \geq \M{0}} \|\M{H}\M[\bar]{W}^\Tra-\M{A}^\Tra\|$.
-In each case, the columns of the transposed variable matrix can be computed independently.
-Considering the $i$th row of $\M[\bar]{H}$, for example, the NNLS problem to solve is
-\begin{align*}
-\begin{multlined}
-\min_{\ME[\bar]{H}{i,1},\ME[\bar]{H}{i,2} \geq 0} \left\|\begin{bmatrix} \V{w}_1 & \V{w}_2 \end{bmatrix} \begin{bmatrix}\ME[\bar]{H}{i,1} \\ \ME[\bar]{H}{i,2} \end{bmatrix} - \V{A}_i \right\| \\ = \min_{\ME[\bar]{H}{i,1},\ME[\bar]{H}{i,2} \geq 0} \left\|\ME[\bar]{H}{i,1} \V{w}_1 + \ME[\bar]{H}{i,2} \V{w}_2 - \V{a}_i \right\|
-\end{multlined}
-\end{align*}
-where $\V{w}_1$ and $\V{W}_2$ are the two columns of $\M{W}$ and $\V{a}_i$ is the $i$ column of $\M{A}$.
-We note that there are four possibilities of solutions, as each of the two scalar variables may be positive or zero.
-
-As shown by Kuang and Park \cite{KP13}, determining which of the four possible solutions is feasible and optimal can be done efficiently by exploiting the following properties:
-\begin{itemize}
-	\item if the solution to the unconstrained least squares problem admits two positive values, it is the optimal solution to the nonnegatively constrained problem,
-	\item if $\M{W}$ and $\M{A}$ are both nonnegative, then the candidate solution with two zero values is never (uniquely) optimal and can be discarded, and
-	\item if the solution to the unconstrained problem does not admit two positive values, the better of the two remaining solutions can be determined by comparing $\V{a}_{j}^\Tra\V{w}_1 / \|\V{w}_1\|$ and $\V{a}_{j}^\Tra\V{w}_2 / \|\V{w}_2\|$.
-\end{itemize}
-If the unconstrained problem is solved via the normal equations, then the temporary matrices computed for the normal equations ($\M{W}^\Tra\M{W}$ and $\M{A}^\Tra\M{W}$) can be re-used to determine the better of the two solutions with a single positive variable.
-
-\Cref{alg:r2nnls} implements this strategy for all rows of $\M{H}$ simultaneously.
-It takes as input the matrices $\M{C}=\M{A}^\Tra\M{W}$ and $\M{G}=\M{W}^\Tra\M{W}$, first solves the normal equations for the unconstrained problem, and then chooses between the two alternate possibilities as necessary.
-We note that each row of $\M{H}$ is independent, and therefore this algorithm is easily parallelized.
-Solving for $\M{W}$ can be done using inputs $\M{C}=\M{A}\M{H}$ and $\M{G}=\M{H}^\Tra\M{H}$.
-
-\begin{algorithm}
-\caption{Rank-2 Nonnegative Least Squares Solve \cite{KP13}}
-\label{alg:r2nnls}
-\begin{algorithmic}[1]
-	\Require{$\M{C}$ is $n\times 2$ and $\M{G}$ is $2\times 2$ and s.p.d.}
-	\Function{$\M{H} =$ Rank2-NLS-Solve}{$\M{C},\M{G}$}
-		\State $\M{H} = \M{C}\M{G}^{-1}$ \hfill \Comment{Solve unconstrained system}
-		\For{$i=1$ to $n$}
-			\If{$\ME{H}{i1} < 0$ or $\ME{H}{i2} < 0$}
-				\State \Comment{Choose between single-variable solutions}
-				\If{$\ME{C}{i1} / \sqrt{\ME{G}{11}}  < \ME{C}{i2} / \sqrt{\ME{G}{22}} $}
-					\State $\ME{H}{i1} = 0$
-					\State $\ME{H}{i2} = \ME{C}{i2} / \ME{G}{22}$
-				\Else
-					\State $\ME{H}{i1} = \ME{C}{i1} / \ME{G}{11}$
-					\State $\ME{H}{i2} = 0$
-				\EndIf
-			\EndIf
-		\EndFor
-	\EndFunction
-	\Ensure{$\displaystyle \M{H} = \argmin_{\M[\bar]{H} \geq \M{0}} \|\M{A}-\M{W}\M[\bar]{H}^\Tra\|$ is $n\times 2$ with $\M{C}=\M{A}^\Tra \M{W}$ and $\M{G}=\M{W}^\Tra\M{W}$}
-\end{algorithmic}
-\end{algorithm}
-
-Given that the computational complexity of \Cref{alg:r2nnls} is $O(n)$ (or $O(m)$ when computing $\M{W}$), and the complexity of computing $\M{W}^\Tra\M{W}$ and $\M{H}^\Tra\M{H}$ is $O(m+n)$, the typical dominant cost of each iteration of Rank-2 NMF is that of computing $\M{A}^\Tra\M{W}$ and $\M{A}\M{H}$, which is $O(mn)$.
-
-\subsubsection{Hierarchical Clustering}
-
-A Rank-2 NMF can be used to partition the columns of the matrix into two parts.
-In this case, the columns of the $\M{W}$ factor represent feature weights for each of the two latent components, and the strength of membership in the two components for each column of $\M{A}$ is given by the two values in the corresponding row of $\M{H}$.
-We can determine part membership by comparing those values: if $\ME{H}{i1} > \ME{H}{i2}$, then column $i$ of $\M{A}$ is assigned to the first part, which is associated with feature vector $\V{w}_1$.
-Membership can be determined by other metrics that also take into account balance across parts or attempt to detect outliers.
-
-Given Rank-2 NMF as a splitting procedure, hierarchical clustering builds a binary tree such that each node corresponds to a subset of samples from the original data set and each node's children correspond to a 2-way partition of the node's samples.
-In this way, the leaves form a partition of the original data, and the internal nodes specify the hierarchical relationship among clusters.
-As the tree is built, nodes are split in order of their score, or relative value to the overall clustering of the data.
-The process can be continued until a target number of leaves is produced or until all remaining leaves have a score below a given threshold.
-
-A node's score can be computed in different ways.
-For document clustering, Kuang and Park \cite{KP13} propose using modified normalized discounted cumulative gain, which measures how distinct a node's children are from each other using the feature weights associated with the node and its children.
-For hyperspectral imaging data, Gillis et al.
-\cite{GKP15} propose using the possible reduction in overall NMF error if the node is split -- the difference in error between using the node itself or using its children.
-We use the latter in our implementation.
-
-In any case, a node's score depends on properties of its children, so the computation for a split must be done before the split is actually accepted.
-To this end, we define a \emph{frontier} node to be a parent of leaves; these are nodes whose children have been computed but whose splits have not been accepted.
-\Cref{fig:tree} depicts the classification of nodes into internal, frontier, and leaf nodes.
-As the tree is built, the algorithm selects the frontier node with the highest score to split, though no computation is required to split the node.
-When a frontier node split is accepted, it becomes an internal node and its children are split (so that their scores can be computed) and added to the set of frontier nodes.
-When the algorithm terminates, the leaves are discarded and the frontier nodes become the leaves of the output tree.
-
-\begin{figure}
-\input{fig/tree}
-\caption{Hierarchy node classification}
-\label{fig:tree}
-\end{figure}
-
-Our hierarchical clustering algorithm is presented in \Cref{alg:hiernmf2} and follows that of Kuang and Park \cite{KP13}.
-Each node includes a field $\M{A}$, which is a subset of columns (samples) of the original data, a feature vector $\V{w}$, which is its corresponding column of the $\M{W}$ matrix from its parent's Rank-2 NMF, a score, and pointers to its left and right children.
-A priority queue $\cal Q$ tracks the frontier nodes so that the node with the highest score is split at each step of the algorithm.
-We use a target number of leaf clusters $k$ as the termination condition.
-When a node is selected from the priority queue, it is removed from the set of frontier nodes and its children are added.
-
-\begin{algorithm}
-\caption{Hierarchical Clustering \cite{KP13}}
-\label{alg:hiernmf2}
-\begin{algorithmic}[1]
-	\Require{$\M{A}$ is $m\times n$, $k$ is target number of leaf clusters}
-	\Function{${\cal T} =$ Hier-R2-NMF}{$\M{A}$}
-		\State ${\cal R} = \text{node}(\M{A})$ \hfill \Comment{create root node}
-		\State \textsc{Split}$({\cal R})$
-		\State inject$({\cal Q},{\cal R}.\text{left})$ \hfill \Comment{create priority queue}
-		\State inject$({\cal Q},{\cal R}.\text{right})$ \hfill \Comment{of frontier nodes}
-		\While{size$({\cal Q}) < k$}
-			\State ${\cal N} = \text{eject}({\cal Q})$ \hfill \Comment{frontier node with max score}
-			\State \textsc{Split}$({\cal N}.\text{left})$ \hfill \Comment{split left child}
-			\State inject$({\cal Q},{\cal N}.\text{left})$ \hfill \Comment{and add to $\cal Q$}
-			\State \textsc{Split}$({\cal N}.\text{right})$ \hfill \Comment{split right child}
-			\State inject$({\cal Q},{\cal N}.\text{right})$ \hfill \Comment{and add to $\cal Q$}
-		\EndWhile
-	\EndFunction
-	\Ensure{$\cal{T}$ is binary tree rooted at $\cal R$ with $k$ frontier nodes, each node has subset of cols of $\M{A}$ and feature vector $\V{w}$}
-\end{algorithmic}
-\end{algorithm}
-
-The splitting procedure is specified in \Cref{alg:split}.
-After the Rank-2 NMF is performed, the $\M{H}$ factor is used to determine part membership, and the columns of the $\M{W}$ factor are assigned to the child nodes.
-The score of the node is computed as the reduction in overall NMF error if the node is split, which can be computed from the principal singular values of the subsets of columns of the node and its children, as given in \Cref{line:score}.
-The principal singular values of the children are computed via the power method.
-Note that the principal singular value of the node itself need not be recomputed as it was needed for its parent's score.
-
-\begin{algorithm}
-\caption{Node Splitting via Rank-Two NMF}
-\label{alg:split}
-\begin{algorithmic}[1]
-	\Require{$\cal N$ has a subset of columns given by field $\M{A}$}
-	\Function{Split}{$\cal N$}
-		\State $[\M{W},\M{H}] = \textsc{Rank2-NMF}({\cal N}.\M{A})$ \hfill \Comment{split $\cal N$}
-		\State partition ${\cal N}.\M{A}$ into $\M{A}_1$ and $\M{A}_2$ using $\M{H}$
-		\State ${\cal N}.\text{left} = \text{node}(\M{A}_1,\V{W}_1)$ \hfill \Comment{create left child}
-		\State ${\cal N}.\text{right} = \text{node}(\M{A}_2,\V{W}_2)$ \hfill \Comment{create right child}
-		\State ${\cal N}.\text{score} = \sigma_1^2(\M{A}_1) + \sigma_1^2(\M{A}_2) - \sigma_1^2({\cal N}.\M{A})$ \label{line:score}
-	\EndFunction
-	\Ensure{$\cal{N}$ has two children and a score}
-\end{algorithmic}
-\end{algorithm}
-
-\subsection{Parallelization}
-\label{sec:paralgs}
-
-In this section, we consider the options for parallelizing Hierarchical Rank-2 NMF Clustering (\Cref{alg:hiernmf2}) and provide an analysis for the approach we take.
-The running time of an algorithm is data dependent because not only does each Rank-2 NMF computation require a variable number of iterations, but also the shape of the tree can vary from a balanced binary tree with $O(\log k)$ levels to a tall, flat tree with $O(k)$ levels.
-For the sake of analysis, we will assume a fixed number of NMF iterations for every node of the tree and we will analyze the cost of complete levels.
-
-The first possibility for parallelization is across the nodes of the tree, as each Rank-2 NMF split is independent.
-We choose not to parallelize across nodes in the tree for two reasons.
-The first reason is that while the NMF computations are independent, choosing which nodes to split may depend on global information.
-In particular, when the global target is to determine $k$ leaf clusters, the nodes must be split in order of their scores, which leads to a serialization of the node splits.
-This serialization might be relaxed using speculative execution, but it risks performing unnecessary computation.
-If the global target is to split all nodes with sufficiently high scores, then this serialization is also avoided and node splits become truly independent.
-We choose not to parallelize in this way to remain agnostic to the global stopping criterion.
-
-The second reason is that parallelizing across nodes requires redistribution of the input data.
-Given a node split by $\hat p$ processors, in order to assign disjoint sets of processors to each child node, each of the $\hat p$ processors would have to redistribute their local data, sending data for samples not in their child's set and receiving data for those in their child's set.
-The communication would be data dependent, but on average, each processor would communicate half of its data in the redistribution set, which could have an all-to-all communication pattern among the $\hat p$ processors.
-For a node with $\hat n$ columns, the communication cost would be at least $O(m\hat n/ \hat p)$ words, which is much larger than the communication cost per iteration of Parallel Rank-2 NMF, as we will see in \Cref{sec:analysis}.
-
-By choosing not to parallelize across nodes in the tree, we employ all $p$ processors on each node, and split nodes in sequence.
-The primary computations used to split a node are the Rank-2 NMF and the score computation, which is based on an approximation of the largest singular value.
-We use an alternating-updating algorithm for Rank-2 NMF as described in \Cref{sec:prelim}, and we parallelize it following the methodology proposed in \cite{EH+19-TR} and presented in \Cref{alg:parrank2nmf}.
-
-The communication cost of the algorithm depends on the parallel distribution of the input matrix data $\M{A}$.
-In order to avoid redistribution of the matrix data, we choose a 1D row distribution so that each processor owns a subset of the rows of $\M{A}$.
-Because the clustering partition splits the columns of $\M{A}$, each processor can partition its local data into left and right children to perform the split without any interprocessor communication.
-If we use a 2D distribution for a given node, then because the partition is data dependent, a data redistribution is required in order to obtain a load balanced distribution of both children.
-\Cref{fig:split} presents a visualization of the node-splitting process using a 1D processor distribution.
-In the following subsections, we describe the parallel algorithms for Rank-2 NMF and approximating the principal singular value given this 1D data distribution and analyze their complexity in the context of the hierarchical clustering algorithm.
-
-\begin{figure}
-\input{fig/split}
-\caption{Parallel splitting using Rank-2 NMF and 1D processor distribution.
-A Rank-2 NMF computes factor matrices $\M{W}$ and $\M{H}$ to approximate $\M{A}$, the values of $\M{H}$ are used to determine child membership of each column (either red or blue), and the corresponding column of the $\M{W}$ matrix represents the part's feature weighting.
-The 1D distribution is depicted for 3 processors to show that the splitting requires no interprocessor redistribution as the children are evenly distributed identically to the parent.}
-\label{fig:split}
-\end{figure}
-
-\subsubsection{Algorithms}
-
-\paragraph{\emph{Parallel Rank-2 NMF}}
-
-\Cref{alg:parrank2nmf} presents the parallelization of an alternating-updating scheme for NMF that uses the exact rank-2 solve algorithm presented in \Cref{alg:r2nnls} to update each factor matrix.
-The algorithm computes the inputs to the rank-2 solves in parallel and then exploits the parallelism across rows of the factor matrix so that each processor solves for a subset of rows simultaneously.
-The distribution of all matrices is 1D row distribution, so that each processor owns a subset of the rows of $\M{A}$, $\M{W}$, and $\M{H}$.
-We use the notation $\M[\hat]{A}$ to refer to the $(m/p) \times n$ local data matrix and $\M[\hat]{W}$ and $\M[\hat]{H}$ to refer to the $(m/p) \times 2$ and $(n/p)\times 2$ local factor matrices.
-With this distribution, the computation of $\M{W}^\Tra\M{W}$ and $\M{H}^\Tra\M{H}$ each is done via local multiplication followed by a single all-reduce collective.
-All processors own the data they need to compute their contribution to $\M{A}^\Tra \M{W}$; in order to distribute the result to compute the rows $\M{H}$ independently, a reduce-scatter collective is used to sum and simultaneously distribute across processors.
-To obtain the data needed to compute $\M[\hat]{W}$, each processor must access all of $\M{H}$, which is performed via an all-gather collective.
-The iteration progresses until a convergence criterion is satisfied.
-For performance benchmarking we use a fixed number of iterations, and in practice we use relative change in objective function value (residual norm).
-
-\begin{algorithm}
-\caption{Parallel Rank-2 NMF}
-\label{alg:parrank2nmf}
-\begin{algorithmic}[1]
-	\Require{$\M{A}$ is $m\times n$ and row-distributed across processors so that $\M[\hat]{A}$ is local $(m/p)\times n$ submatrix}
-	\Function{$[\M{W},\M{H}] =$ Parallel-Rank2-NMF}{$\M{A}$}
-		\State Initialize local $\M[\hat]{W}$ randomly
-		\While{not converged}
-			\State \Comment{Compute $\M{H}$}
-			\State $\M[\hat]{G}_W = \M[\hat]{W}^\Tra \M[\hat]{W}$
-			\State $\M{G}_W = \textsc{All-Reduce}(\M[\hat]{G})$
-			\State $\M[\hat]{B} = \M[\hat]{A}^\Tra \M[\hat]{W}$ %\hfill \Comment{local matrix multiplication}
-			\State $\M[\hat]{C} = \textsc{Reduce-Scatter}(\M[\hat]{B})$
-			\State $\M[\hat]{H} = \textsc{Rank2-NLS-Solve}(\M[\hat]{C},\M{G}_W)$
-			\State \Comment{Compute $\M{W}$}
-			\State $\M[\hat]{G}_H = \M[\hat]{H}^\Tra \M[\hat]{H}$
-			\State $\M{G}_H = \textsc{All-Reduce}(\M[\hat]{G}_H)$
-			\State $\M{H} = \textsc{All-Gather}(\M[\hat]{H})$
-			\State $\M[\hat]{D} = \M[\hat]{A} \M{H}$ %\hfill \Comment{local matrix multiplication}
-			\State $\M[\hat]{W} = \textsc{Rank2-NLS-Solve}(\M[\hat]{D},\M{G}_H)$
-		\EndWhile
-	\EndFunction
-	\Ensure{$\M{A} \approx \M{W}\M{H}^\Tra$ with $\M{W}$, $\M{H}$ row-distributed}
-\end{algorithmic}
-\end{algorithm}
-
-\paragraph{\emph{Parallel Power Method}}
-
-In order to compute the score for a frontier node, we use the difference between the principal singular value of the matrix columns of the node and the sum of those of its children.
-Thus, we must determine the principal singular value of every node in the tree once, including leaf nodes.
-We use the power method to approximate it, repeatedly applying $\M{A}\M{A}^\Tra$ to a vector until it converges to the leading right singular vector.
-We present the power method in \Cref{alg:parpowmeth}.
-Note that we do not normalize the approximate left singular vector so that the computed value approximates the square of the largest singular value.
-
-Given the 1D distribution, only one communication collective is required for the pair of matrix-vector multiplications.
-That is, the approximate right singular vector $\V{v}$ is redundantly owned on each processor, and the approximate left singular vector $\V{u}$ is distributed across processors.
-Each processor can compute its local $\V[\hat]{u}$ from $\V{v}$ without communication and use the result for its contribution to $\V{v}=\M{A}^\Tra \V{u}$.
-An all-reduce collective is used to obtain a copy of $\V{v}$ on every processor for the next iteration, and the norm is redundantly computed without further communication.
-We used the relative change in $\sigma$ as the stopping criterion for benchmarking.
-
-\begin{algorithm}
-\caption{Parallel Power Method}
-\label{alg:parpowmeth}
-\begin{algorithmic}[1]
-	\Require{$\M{A}$ is $m\times n$ and row-distributed across processors so that $\M[\hat]{A}$ is local $(m/p)\times n$ submatrix}
-	\Function{$\sigma =$ Parallel-Power-Method}{$\M{A}$}
-		\State Initialize $\V{v}$ randomly and redundantly
-		\While{not converged}
-			\State $\V[\hat]{u} = \M[\hat]{A} \V{v}$
-			\State $\V[\hat]{z} = \M[\hat]{A}^\Tra \V[\hat]{u}$
-			\State $\V{v} = \textsc{All-Reduce}(\V[\hat]{z})$
-			\State $\sigma = \|\V{v}\|$
-			\State $\V{v} = \V{v}/\sigma$
-		\EndWhile
-	\EndFunction
-	\Ensure{$\sigma \approx \sigma_1^2(\M{A})$ is  redundantly owned by all procs}
-\end{algorithmic}
-\end{algorithm}
-
-\subsubsection{Analysis}
-\label{sec:analysis}
-
-\paragraph{\emph{Parallel Rank-2 NMF}}
-
-Each iteration of \Cref{alg:parrank2nmf} incurs the same cost, so we analyze per-iteration computation and communication costs.
-We first consider the cost of the Rank-2 NNLS solves, which are local computations.
-In the notation of \Cref{alg:r2nnls}, matrix $\M{G}$ is $2\times 2$, so solving the unconstrained system (via Cholesky decomposition) and then choosing between single-positive-variables solutions if necessary requires constant time per row of $\M{C}$.
-Thus, the cost of \Cref{alg:r2nnls} is proportional to the number of rows of the first input matrix.
-In the context of \Cref{alg:parrank2nmf}, the per-iteration computational cost of rank-2 solves is then $O((m+n)/p)$.
-The other local computations are the matrix multiplications $\M[\hat]{W}^\Tra \M[\hat]{W}$ and $\M[\hat]{H}^\Tra \M[\hat]{H}$, which also amount to $O((m+n)/p)$ flops, and $\M[\hat]{A}^\Tra \M[\hat]{W}$ and $\M[\hat]{A} \M{H}$, which require $O(mn/p)$ flops because they involve the data matrix.
-Thus, the computation cost is $\gamma \cdot O((mn+m+n)/p)$ and typically dominated by the multiplications involving $\M{A}$.
-We track the lower order terms corresponding to NNLS solves because their hidden constants are larger than that of the dominating term.
-
-There are four communication collectives each iteration, and each involves all $p$ processors.
-The two all-reduce collectives to compute the Gram matrices of the factor matrices involve $2\times 2$ matrices and incur a communication cost of $(\beta + \alpha) \cdot O(\log p)$.
-The reduce-scatter and all-gather collectives involve $n\times 2$ matrices (the size of $\M{H}$) and require $\beta \cdot O(n) + \alpha \cdot O(\log p)$ in communication cost (we ignore the computation cost of the reduce-scatter because it is typically dominated by the bandwidth cost).
-If the algorithm performs $\imath$ iterations, the overall cost of \Cref{alg:parrank2nmf} is
-\begin{equation}
-\label{eq:r2nmfcost}
-\gamma \cdot O\left( \frac{\imath (mn+m+n)}{p} \right) + \beta \cdot O(\imath n) + \alpha \cdot O(\imath \log p).
-\end{equation}
-
-\paragraph{\emph{Parallel Power Method}}
-
-Similar to the previous analysis, we consider a single iteration of the power method.
-The local computation is dominated by two matrix-vector products involving the local data matrix of size $O(mn/p)$ words, incurring $O(mn/p)$ flops.
-The single communication collective is an all-reduce of the approximate right singular vector, which is of size $n$, incurring $\beta \cdot O(n) + \alpha \cdot O(\log p)$ communication.
-We ignore the $O(n)$ computation cost of normalizing the vector, as it will typically be dominated by the communication cost of the all-reduce.
-Over $\jmath$ iterations, \Cref{alg:parpowmeth} has an overall cost of
-\begin{equation}
-\label{eq:powmethcost}
-\gamma \cdot O\left( \frac{\jmath mn}{p} \right) + \beta \cdot O(\jmath n) + \alpha \cdot O(\jmath \log p).
-\end{equation}
-Note the per-iteration cost of the power method differs by only a constant from the per-iteration cost of Rank-2 NMF.
-Because the power method involves single vectors rather than factor matrices with two columns, its constants are smaller than half the size of their counterparts.
-
-\paragraph{\emph{Hierarchical Clustering}}
-
-To analyze the overall cost of the hierarchical clustering algorithm, we sum the costs over all nodes in the tree.
-Because the shape of the tree is data dependent and affects the overall costs, for the sake of analysis we will analyze only complete levels.
-The number of rows in any node is $m$, the same as the root node, as each splitting corresponds to a partition of the columns.
-Furthermore, because each split is a partition, every column of $\M{A}$ is represented exactly once in every complete level of the tree.
-If we assume that all nodes perform the same number of NMF iterations ($\imath$) and power method iterations $(\jmath)$, then the dominating costs of a node with $\overline{n}$ columns is
-\begin{equation*}
-\label{eq:nodecost}
-\begin{split}
-\gamma \cdot O\left( \frac{(\imath+\jmath) m\overline{n} + \imath (m+\overline{n}) }{p} \right) + \beta \cdot O((\imath+\jmath) \overline{n}) \\ + \alpha \cdot O((\imath+\jmath) \log p).
-\end{split}
-\end{equation*}
-Because the sum of the number of columns across any level of the tree is $n$, the cost of the $\ell$th level of the tree is
-\begin{equation}
-\label{eq:levelcost}
-\begin{split}
-\gamma \cdot O\left( \frac{(\imath+\jmath) mn + \imath m 2^\ell}{p} \right) + \beta \cdot O((\imath+\jmath) n) \\ + \alpha \cdot O((\imath+\jmath) 2^\ell \log p).
-\end{split}
-\end{equation}
-Note that the only costs that depend on the level index $\ell$ are the latency cost and a lower-order computational cost.
-
-Summing over levels and assuming the tree is nearly balanced and has height $O(\log k)$ where $k$ is the number of frontier nodes, we obtain an overall cost of \Cref{alg:hiernmf2} of
-\begin{equation}
-\label{eq:treecost}
-\begin{split}
-\gamma \cdot O\left( \frac{(\imath+\jmath) mn}{p}  \log k + \frac{\imath mk}{p} \right) + \beta \cdot O((\imath+\jmath) n \log k) \\ + \alpha \cdot O((\imath+\jmath) k \log p).
-\end{split}
-\end{equation}
-
-We see that the leading order computational cost is logarithmic in $k$ and perfectly load balanced.
-If the overall running time is dominated by the computation (and in particular the matrix multiplications involving $\M{A}$), we expect near-perfect strong scaling.
-The bandwidth cost is also logarithmic in $k$ but does not scale with the number of processors.
-The latency cost grows most quickly with the target number of clusters $k$ but is also independent of the matrix dimensions $m$ and $n$.
-
-\section{Experimental Results}
-\label{sec:results}
-
-\subsection{Experimental Platform}
-\label{sec:summit}
-
-All the experiments in this section were conducted on Summit. Summit is a supercomputer created by IBM for the Oak Ridge National Laboratory. 
-There are approximately 4,600 nodes on Summit. Each node contains two IBM POWER9 processors on separate sockets with dual NVLINK bricks to facilitate a transfer rate of 
-25 GB/s between the processors. Each node contains 512 GB of DDR4 memory for use by both processors. Each POWER9 processor utilizes 22 IBM SIMD Multi-Cores (SMCs), 
-although one of these SMCs on each processor is dedicated to memory transfer and is therefore not available for computation. 
-For node scaling experiments, all 42 available SMCs were utilized in each node so that every node computed with 42 separate MPI processes.
-Additionally, every node also supports six NVIDIA Volta V100 accelerators but these were unused by our algorithm.
-
-H2NMF uses the Armadillo library (version 9.900.1) for all matrix operations. 
-In Armadillo, sparse matrices are stored in Compressed Sparse Column (CSC) format and dense matrices are stored in column-major ordering.
-On Summit, we linked this version of Armadillo with OpenBLAS (version 0.3.9) and IBM's Spectrum MPI (version 10.3.1.2-20200121).
-
-\GB{specify that $\imath=\jmath=100$ in all experiments, also what do we set $k$ to be?
-fixed number for each dataset?
-any other experimental parameters we should specify?
-could also specify these at the beginning of \Cref{sec:perf}}
-
-\subsection{Datasets}
-
-\GB{need shortcut names for datasets: synthetic, DC, and something better than \image{}}
-
-\begin{itemize}
-	\item \textit{Hyperspectral Imaging}:
-	For Hyperspectral imaging, we chose the Hyperspectral digital imagery collection experiment (HYDICE) image of the Washington DC Mall. We will refer
-	to this dataset as \hyper{}\cite{DC-HYDICE}.
-
-	\hyper{} is formatted into a 3-way tensor representing two spatial dimensions of pixels and one dimension of spectral bands. So, a slice along
-	the spectral band dimnesion would be the full \hyper{} image in that spectral band. For H2NMF, these tensors are flattened so that the rows represent the
-	$191$ spectral bands and the columns represent the $392960$ pixels.
-
-	Larger hyperspectral datasets tend to not grow larger than around 200 spectral bands and have millions or billions of pixels. So, scaling is 
-	limited for this dataset given our data distribution scheme.
-	\item \textit{Image Classification}:
-	Image Classification datasets contain tens of thousands of images of equal size. The dataset we chose is the SIIM-ISIC Melanoma classification
-	dataset, which we will refer to as \image{}\cite{SIIM-ISIC}. \image{} consists of $33126$ RGB training images equally sized at $1024 \times 1024$. Unlike with hyperspectral imaging, the resulting
-	matrix used in H2NMF consists of image pixels along the rows and individual images along the columns. So, the resulting sized matrix is
-	$3145728 \times 33126$, which is approximately 800GB in size.
-
-	Given its size, \image{} can only scale from $10$ Summit nodes.
-	\item \textit{Synthetic Dataset}:
-	Our synthetic dataset is the same shape of \image{} but consists of fewer rows and columns by a factor of $3$. The resulting matrix is $1048576 \times 11042$. 
-	We chose a smaller size dataset to better see the limits of scaling and to scale from fewer Summit nodes.
-\end{itemize}
-
-\subsection{Performance}
-\label{sec:perf}
-
-\subsubsection{Single-Node Scaling for DC Dataset}
-
-\GB{add only speedup plot for entire tree (at least all complete levels) from 1 core to 42 cores}
-
-\begin{itemize}
-	\item emphasize we start with a small data set that fits on one node and has only 191 rows
-	\item using 1 node (2 sockets), because we are memory bandwidth bound we expect limited speedup but we can save time with parallelism even for this small a data set
-\end{itemize}
-
-\subsubsection{Rank-2 NMF Strong Scaling}
-
-We perform strong scaling experiments for a single Rank-2 NMF (\Cref{alg:parrank2nmf}) on the synthetic and \image{} datasets.
-The theory (\Cref{eq:r2nmfcost}) suggests that perfect strong scaling is possible as long as the execution time is dominated by local computation.
-Both the matrix multiplications and NNLS solves scale linearly with $1/p$ (we expect MatMul to dominate), but the bandwidth cost is independent of $p$ and the latency cost increases slightly with $p$.
-
-\Cref{fig:synrank2speedup,fig:rwrank2speedup} show performance relative to the smallest number of nodes required to store data and factor matrices.
-The synthetic data is chosen to fit on a single node, while the \image{} data requires 10 nodes to store the input matrix as well as temporary and output matrices.
-For these data sets, we observe nearly perfect strong scaling, with 47$\times$ speedup on 50 nodes (over 1 node) for synthetic data and 7.1$\times$ speedup on 80 nodes (over 10 nodes) for \image{} data.
-
-\begin{figure}
-\begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_speedup.pdf}
-\caption{Strong Scaling Speedup for Rank-2 NMF on synthetic data}
-\label{fig:synrank2speedup}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_strongscaling.pdf}
-\caption{Relative Time Breakdown for Rank-2 NMF on synthetic data}
-\label{fig:synrank2strongscaling}
-\end{center}
-\end{figure}
-
-The relative time breakdowns are presented in \Cref{fig:synrank2strongscaling,fig:rwrank2strongscaling} and explain the strong scaling performance.
-Each experiment is normalized to 100\% time, so comparisons cannot be readily made across numbers of compute nodes. 
-For both data sets, we see that the time is dominated by MatMul, which is the primary reason for the scalability.
-The dominant matrix multiplications are between a large matrix and a matrix with 2 columns, so it is locally memory bandwidth bound, with performance proportional to the size of the large matrix.
-In each plot, we also see the relative time of all-gather and reduce-scatter increasing, which is because the local computation is decreasing while the communication cost is slightly increasing with $p$.
-This pattern will continue as $p$ increases, which will eventually limit scalability, but for these data sets the MatMul takes around 80\% of the time at over 2000 cores.
-
-\begin{figure}
-\begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_speedup.pdf}
-\caption{Strong Scaling Speedup for Rank-2 NMF on \image{} data}
-\label{fig:rwrank2speedup}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_strongscaling.pdf}
-\caption{Relative Time Breakdown for Rank-2 NMF on \image{} data}
-\label{fig:rwrank2strongscaling}
-\end{center}
-\end{figure}
-
-\subsubsection{Hierarchical Clustering Strong Scaling}
-
-\GB{Do these breakdowns include the score function?  actually this may explain why all-reduce costs more, it also has $O(n)$ data for score...}
-
-\GB{4 plots: 2 data sets with speedup/breakdown, note that we include time for complete levels, ignoring nodes in incomplete levels for easier interpretation}
-
-\begin{itemize}
-	\item theory (\Cref{eq:treecost}) says...
-	\item Fig 7 shows limit of strong scaling for synthetic data (time increases at 50 nodes compared to 40), Fig 8 gives breakdown and explains
-	\begin{itemize}
-		\item drastic performance dropoff may be artifact of all-gather, which should take about the same time as reduce-scatter, but speedup from 30 to 40 nodes is also slight
-		\item by 50 nodes, about 50\% of time in communication; still mostly dominated by bandwidth cost because ag and rs dominate ar
-		\item note the difference between the scaling of single rank-2 and entire tree: poor scaling must be due to lower levels, which we explore in the next section
-	\end{itemize}
-	\item Fig 9 shows reasonable scaling for entire tree -- speedup drops from 7x to 6x over 80 nodes; Fig 10 shows that MM is still 70\% of run time on 80 nodes, which allows for near-linear scaling
-\end{itemize}
-
-\begin{figure}
-\begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hierarchical_speedup.pdf}
-\caption{Strong Scaling Speedup for Hierarchical Clustering. The total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
-\label{fig:synhierspeedup}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hier_strongscaling.pdf}
-\caption{Relative Time Breakdown for Hierarchical Clustering. The plot uses the total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
-\label{fig:synhierstrongscaling}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/realworld_hierarchical_speedup.pdf}
-\caption{Strong Scaling Speedup for Hierarchical Clustering on \image{} Data. The total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
-\label{fig:rwhierspeedup}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/realworld_hier_strongscaling.pdf}
-\caption{Relative Time Breakdown for Hierarchical Clustering on \image{}. The plot uses the total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
-\label{fig:rwhierstrongscaling}
-\end{center}
-\end{figure}
-
-\subsubsection{Level Scaling}
-
-\GB{3 plots: for synthetic level breakdown for 1 node and for 40 nodes; for realworld just for 80 nodes}
-
-\begin{itemize}
-	\item theory says per-level cost is \Cref{eq:levelcost}
-\end{itemize}
-
-\begin{figure}
-\begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_sequential_level_breakdown.pdf}
-\caption{Sequential Breakdown plot for levels 0 to 5. y-axis is absolute wall clock time in seconds}
-\label{fig:seqlevelbreakdown}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_parallel_level_breakdown.pdf}
-\caption{Parallel Breakdown plot for levels 0 to 5 on 40 nodes. y-axis is absolute wall clock time in seconds}
-\label{fig:parallellevelbreakdown}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/realworld_parallel_level_breakdown.pdf}
-\caption{Parallel Breakdown plot for levels 0 to 5 on 80 nodes. y-axis is absolute wall clock time in seconds}
-\label{fig:rwparallellevelbreakdown}
-\end{center}
-\end{figure}
-
-\subsection{DC Clustering Results}
-
-\subsubsection{Hyperspectral Imaging}
-\begin{figure}
-\input{fig/dc}
-\caption{Hierarchical Cluster from DC Mall Image}
-\label{fig:dc}
-\end{figure}
-
-\section{Conclusion}
-
-\begin{itemize}
-	\item consider sparse case
-	\item consider parallelizing across tree
-	\item consider redistribution for 2D
-\end{itemize}
-
-\section*{Acknowledgment}
-
-The authors would like to thank Simin Ma for contributions to the algorithmic analysis and John Farrell for his contributions to the implementation of the parallel algorithm.
+\input{tex/ack.tex}
 

### 2020-07-16 latex/tex/abstract.tex
split the paper.tex into multiple files

\begin{abstract}
Nonnegative Matrix Factorization (NMF) is an effective tool for clustering nonnegative data, either for computing a flat partitioning of a dataset or for determining a hierarchy of similarity.
In this paper, we propose a parallel algorithm for hierarchical clustering that uses a divide-and-conquer approach based on rank-two NMF to split a data set into two cohesive parts.
Not only does this approach uncover more structure in the data than a flat NMF clustering, but also rank-two NMF can be computed more quickly than for general ranks, providing comparable overall time to solution.
Our data distribution and parallelization strategies are designed to maintain computational load balance throughout the data-dependent hierarchy of computation while limiting interprocess communication, allowing the algorithm to scale to large dense and sparse data sets.
We demonstrate the scalability of our parallel algorithm in terms of data size (up to 800 GB) and number of processors (up to 80 nodes of the Summit supercomputer), applying the hierarchical clustering approach to hyperspectral imaging and image classification data. 
Our algorithm for Rank-2 NMF scales perfectly on up to 1000s of cores and the entire hierarchical clustering method achieves 5.9x speedup scaling from 10 to 80 nodes on an 800 GB dataset.
\end{abstract}

### 2020-07-16 latex/tex/ack.tex
split the paper.tex into multiple files

\section*{Acknowledgment}

The authors would like to thank Simin Ma for contributions to the algorithmic analysis and John Farrell for his contributions to the implementation of the parallel algorithm.

### 2020-07-16 latex/tex/algorithm.tex
split the paper.tex into multiple files

\section{Algorithms}

\subsection{Sequential Algorithms}
\label{sec:seqalgs}

\subsubsection{Rank-2 NMF}
\label{sec:r2nmf}

Using the 2-block BCD approach for a rank-2 NMF yields NNLS subproblems of the form $\displaystyle \min_{\M[\bar]{H} \geq \M{0}} \|\M{W}\M[\bar]{H}^\Tra-\M{A}\|$ and $\min_{\M[\bar]{W} \geq \M{0}} \|\M{H}\M[\bar]{W}^\Tra-\M{A}^\Tra\|$.
In each case, the columns of the transposed variable matrix can be computed independently.
Considering the $i$th row of $\M[\bar]{H}$, for example, the NNLS problem to solve is
\begin{align*}
\begin{multlined}
\min_{\ME[\bar]{H}{i,1},\ME[\bar]{H}{i,2} \geq 0} \left\|\begin{bmatrix} \V{w}_1 & \V{w}_2 \end{bmatrix} \begin{bmatrix}\ME[\bar]{H}{i,1} \\ \ME[\bar]{H}{i,2} \end{bmatrix} - \V{A}_i \right\| \\ = \min_{\ME[\bar]{H}{i,1},\ME[\bar]{H}{i,2} \geq 0} \left\|\ME[\bar]{H}{i,1} \V{w}_1 + \ME[\bar]{H}{i,2} \V{w}_2 - \V{a}_i \right\|
\end{multlined}
\end{align*}
where $\V{w}_1$ and $\V{W}_2$ are the two columns of $\M{W}$ and $\V{a}_i$ is the $i$ column of $\M{A}$.
We note that there are four possibilities of solutions, as each of the two scalar variables may be positive or zero.

As shown by Kuang and Park \cite{KP13}, determining which of the four possible solutions is feasible and optimal can be done efficiently by exploiting the following properties:
\begin{itemize}
	\item if the solution to the unconstrained least squares problem admits two positive values, it is the optimal solution to the nonnegatively constrained problem,
	\item if $\M{W}$ and $\M{A}$ are both nonnegative, then the candidate solution with two zero values is never (uniquely) optimal and can be discarded, and
	\item if the solution to the unconstrained problem does not admit two positive values, the better of the two remaining solutions can be determined by comparing $\V{a}_{j}^\Tra\V{w}_1 / \|\V{w}_1\|$ and $\V{a}_{j}^\Tra\V{w}_2 / \|\V{w}_2\|$.
\end{itemize}
If the unconstrained problem is solved via the normal equations, then the temporary matrices computed for the normal equations ($\M{W}^\Tra\M{W}$ and $\M{A}^\Tra\M{W}$) can be re-used to determine the better of the two solutions with a single positive variable.

\Cref{alg:r2nnls} implements this strategy for all rows of $\M{H}$ simultaneously.
It takes as input the matrices $\M{C}=\M{A}^\Tra\M{W}$ and $\M{G}=\M{W}^\Tra\M{W}$, first solves the normal equations for the unconstrained problem, and then chooses between the two alternate possibilities as necessary.
We note that each row of $\M{H}$ is independent, and therefore this algorithm is easily parallelized.
Solving for $\M{W}$ can be done using inputs $\M{C}=\M{A}\M{H}$ and $\M{G}=\M{H}^\Tra\M{H}$.

\begin{algorithm}
\caption{Rank-2 Nonnegative Least Squares Solve \cite{KP13}}
\label{alg:r2nnls}
\begin{algorithmic}[1]
	\Require{$\M{C}$ is $n\times 2$ and $\M{G}$ is $2\times 2$ and s.p.d.}
	\Function{$\M{H} =$ Rank2-NLS-Solve}{$\M{C},\M{G}$}
		\State $\M{H} = \M{C}\M{G}^{-1}$ \hfill \Comment{Solve unconstrained system}
		\For{$i=1$ to $n$}
			\If{$\ME{H}{i1} < 0$ or $\ME{H}{i2} < 0$}
				\State \Comment{Choose between single-variable solutions}
				\If{$\ME{C}{i1} / \sqrt{\ME{G}{11}}  < \ME{C}{i2} / \sqrt{\ME{G}{22}} $}
					\State $\ME{H}{i1} = 0$
					\State $\ME{H}{i2} = \ME{C}{i2} / \ME{G}{22}$
				\Else
					\State $\ME{H}{i1} = \ME{C}{i1} / \ME{G}{11}$
					\State $\ME{H}{i2} = 0$
				\EndIf
			\EndIf
		\EndFor
	\EndFunction
	\Ensure{$\displaystyle \M{H} = \argmin_{\M[\bar]{H} \geq \M{0}} \|\M{A}-\M{W}\M[\bar]{H}^\Tra\|$ is $n\times 2$ with $\M{C}=\M{A}^\Tra \M{W}$ and $\M{G}=\M{W}^\Tra\M{W}$}
\end{algorithmic}
\end{algorithm}

Given that the computational complexity of \Cref{alg:r2nnls} is $O(n)$ (or $O(m)$ when computing $\M{W}$), and the complexity of computing $\M{W}^\Tra\M{W}$ and $\M{H}^\Tra\M{H}$ is $O(m+n)$, the typical dominant cost of each iteration of Rank-2 NMF is that of computing $\M{A}^\Tra\M{W}$ and $\M{A}\M{H}$, which is $O(mn)$.

\subsubsection{Hierarchical Clustering}

A Rank-2 NMF can be used to partition the columns of the matrix into two parts.
In this case, the columns of the $\M{W}$ factor represent feature weights for each of the two latent components, and the strength of membership in the two components for each column of $\M{A}$ is given by the two values in the corresponding row of $\M{H}$.
We can determine part membership by comparing those values: if $\ME{H}{i1} > \ME{H}{i2}$, then column $i$ of $\M{A}$ is assigned to the first part, which is associated with feature vector $\V{w}_1$.
Membership can be determined by other metrics that also take into account balance across parts or attempt to detect outliers.

Given Rank-2 NMF as a splitting procedure, hierarchical clustering builds a binary tree such that each node corresponds to a subset of samples from the original data set and each node's children correspond to a 2-way partition of the node's samples.
In this way, the leaves form a partition of the original data, and the internal nodes specify the hierarchical relationship among clusters.
As the tree is built, nodes are split in order of their score, or relative value to the overall clustering of the data.
The process can be continued until a target number of leaves is produced or until all remaining leaves have a score below a given threshold.

A node's score can be computed in different ways.
For document clustering, Kuang and Park \cite{KP13} propose using modified normalized discounted cumulative gain, which measures how distinct a node's children are from each other using the feature weights associated with the node and its children.
For hyperspectral imaging data, Gillis et al. \cite{GKP15} propose using the possible reduction in overall NMF error if the node is split -- the difference in error between using the node itself or using its children.
We use the latter in our implementation.

In any case, a node's score depends on properties of its children, so the computation for a split must be done before the split is actually accepted.
To this end, we define a \emph{frontier} node to be a parent of leaves; these are nodes whose children have been computed but whose splits have not been accepted.
\Cref{fig:tree} depicts the classification of nodes into internal, frontier, and leaf nodes.
As the tree is built, the algorithm selects the frontier node with the highest score to split, though no computation is required to split the node.
When a frontier node split is accepted, it becomes an internal node and its children are split (so that their scores can be computed) and added to the set of frontier nodes.
When the algorithm terminates, the leaves are discarded and the frontier nodes become the leaves of the output tree.

\begin{figure}
\input{fig/tree}
\caption{Hierarchy node classification}
\label{fig:tree}
\end{figure}

Our hierarchical clustering algorithm is presented in \Cref{alg:hiernmf2} and follows that of Kuang and Park \cite{KP13}.
Each node includes a field $\M{A}$, which is a subset of columns (samples) of the original data, a feature vector $\V{w}$, which is its corresponding column of the $\M{W}$ matrix from its parent's Rank-2 NMF, a score, and pointers to its left and right children.
A priority queue $\cal Q$ tracks the frontier nodes so that the node with the highest score is split at each step of the algorithm.
We use a target number of leaf clusters $k$ as the termination condition.
When a node is selected from the priority queue, it is removed from the set of frontier nodes and its children are added.

\begin{algorithm}
\caption{Hierarchical Clustering \cite{KP13}}
\label{alg:hiernmf2}
\begin{algorithmic}[1]
	\Require{$\M{A}$ is $m\times n$, $k$ is target number of leaf clusters}
	\Function{${\cal T} =$ Hier-R2-NMF}{$\M{A}$}
		\State ${\cal R} = \text{node}(\M{A})$ \hfill \Comment{create root node}
		\State \textsc{Split}$({\cal R})$
		\State inject$({\cal Q},{\cal R}.\text{left})$ \hfill \Comment{create priority queue}
		\State inject$({\cal Q},{\cal R}.\text{right})$ \hfill \Comment{of frontier nodes}
		\While{size$({\cal Q}) < k$}
			\State ${\cal N} = \text{eject}({\cal Q})$ \hfill \Comment{frontier node with max score}
			\State \textsc{Split}$({\cal N}.\text{left})$ \hfill \Comment{split left child}
			\State inject$({\cal Q},{\cal N}.\text{left})$ \hfill \Comment{and add to $\cal Q$}
			\State \textsc{Split}$({\cal N}.\text{right})$ \hfill \Comment{split right child}
			\State inject$({\cal Q},{\cal N}.\text{right})$ \hfill \Comment{and add to $\cal Q$}
		\EndWhile
	\EndFunction
	\Ensure{$\cal{T}$ is binary tree rooted at $\cal R$ with $k$ frontier nodes, each node has subset of cols of $\M{A}$ and feature vector $\V{w}$}
\end{algorithmic}
\end{algorithm}

The splitting procedure is specified in \Cref{alg:split}.
After the Rank-2 NMF is performed, the $\M{H}$ factor is used to determine part membership, and the columns of the $\M{W}$ factor are assigned to the child nodes.
The score of the node is computed as the reduction in overall NMF error if the node is split, which can be computed from the principal singular values of the subsets of columns of the node and its children, as given in \Cref{line:score}.
The principal singular values of the children are computed via the power method.
Note that the principal singular value of the node itself need not be recomputed as it was needed for its parent's score.

\begin{algorithm}
\caption{Node Splitting via Rank-Two NMF}
\label{alg:split}
\begin{algorithmic}[1]
	\Require{$\cal N$ has a subset of columns given by field $\M{A}$}
	\Function{Split}{$\cal N$}
		\State $[\M{W},\M{H}] = \textsc{Rank2-NMF}({\cal N}.\M{A})$ \hfill \Comment{split $\cal N$}
		\State partition ${\cal N}.\M{A}$ into $\M{A}_1$ and $\M{A}_2$ using $\M{H}$
		\State ${\cal N}.\text{left} = \text{node}(\M{A}_1,\V{W}_1)$ \hfill \Comment{create left child}
		\State ${\cal N}.\text{right} = \text{node}(\M{A}_2,\V{W}_2)$ \hfill \Comment{create right child}
		\State ${\cal N}.\text{score} = \sigma_1^2(\M{A}_1) + \sigma_1^2(\M{A}_2) - \sigma_1^2({\cal N}.\M{A})$ \label{line:score}
	\EndFunction
	\Ensure{$\cal{N}$ has two children and a score}
\end{algorithmic}
\end{algorithm}

\subsection{Parallelization}
\label{sec:paralgs}

In this section, we consider the options for parallelizing Hierarchical Rank-2 NMF Clustering (\Cref{alg:hiernmf2}) and provide an analysis for the approach we take.
The running time of an algorithm is data dependent because not only does each Rank-2 NMF computation require a variable number of iterations, but also the shape of the tree can vary from a balanced binary tree with $O(\log k)$ levels to a tall, flat tree with $O(k)$ levels.
For the sake of analysis, we will assume a fixed number of NMF iterations for every node of the tree and we will analyze the cost of complete levels.

The first possibility for parallelization is across the nodes of the tree, as each Rank-2 NMF split is independent.
We choose not to parallelize across nodes in the tree for two reasons.
The first reason is that while the NMF computations are independent, choosing which nodes to split may depend on global information.
In particular, when the global target is to determine $k$ leaf clusters, the nodes must be split in order of their scores, which leads to a serialization of the node splits.
This serialization might be relaxed using speculative execution, but it risks performing unnecessary computation.
If the global target is to split all nodes with sufficiently high scores, then this serialization is also avoided and node splits become truly independent.
We choose not to parallelize in this way to remain agnostic to the global stopping criterion.

The second reason is that parallelizing across nodes requires redistribution of the input data.
Given a node split by $\hat p$ processors, in order to assign disjoint sets of processors to each child node, each of the $\hat p$ processors would have to redistribute their local data, sending data for samples not in their child's set and receiving data for those in their child's set.
The communication would be data dependent, but on average, each processor would communicate half of its data in the redistribution set, which could have an all-to-all communication pattern among the $\hat p$ processors.
For a node with $\hat n$ columns, the communication cost would be at least $O(m\hat n/ \hat p)$ words, which is much larger than the communication cost per iteration of Parallel Rank-2 NMF, as we will see in \Cref{sec:analysis}.

By choosing not to parallelize across nodes in the tree, we employ all $p$ processors on each node, and split nodes in sequence.
The primary computations used to split a node are the Rank-2 NMF and the score computation, which is based on an approximation of the largest singular value.
We use an alternating-updating algorithm for Rank-2 NMF as described in \Cref{sec:prelim}, and we parallelize it following the methodology proposed in \cite{EH+19-TR} and presented in \Cref{alg:parrank2nmf}.

The communication cost of the algorithm depends on the parallel distribution of the input matrix data $\M{A}$.
In order to avoid redistribution of the matrix data, we choose a 1D row distribution so that each processor owns a subset of the rows of $\M{A}$.
Because the clustering partition splits the columns of $\M{A}$, each processor can partition its local data into left and right children to perform the split without any interprocessor communication.
If we use a 2D distribution for a given node, then because the partition is data dependent, a data redistribution is required in order to obtain a load balanced distribution of both children.
\Cref{fig:split} presents a visualization of the node-splitting process using a 1D processor distribution.
In the following subsections, we describe the parallel algorithms for Rank-2 NMF and approximating the principal singular value given this 1D data distribution and analyze their complexity in the context of the hierarchical clustering algorithm.

\begin{figure}
\input{fig/split}
\caption{Parallel splitting using Rank-2 NMF and 1D processor distribution.  A Rank-2 NMF computes factor matrices $\M{W}$ and $\M{H}$ to approximate $\M{A}$, the values of $\M{H}$ are used to determine child membership of each column (either red or blue), and the corresponding column of the $\M{W}$ matrix represents the part's feature weighting.  The 1D distribution is depicted for 3 processors to show that the splitting requires no interprocessor redistribution as the children are evenly distributed identically to the parent.}
\label{fig:split}
\end{figure}

\subsubsection{Algorithms}

\paragraph{\emph{Parallel Rank-2 NMF}}

\Cref{alg:parrank2nmf} presents the parallelization of an alternating-updating scheme for NMF that uses the exact rank-2 solve algorithm presented in \Cref{alg:r2nnls} to update each factor matrix.
The algorithm computes the inputs to the rank-2 solves in parallel and then exploits the parallelism across rows of the factor matrix so that each processor solves for a subset of rows simultaneously.
The distribution of all matrices is 1D row distribution, so that each processor owns a subset of the rows of $\M{A}$, $\M{W}$, and $\M{H}$.
We use the notation $\M[\hat]{A}$ to refer to the $(m/p) \times n$ local data matrix and $\M[\hat]{W}$ and $\M[\hat]{H}$ to refer to the $(m/p) \times 2$ and $(n/p)\times 2$ local factor matrices.
With this distribution, the computation of $\M{W}^\Tra\M{W}$ and $\M{H}^\Tra\M{H}$ each is done via local multiplication followed by a single all-reduce collective.
All processors own the data they need to compute their contribution to $\M{A}^\Tra \M{W}$; in order to distribute the result to compute the rows $\M{H}$ independently, a reduce-scatter collective is used to sum and simultaneously distribute across processors.
To obtain the data needed to compute $\M[\hat]{W}$, each processor must access all of $\M{H}$, which is performed via an all-gather collective.
The iteration progresses until a convergence criterion is satisfied.
For performance benchmarking we use a fixed number of iterations, and in practice we use relative change in objective function value (residual norm).

\begin{algorithm}
\caption{Parallel Rank-2 NMF}
\label{alg:parrank2nmf}
\begin{algorithmic}[1]
	\Require{$\M{A}$ is $m\times n$ and row-distributed across processors so that $\M[\hat]{A}$ is local $(m/p)\times n$ submatrix}
	\Function{$[\M{W},\M{H}] =$ Parallel-Rank2-NMF}{$\M{A}$}
		\State Initialize local $\M[\hat]{W}$ randomly
		\While{not converged}
			\State \Comment{Compute $\M{H}$}
			\State $\M[\hat]{G}_W = \M[\hat]{W}^\Tra \M[\hat]{W}$
			\State $\M{G}_W = \textsc{All-Reduce}(\M[\hat]{G})$
			\State $\M[\hat]{B} = \M[\hat]{A}^\Tra \M[\hat]{W}$ %\hfill \Comment{local matrix multiplication}
			\State $\M[\hat]{C} = \textsc{Reduce-Scatter}(\M[\hat]{B})$
			\State $\M[\hat]{H} = \textsc{Rank2-NLS-Solve}(\M[\hat]{C},\M{G}_W)$
			\State \Comment{Compute $\M{W}$}
			\State $\M[\hat]{G}_H = \M[\hat]{H}^\Tra \M[\hat]{H}$
			\State $\M{G}_H = \textsc{All-Reduce}(\M[\hat]{G}_H)$
			\State $\M{H} = \textsc{All-Gather}(\M[\hat]{H})$
			\State $\M[\hat]{D} = \M[\hat]{A} \M{H}$ %\hfill \Comment{local matrix multiplication}
			\State $\M[\hat]{W} = \textsc{Rank2-NLS-Solve}(\M[\hat]{D},\M{G}_H)$
		\EndWhile
	\EndFunction
	\Ensure{$\M{A} \approx \M{W}\M{H}^\Tra$ with $\M{W}$, $\M{H}$ row-distributed}
\end{algorithmic}
\end{algorithm}

\paragraph{\emph{Parallel Power Method}}

In order to compute the score for a frontier node, we use the difference between the principal singular value of the matrix columns of the node and the sum of those of its children.
Thus, we must determine the principal singular value of every node in the tree once, including leaf nodes.
We use the power method to approximate it, repeatedly applying $\M{A}\M{A}^\Tra$ to a vector until it converges to the leading right singular vector.
We present the power method in \Cref{alg:parpowmeth}.
Note that we do not normalize the approximate left singular vector so that the computed value approximates the square of the largest singular value.

Given the 1D distribution, only one communication collective is required for the pair of matrix-vector multiplications.
That is, the approximate right singular vector $\V{v}$ is redundantly owned on each processor, and the approximate left singular vector $\V{u}$ is distributed across processors.
Each processor can compute its local $\V[\hat]{u}$ from $\V{v}$ without communication and use the result for its contribution to $\V{v}=\M{A}^\Tra \V{u}$.
An all-reduce collective is used to obtain a copy of $\V{v}$ on every processor for the next iteration, and the norm is redundantly computed without further communication.
We used the relative change in $\sigma$ as the stopping criterion for benchmarking.

\begin{algorithm}
\caption{Parallel Power Method}
\label{alg:parpowmeth}
\begin{algorithmic}[1]
	\Require{$\M{A}$ is $m\times n$ and row-distributed across processors so that $\M[\hat]{A}$ is local $(m/p)\times n$ submatrix}
	\Function{$\sigma =$ Parallel-Power-Method}{$\M{A}$}
		\State Initialize $\V{v}$ randomly and redundantly
		\While{not converged}
			\State $\V[\hat]{u} = \M[\hat]{A} \V{v}$
			\State $\V[\hat]{z} = \M[\hat]{A}^\Tra \V[\hat]{u}$
			\State $\V{v} = \textsc{All-Reduce}(\V[\hat]{z})$
			\State $\sigma = \|\V{v}\|$
			\State $\V{v} = \V{v}/\sigma$
		\EndWhile
	\EndFunction
	\Ensure{$\sigma \approx \sigma_1^2(\M{A})$ is  redundantly owned by all procs}
\end{algorithmic}
\end{algorithm}

### 2020-07-16 latex/tex/analysis.tex
split the paper.tex into multiple files

\subsubsection{Analysis}
\label{sec:analysis}

\paragraph{\emph{Parallel Rank-2 NMF}}

Each iteration of \Cref{alg:parrank2nmf} incurs the same cost, so we analyze per-iteration computation and communication costs.
We first consider the cost of the Rank-2 NNLS solves, which are local computations.
In the notation of \Cref{alg:r2nnls}, matrix $\M{G}$ is $2\times 2$, so solving the unconstrained system (via Cholesky decomposition) and then choosing between single-positive-variables solutions if necessary requires constant time per row of $\M{C}$.
Thus, the cost of \Cref{alg:r2nnls} is proportional to the number of rows of the first input matrix.
In the context of \Cref{alg:parrank2nmf}, the per-iteration computational cost of rank-2 solves is then $O((m+n)/p)$.
The other local computations are the matrix multiplications $\M[\hat]{W}^\Tra \M[\hat]{W}$ and $\M[\hat]{H}^\Tra \M[\hat]{H}$, which also amount to $O((m+n)/p)$ flops, and $\M[\hat]{A}^\Tra \M[\hat]{W}$ and $\M[\hat]{A} \M{H}$, which require $O(mn/p)$ flops because they involve the data matrix.
Thus, the computation cost is $\gamma \cdot O((mn+m+n)/p)$ and typically dominated by the multiplications involving $\M{A}$.
We track the lower order terms corresponding to NNLS solves because their hidden constants are larger than that of the dominating term.

There are four communication collectives each iteration, and each involves all $p$ processors.
The two all-reduce collectives to compute the Gram matrices of the factor matrices involve $2\times 2$ matrices and incur a communication cost of $(\beta + \alpha) \cdot O(\log p)$.
The reduce-scatter and all-gather collectives involve $n\times 2$ matrices (the size of $\M{H}$) and require $\beta \cdot O(n) + \alpha \cdot O(\log p)$ in communication cost (we ignore the computation cost of the reduce-scatter because it is typically dominated by the bandwidth cost).
If the algorithm performs $\imath$ iterations, the overall cost of \Cref{alg:parrank2nmf} is
\begin{equation}
\label{eq:r2nmfcost}
\gamma \cdot O\left( \frac{\imath (mn+m+n)}{p} \right) + \beta \cdot O(\imath n) + \alpha \cdot O(\imath \log p).
\end{equation}

\paragraph{\emph{Parallel Power Method}}

Similar to the previous analysis, we consider a single iteration of the power method.
The local computation is dominated by two matrix-vector products involving the local data matrix of size $O(mn/p)$ words, incurring $O(mn/p)$ flops.
The single communication collective is an all-reduce of the approximate right singular vector, which is of size $n$, incurring $\beta \cdot O(n) + \alpha \cdot O(\log p)$ communication.
We ignore the $O(n)$ computation cost of normalizing the vector, as it will typically be dominated by the communication cost of the all-reduce.
Over $\jmath$ iterations, \Cref{alg:parpowmeth} has an overall cost of
\begin{equation}
\label{eq:powmethcost}
\gamma \cdot O\left( \frac{\jmath mn}{p} \right) + \beta \cdot O(\jmath n) + \alpha \cdot O(\jmath \log p).
\end{equation}
Note the per-iteration cost of the power method differs by only a constant from the per-iteration cost of Rank-2 NMF.
Because the power method involves single vectors rather than factor matrices with two columns, its constants are smaller than half the size of their counterparts.

\paragraph{\emph{Hierarchical Clustering}}

To analyze the overall cost of the hierarchical clustering algorithm, we sum the costs over all nodes in the tree.
Because the shape of the tree is data dependent and affects the overall costs, for the sake of analysis we will analyze only complete levels.
The number of rows in any node is $m$, the same as the root node, as each splitting corresponds to a partition of the columns.
Furthermore, because each split is a partition, every column of $\M{A}$ is represented exactly once in every complete level of the tree.
If we assume that all nodes perform the same number of NMF iterations ($\imath$) and power method iterations $(\jmath)$, then the dominating costs of a node with $\overline{n}$ columns is
\begin{equation*}
\label{eq:nodecost}
\begin{split}
\gamma \cdot O\left( \frac{(\imath+\jmath) m\overline{n} + \imath (m+\overline{n}) }{p} \right) + \beta \cdot O((\imath+\jmath) \overline{n}) \\ + \alpha \cdot O((\imath+\jmath) \log p).
\end{split}
\end{equation*}
Because the sum of the number of columns across any level of the tree is $n$, the cost of the $\ell$th level of the tree is
\begin{equation}
\label{eq:levelcost}
\begin{split}
\gamma \cdot O\left( \frac{(\imath+\jmath) mn + \imath m 2^\ell}{p} \right) + \beta \cdot O((\imath+\jmath) n) \\ + \alpha \cdot O((\imath+\jmath) 2^\ell \log p).
\end{split}
\end{equation}
Note that the only costs that depend on the level index $\ell$ are the latency cost and a lower-order computational cost.

Summing over levels and assuming the tree is nearly balanced and has height $O(\log k)$ where $k$ is the number of frontier nodes, we obtain an overall cost of \Cref{alg:hiernmf2} of
\begin{equation}
\label{eq:treecost}
\begin{split}
\gamma \cdot O\left( \frac{(\imath+\jmath) mn}{p}  \log k + \frac{\imath mk}{p} \right) + \beta \cdot O((\imath+\jmath) n \log k) \\ + \alpha \cdot O((\imath+\jmath) k \log p).
\end{split}
\end{equation}

We see that the leading order computational cost is logarithmic in $k$ and perfectly load balanced.
If the overall running time is dominated by the computation (and in particular the matrix multiplications involving $\M{A}$), we expect near-perfect strong scaling.
The bandwidth cost is also logarithmic in $k$ but does not scale with the number of processors.
The latency cost grows most quickly with the target number of clusters $k$ but is also independent of the matrix dimensions $m$ and $n$.

### 2020-07-16 latex/tex/conclusion.tex
split the paper.tex into multiple files

\section{Conclusion}

\begin{itemize}
	\item consider sparse case
	\item consider parallelizing across tree
	\item consider redistribution for 2D
\end{itemize}

### 2020-07-16 latex/tex/experiments.tex
split the paper.tex into multiple files

\section{Experimental Results}
\label{sec:results}

\subsection{Experimental Platform}
\label{sec:summit}

All the experiments in this section were conducted on Summit. Summit is a supercomputer created by IBM for the Oak Ridge National Laboratory. 
There are approximately 4,600 nodes on Summit. Each node contains two IBM POWER9 processors on separate sockets with dual NVLINK bricks to facilitate a transfer rate of 
25 GB/s between the processors. Each node contains 512 GB of DDR4 memory for use by both processors. Each POWER9 processor utilizes 22 IBM SIMD Multi-Cores (SMCs), 
although one of these SMCs on each processor is dedicated to memory transfer and is therefore not available for computation. 
For node scaling experiments, all 42 available SMCs were utilized in each node so that every node computed with 42 separate MPI processes.
Additionally, every node also supports six NVIDIA Volta V100 accelerators but these were unused by our algorithm.

H2NMF uses the Armadillo library (version 9.900.1) for all matrix operations. 
In Armadillo, sparse matrices are stored in Compressed Sparse Column (CSC) format and dense matrices are stored in column-major ordering.
On Summit, we linked this version of Armadillo with OpenBLAS (version 0.3.9) and IBM's Spectrum MPI (version 10.3.1.2-20200121).

\GB{specify that $\imath=\jmath=100$ in all experiments, also what do we set $k$ to be?  fixed number for each dataset?  any other experimental parameters we should specify? could also specify these at the beginning of \Cref{sec:perf}}

\subsection{Datasets}
\begin{itemize}
	\item \textit{Hyperspectral Imaging}:
	For Hyperspectral imaging, we chose the Hyperspectral digital imagery collection experiment (HYDICE) image of the Washington DC Mall. We will refer
	to this dataset as \hyper{}\cite{DC-HYDICE}.

	\hyper{} is formatted into a 3-way tensor representing two spatial dimensions of pixels and one dimension of spectral bands. So, a slice along
	the spectral band dimnesion would be the full \hyper{} image in that spectral band. For H2NMF, these tensors are flattened so that the rows represent the
	$191$ spectral bands and the columns represent the $392960$ pixels.

	Larger hyperspectral datasets tend to not grow larger than around 200 spectral bands and have millions or billions of pixels. So, scaling is 
	limited for this dataset given our data distribution scheme.
	\item \textit{Image Classification}:
	Image Classification datasets contain tens of thousands of images of equal size. The dataset we chose is the SIIM-ISIC Melanoma classification
	dataset, which we will refer to as \image{}\cite{SIIM-ISIC}. \image{} consists of $33126$ RGB training images equally sized at $1024 \times 1024$. Unlike with hyperspectral imaging, the resulting
	matrix used in H2NMF consists of image pixels along the rows and individual images along the columns. So, the resulting sized matrix is
	$3145728 \times 33126$, which is approximately 800GB in size.

	Given its size, \image{} can only scale from $10$ Summit nodes.
	\item \textit{Synthetic Dataset}:
	Our synthetic dataset is the same shape of \image{} but consists of fewer rows and columns by a factor of $3$. The resulting matrix is $1048576 \times 11042$. 
	We chose a smaller size dataset to better see the limits of scaling and to scale from fewer Summit nodes.
\end{itemize}

\subsection{Performance}
\label{sec:perf}

\subsubsection{Single-Node Scaling for DC Dataset}

\GB{add only speedup plot for entire tree (at least all complete levels) from 1 core to 42 cores}

\begin{itemize}
	\item emphasize we start with a small data set that fits on one node and has only 191 rows
	\item using 1 node (2 sockets), because we are memory bandwidth bound we expect limited speedup but we can save time with parallelism even for this small a data set
\end{itemize}

\subsubsection{Rank-2 NMF Strong Scaling}

We perform strong scaling experiments for a single Rank-2 NMF (\Cref{alg:parrank2nmf}) on the synthetic and \image{} datasets.
The theory (\Cref{eq:r2nmfcost}) suggests that perfect strong scaling is possible as long as the execution time is dominated by local computation.
Both the matrix multiplications and NNLS solves scale linearly with $1/p$ (we expect MatMul to dominate), but the bandwidth cost is independent of $p$ and the latency cost increases slightly with $p$.

\Cref{fig:synrank2speedup,fig:rwrank2speedup} show performance relative to the smallest number of nodes required to store data and factor matrices.
The synthetic data is chosen to fit on a single node, while the \image{} data requires 10 nodes to store the input matrix as well as temporary and output matrices.
For these data sets, we observe nearly perfect strong scaling, with 47$\times$ speedup on 50 nodes (over 1 node) for synthetic data and 7.1$\times$ speedup on 80 nodes (over 10 nodes) for \image{} data.

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_speedup.pdf}
\caption{Strong Scaling Speedup for Rank-2 NMF on synthetic data}
\label{fig:synrank2speedup}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_strongscaling.pdf}
\caption{Relative Time Breakdown for Rank-2 NMF on synthetic data}
\label{fig:synrank2strongscaling}
\end{center}
\end{figure}

The relative time breakdowns are presented in \Cref{fig:synrank2strongscaling,fig:rwrank2strongscaling} and explain the strong scaling performance.
Each experiment is normalized to 100\% time, so comparisons cannot be readily made across numbers of compute nodes. 
For both data sets, we see that the time is dominated by MatMul, which is the primary reason for the scalability.
The dominant matrix multiplications are between a large matrix and a matrix with 2 columns, so it is locally memory bandwidth bound, with performance proportional to the size of the large matrix.
In each plot, we also see the relative time of all-gather and reduce-scatter increasing, which is because the local computation is decreasing while the communication cost is slightly increasing with $p$.
This pattern will continue as $p$ increases, which will eventually limit scalability, but for these data sets the MatMul takes around 80\% of the time at over 2000 cores.

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_speedup.pdf}
\caption{Strong Scaling Speedup for Rank-2 NMF on \image{} data}
\label{fig:rwrank2speedup}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_strongscaling.pdf}
\caption{Relative Time Breakdown for Rank-2 NMF on \image{} data}
\label{fig:rwrank2strongscaling}
\end{center}
\end{figure}

\subsubsection{Hierarchical Clustering Strong Scaling}

\GB{Do these breakdowns include the score function?  actually this may explain why all-reduce costs more, it also has $O(n)$ data for score...}

\GB{4 plots: 2 data sets with speedup/breakdown, note that we include time for complete levels, ignoring nodes in incomplete levels for easier interpretation}

From equation \ref{eq:treecost}, we expect to see perfect strong scaling in a computationally bound H2NMF problem with fixed frontier node $k$ count.
However, the problem should become latency bound with its $\log p$ cost as we scale. Figure \ref{fig:fig:synhierspeedup} shows this latency limit as
we scale to 50 compute nodes on the synthetic data. Figure \ref{fig:synhierstrongscaling}

From equation \ref{eq:levelcost}
\begin{itemize}
	\item theory (\Cref{eq:treecost}) says...
	\item Fig 7 shows limit of strong scaling for synthetic data (time increases at 50 nodes compared to 40), Fig 8 gives breakdown and explains
	\begin{itemize}
		\item drastic performance dropoff may be artifact of all-gather, which should take about the same time as reduce-scatter, but speedup from 30 to 40 nodes is also slight
		\item by 50 nodes, about 50\% of time in communication; still mostly dominated by bandwidth cost because ag and rs dominate ar
		\item note the difference between the scaling of single rank-2 and entire tree: poor scaling must be due to lower levels, which we explore in the next section
	\end{itemize}
	\item Fig 9 shows reasonable scaling for entire tree -- speedup drops from 7x to 6x over 80 nodes; Fig 10 shows that MM is still 70\% of run time on 80 nodes, which allows for near-linear scaling
\end{itemize}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hierarchical_speedup.pdf}
\caption{Strong Scaling Speedup for Hierarchical Clustering. The total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
\label{fig:synhierspeedup}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hier_strongscaling.pdf}
\caption{Relative Time Breakdown for Hierarchical Clustering. The plot uses the total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
\label{fig:synhierstrongscaling}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/realworld_hierarchical_speedup.pdf}
\caption{Strong Scaling Speedup for Hierarchical Clustering on \image{} Data. The total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
\label{fig:rwhierspeedup}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/realworld_hier_strongscaling.pdf}
\caption{Relative Time Breakdown for Hierarchical Clustering on \image{}. The plot uses the total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
\label{fig:rwhierstrongscaling}
\end{center}
\end{figure}

\subsubsection{Level Scaling}

\GB{3 plots: for synthetic level breakdown for 1 node and for 40 nodes; for realworld just for 80 nodes}

\begin{itemize}
	\item theory says per-level cost is \Cref{eq:levelcost}
\end{itemize}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_sequential_level_breakdown.pdf}
\caption{Sequential Breakdown plot for levels 0 to 5. y-axis is absolute wall clock time in seconds}
\label{fig:seqlevelbreakdown}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_parallel_level_breakdown.pdf}
\caption{Parallel Breakdown plot for levels 0 to 5 on 40 nodes. y-axis is absolute wall clock time in seconds}
\label{fig:parallellevelbreakdown}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/realworld_parallel_level_breakdown.pdf}
\caption{Parallel Breakdown plot for levels 0 to 5 on 80 nodes. y-axis is absolute wall clock time in seconds}
\label{fig:rwparallellevelbreakdown}
\end{center}
\end{figure}

\subsection{DC Clustering Results}

\subsubsection{Hyperspectral Imaging}
\begin{figure}
\input{fig/dc}
\caption{Hierarchical Cluster from DC Mall Image}
\label{fig:dc}
\end{figure}

### 2020-07-16 latex/tex/introduction.tex
split the paper.tex into multiple files

\section{Introduction}

Nonnegative Matrix Factorization (NMF) has been demonstrated to be an effective tool for unsupervised learning problems including clustering \cite{XLG03, SBPP06, DHS05}. 
An NMF consists of two tall-and-skinny non-negative matrices whose product approximates a nonnegative data matrix.
That is, given an $m\times n$ data matrix $\M{A}$, we seek nonnegative matrices $\M{W}$ and $\M{H}$ that each have $k$ columns so that $\M{A} \approx \M{W}\M{H}^\Tra$.
Each pair of corresponding columns of $\M{W}$ and $\M{H}$ form a latent component of the NMF.
If the rows of $\M{A}$ correspond to features and the columns to samples, the $i$th row of the $\M{H}$ matrix represents the loading of sample $i$ onto each latent component and provides a soft clustering.
Because the $\M{W}$ factor is also nonnegative, each column can typically be interpreted as a latent feature vector for each cluster.

Hierarchical clustering is the process of recursively breaking a group into two similar groups until every sample is a cluster by itself. That is., we consider all the data points as a single cluster and in each iteration, the data points in the current cluster are split into two separate clusters till there is only one sample in a cluster.
While standard NMF is interpreted as a flat clustering, it can also be extended for hierarchical clustering.
Kuang and Park \cite{KP13} propose a method that uses rank-2 NMF to recursively bipartition the samples.
An NMF with $k=2$ determines a row of $\M{H}$ of length 2 for each sample that determines its part membership.
By recursively bipartitioning the samples in each part, the method determines a binary tree such that each sample belongs to a unique leaf and internal nodes of the tree determine higher-level clusters of the samples.
A single latent feature vector for each node can also be used for interpreting clusters.
Similarly, Du et al. \cite{DKDP17} show that the latent features vectors of the leaves can also be used to initialize a $\M{W}$ within an algorithm for flat NMF, yielding fast convergence and overall speedup compared to initializing the algorithm with random data.
We discuss the hierarchical method in more detail in \Cref{sec:prelim} and \Cref{sec:seqalgs}.

We illustrate the output of the hierarchical clustering method with an example data set and output tree.
Following Gillis et al. \cite{GKP15}, we apply the method to a hyperspectral imaging (HSI) data set of the Washington, D.C~national mall, which has pixel dimensions $1280 \times 307$ and 191 spectral bands.
\Cref{fig:dc} visualizes the output tree with 6 leaves along with their hierarchical relationships.
The root node, labeled 0, is a flattening of the HSI data to a 2D grayscale image.
Each other node is represented by an overlay of the member pixels of the clusters (in blue) on the original grayscale image.
The first bipartitioning separates vegetation (cluster 1) from non-vegetation (cluster 2), the bipartitioning of cluster 1 separates grass (cluster 3) from trees (cluster 4), the bipartitioning of cluster 2 separates buildings (cluster 5) from sidewalks/water (cluster 6), and so on.
If the algorithm continues, it chooses to split the leaf node that provides the greatest benefit to the overall tree, which can be quantified as a node's ``score'' in various ways.
The algorithm proceeds to split nodes until a target number of leaves is obtained, until no node is worth splitting, or some other stopping criterion is met.

While the hierarchical clustering method offers advantages in terms of interpretation as well as execution time compared to flat NMF, implementations of the algorithm are limited to single workstations and the dataset must fit in the available memory.
Currently available implementations can utilize multiple cores via MATLAB \cite{KP13} or explicit shared-memory parallelization in the SmallK library \cite{SmallK}.
The goal of this work is to use distributed-memory parallelism to scale the algorithm to large datasets that require the memory of multiple compute nodes and to high processor counts.
While flat NMF algorithms have been scaled to HPC platforms \cite{FKPB15} \GB{add citations beyond our own}, our implementation is the first to our knowledge to scale a hierarchical NMF method to 1000s of cores. 
As discussed in detail in \Cref{sec:paralgs}, we choose to parallelize the computations associated with each node in the tree, which involve a Rank-2 NMF and the computation of the node's score.
We choose a data matrix distribution across processors that avoids any redistribution of the input matrix regardless of the data-dependent structure of the tree's splitting decisions so that the communication required involves only the small factor matrices.
Analysis of the algorithm shows the dependence of execution time on computation and communication costs as well as on $k$, the number of clusters computed.
In particular, we confirm that many of the dominant costs are logarithmic in $k$, which is favorable to the linear or sometimes superlinear dependence of flat NMF algorithms.

We demonstrate in \Cref{sec:results} the efficiency and scalability of our parallel algorithm on three data sets, including the HSI data of the DC mall and an image classification data set involving skin melanoma.
The experimental results show that our parallelization of Rank-2 NMF is highly scalable, maintaining computation bound performance on 1000s of cores.
We also show the limits of strong scalability when scaling to large numbers of clusters (leaf nodes), as the execution time shifts to becoming interprocessor bandwidth bound and eventually latency bound.
The image classification data set requires 800 GB of memory across multiple nodes to process, and in scaling from 10 nodes to 80 nodes of the Summit supercomputer (see \Cref{sec:summit}), we demonstrate parallel speedups of 7.1$\times$ for a single Rank-2 NMF and ZZ$\times$ for a complete hierarchical clustering.

\GB{we should have perhaps compared against SmallK and also against flat NMF...}

\begin{figure}
\input{fig/dc}
\caption{Hierarchical Clustering of DC Mall Hyperspectral Image}
\label{fig:dc}
\end{figure}

### 2020-07-16 latex/tex/prelims.tex
split the paper.tex into multiple files

\section{Preliminaries and Related Work}
\label{sec:prelim}

\subsection{NMF}

The NMF constrained optimization problem
$$\min_{\M{W},\M{H}\geq \M{0}} \|\M{A} - \M{W}\M{H}^\Tra\|_2$$
is nonlinear and nonconvex, and various optimization techniques can be used to approximately solve it.
A popular approach is to use alternating optimization of the two factor matrices because each subproblem is a nonnegative least squares (NNLS) problem, which is convex and can be solved exactly.
Many block coordinate descent (BCD) approaches are possible \cite{KHP14}, and one 2-block BCD algorithm that solves the NNLS subproblems exactly is block principal pivoting \cite{KP11}.
This NNLS algorithm is an active-set-like method that determines the sets of entries in the solution vectors that are zero and those that are positive through an iterative but finite process.

When the rank of the factorization (the number of columns of $\M{W}$ and $\M{H}$) is 2, the NNLS subproblems can be solved much more quickly because the number of possible active sets is only 4.
As explained in more detail in \Cref{sec:r2nmf}, the optimal solution across the 4 sets can be determined efficiently to solve the NNLS subproblem more quickly than general-rank approaches like block principal pivoting.
Because of the relative ease of solving the NMF problem for the rank-2 case, Kuang and Park \cite{KP13} propose a recursive method to use a rank-2 NMF to partition the input data into 2 parts, whereby each part can be further partitioned via rank-2 NMF of the corresponding original data.
This approach yields a hierarchical factorization, potentially uncovering more global structure of the input data and allowing for better scalability of the algorithm to large NMF ranks.

The hierarchical rank-2 NMF method has been applied to document clustering \cite{KP13} and hyperspectral image segmentation \cite{GKP15}.
The leaves of the tree also yield a set of column vectors that can be aggregated into an approximate $\M{W}$ factor (ignoring their hierarchical structure).
Using this factor matrix to initialize a higher-rank NMF computation leads to quick convergence and overall faster performance than initializing NMF with random data; this approach is known as Divide-and-Conquer NMF \cite{DKDP17}.
We focus in this paper on parallelizing the hierarchical algorithms proposed by Kuang and Park \cite{KP13} and Gillis et al. \cite{GKP15}.

\subsection{Parallel NMF}

Scaling algorithms for NMF to large data often requires parallelization in order to fit the data across the memories of multiple compute nodes or speed up the computation to complete in reasonable time.
Parallelizations of multiple optimization approaches have been proposed for general NMF. \GB{add more citations here, including to SmallK}
In particular, we build upon the work of Kannan et al. \cite{KBP16,KBP17,EH+19-TR} and the open-source library PLANC, designed for nonnegative matrix and tensor factorizations of dense and sparse data.
In this parallelization, the alternating optimization approach is employed with various options for the algorithm used to (approximately) solve the NNLS subproblems.
The efficiency of the parallelization is based on scalable algorithms for the parallel matrix multiplications involved in all NNLS algorithms; these algorithms are based on Cartesian distributions of the input matrix across 1D or 2D processor grids.

\subsection{Communication Model}

We use the $\alpha$-$\beta$-$\gamma$ model \cite{TRG05,CH+07,BCDH+14} for analysis of distributed-memory parallel algorithms. 
In this model, the cost of sending a single message of $n$ words of data between two processors is $\alpha + \beta \cdot n$, so that $\alpha$ represents the latency cost of the message and $\beta$ represents the bandwidth cost of each word in the message.
The $\gamma$ parameter represents the computational cost of a single floating point operation (flop).
In this simplified communication model, we ignore contention in the network, assuming in effect a fully connected network, and other limiting factors in practice such as the number of hops between nodes and the network injection rate \cite{GOS16}.
We let $p$ represent the number of processors available on the machine.

All of the interprocessor communication in the algorithms presented in this work are encapsulated in collective communication operations that involve the full set of processors.
Algorithms for implementing the collective operations are built out of pairwise send and receive operations, and we assume the most efficient algorithms are used in our analysis \cite{TRG05,CH+07}.
The collectives used in our algorithms are all-reduce, all-gather, and reduce-scatter.
In an all-reduce, all processors start out with the same amount of data and all end with a copy of the same result, which is in our case a sum of all the inputs (and the same size as a single input).
The cost of an all-reduce of size $n$ words is $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n)$ for $n>p$ and $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n\log p)$ for $n<p$.
In an all-gather, all processors start out with separate data and all end with a copy of the same result, which is the union of all the input data.
If each processor starts with $n/p$ data and ends with $n$ data, the cost of the all-gather is $\alpha \cdot O(\log p) + \beta \cdot O(n)$.
In a reduce-scatter, all processors start out with the same amount of data and all end with a subset of the result, which is in our case a sum of all the inputs (and is smaller than its input).
If each processor starts with $n$ data and ends with $n/p$ data, the cost of the reduce-scatter is $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n)$.
In the case of all-reduce and reduce-scatter, the computational cost is typically dominated by the bandwidth cost because $\beta \gg \gamma$.

### 2020-07-16 latex/tex/introduction.tex
added citations to parallel NMF

@@
+% !TEX root = ../paper.tex
+
 \section{Introduction}
@@
 The goal of this work is to use distributed-memory parallelism to scale the algorithm to large datasets that require the memory of multiple compute nodes and to high processor counts.
-While flat NMF algorithms have been scaled to HPC platforms \cite{FKPB15} \GB{add citations beyond our own}, our implementation is the first to our knowledge to scale a hierarchical NMF method to 1000s of cores.
+While flat NMF algorithms have been scaled to HPC platforms \cite{FKPB15,BW09,MESPS20,KBP17}, our implementation is the first to our knowledge to scale a hierarchical NMF method to 1000s of cores. 
 As discussed in detail in \Cref{sec:paralgs}, we choose to parallelize the computations associated with each node in the tree, which involve a Rank-2 NMF and the computation of the node's score.

### 2020-07-16 latex/tex/introduction.tex
ramki completed introduction

--- 2020-07-16 latex/tex/introduction.tex
@@
 Currently available implementations can utilize multiple cores via MATLAB \cite{KP13} or explicit shared-memory parallelization in the SmallK library \cite{SmallK}.
-The goal of this work is to use distributed-memory parallelism to scale the algorithm to large datasets that require the memory of multiple compute nodes and to high processor counts.
+{\em The goal of this work is to use distributed-memory parallelism to scale the algorithm to large datasets that require the memory of multiple compute nodes and to high processor counts.}
 While flat NMF algorithms have been scaled to HPC platforms \cite{FKPB15} \GB{add citations beyond our own}, our implementation is the first to our knowledge to scale a hierarchical NMF method to 1000s of cores.
+
 As discussed in detail in \Cref{sec:paralgs}, we choose to parallelize the computations associated with each node in the tree, which involve a Rank-2 NMF and the computation of the node's score.

### 2020-07-16 latex/tex/experiments.tex
added texshop hooks and references

@@
+% !TEX root = ../paper.tex
+
 \section{Experimental Results}
@@
 	\item \textit{Synthetic Dataset}:
-	Our synthetic dataset is the same shape of \image{} but consists of fewer rows and columns by a factor of $3$. The resulting matrix is $1048576 \times 11042$. 
-	We chose a smaller size dataset to better see the limits of scaling and to scale from fewer Summit nodes.
+	Our synthetic dataset has the same aspect ratio of \image{} but consists of fewer rows and columns by a factor of $3$. The resulting matrix is $1048576 \times 11042$. 
+	We chose a smaller size dataset to fit on a single node.
 \end{itemize}

### 2020-07-16 latex/tex/prelims.tex
added texshop hooks and references

@@
+% !TEX root = ../paper.tex
+
 \section{Preliminaries and Related Work}
@@
 Scaling algorithms for NMF to large data often requires parallelization in order to fit the data across the memories of multiple compute nodes or speed up the computation to complete in reasonable time.
-Parallelizations of multiple optimization approaches have been proposed for general NMF. \GB{add more citations here, including to SmallK}
+Parallelizations of multiple optimization approaches have been proposed for general NMF \cite{FKPB15,BW09,MESPS20,KBP17,SmallK}. 
 In particular, we build upon the work of Kannan et al.

### 2020-07-16 latex/tex/introduction.tex

--- 2020-07-16 latex/tex/introduction.tex
@@
 Currently available implementations can utilize multiple cores via MATLAB \cite{KP13} or explicit shared-memory parallelization in the SmallK library \cite{SmallK}.
-The goal of this work is to use distributed-memory parallelism to scale the algorithm to large datasets that require the memory of multiple compute nodes and to high processor counts.
+
+{\em The goal of this work is to use distributed-memory parallelism to scale the algorithm to large datasets that require the memory of multiple compute nodes and to high processor counts.}
 While flat NMF algorithms have been scaled to HPC platforms \cite{FKPB15,BW09,MESPS20,KBP17}, our implementation is the first to our knowledge to scale a hierarchical NMF method to 1000s of cores. 

### 2020-07-16 latex/tex/experiments.tex
removed dc mall tree from exp section
single node scaling

@@
 On Summit, we linked this version of Armadillo with OpenBLAS (version 0.3.9) and IBM's Spectrum MPI (version 10.3.1.2-20200121).
-
-\GB{specify that $\imath=\jmath=100$ in all experiments, also what do we set $k$ to be?
-fixed number for each dataset?
-any other experimental parameters we should specify?
-could also specify these at the beginning of \Cref{sec:perf}}
 
@@
 \GB{add only speedup plot for entire tree (at least all complete levels) from 1 core to 42 cores}
+
+\hyper{} is small compared to the other datasets, so it can easily fit on one compute node.
+Also, its small number of 191 rows doesn't allow for parallelizing beyond that number of MPI processes.
+So, this dataset was used for a single-node scaling experiment on Summit from 1 to 42 cores.
+Because the hierarchical clustering algorithm is memory bandwidth bound, there should be limited speedup on one node due to it being essentially shared memory 
+parallelism. Figure \LM{FIXME: DC REF} shows that there is enough speedup for it to be worth parallelizing such a small problem. In this experiment, the
+processes were distributed across both sockets so that a minimum amount of shared memory between processes was used.
 
@@
 \end{figure}
-
-\subsection{DC Clustering Results}
-
-\subsubsection{Hyperspectral Imaging}
-\begin{figure}
-\input{fig/dc}
-\caption{Hierarchical Cluster from DC Mall Image}
-\label{fig:dc}
-\end{figure}

### 2020-07-16 latex/tex/experiments.tex
performance preamble

@@
 \label{sec:perf}
+For all hierarchical clustering experiments in this section, the number of tree leaf nodes $k$ was set at $100$, the tolerance
+for the power iteration was set at $0$, and only complete levels were considered for analysis purposes for both level and strong scaling plots.
 
@@
 
-\Cref{fig:synrank2speedup,fig:rwrank2speedup} show performance relative to the smallest number of nodes required to store data and factor matrices.
-The synthetic data is chosen to fit on a single node, while the \image{} data requires 10 nodes to store the input matrix as well as temporary and output matrices.
-For these data sets, we observe nearly perfect strong scaling, with 47$\times$ speedup on 50 nodes (over 1 node) for synthetic data and 7.1$\times$ speedup on 80 nodes (over 10 nodes) for \image{} data.
+\Cref{fig:synrank2speedup,fig:rwrank2speedup} show performance relative to the smallest number of compute nodes required to store data and factor matrices.
+The synthetic data is chosen to fit on a single node, while the \image{} data requires 10 compute nodes to store the input matrix as well as temporary and output matrices.
+For these data sets, we observe nearly perfect strong scaling, with 47$\times$ speedup on 50 compute nodes (over 1 compute node) for synthetic data and 7.1$\times$ speedup on 80 compute nodes (over 10 compute nodes) for \image{} data.
 

### 2020-07-16 latex/tex/experiments.tex
hierarchical scaling

@@
 
-From equation \ref{eq:treecost}, we expect to see perfect strong scaling in a computationally bound H2NMF problem with fixed frontier node $k$ count.
-However, the problem should become latency bound with its $\log p$ cost as we scale. Figure \ref{fig:fig:synhierspeedup} shows this latency limit as
-we scale to 50 compute nodes on the synthetic data. Figure \ref{fig:synhierstrongscaling}
+From equation \ref{eq:treecost}, we expect to see perfect strong scaling in a computationally bound hierarchical clustering problem with fixed frontier node $k=100$ count.
+The problem should become latency bound with its $\log p$ cost as we scale.
 
-From equation \ref{eq:levelcost}
+Figure \ref{fig:synhierstrongscaling} shows the limits of scaling as the hierarchical cost is becoming saturated with communication.
+The AllReduce segment would consistently grow if the problem was becoming latency bound, which is not the case.
+So, the problem is becoming bandwidth bound as seen through the ReduceScatter and AllGather segments. 
+However, computation still makes up more than 60\% of the total cost.
+
+With the much larger \image{} dataset, it's possible to scale much further as seen in Figure \ref{fig:rwhierspeedup} without being overwhelmed by the communication cost.
+From Figure \ref{fig:fig:rwhierstrongscaling}, the communication cost doesn't make up more than 20\% of the total cost even scaling to 80 compute nodes.
+
+Note that in order to compare hierarchical clusters between differing compute nodes, we limited the output to include only complete levels in the tree.
+
 \begin{itemize}

### 2020-07-16 latex/tex/conclusion.tex
moved comment from intro to conclusion
drafted conclusion

@@
+% !TEX root = ../paper.tex
+
 \section{Conclusion}
 
-\begin{itemize}
-	\item consider sparse case
-	\item consider parallelizing across tree
-	\item consider redistribution for 2D
-\end{itemize}
+As shown in the theoretical analysis (\Cref{sec:analysis}) and experimental results (\Cref{sec:hiernmf2scaling}), \Cref{alg:hiernmf2} can efficiently scale to large $p$ as long as the execution time is dominated by local matrix multiplication.
+The principal barriers to scalability are the bandwidth cost due to Rank-2 NMF, which is consistent across levels of the tree and proportional to the number of columns $n$ of the original data set, and the latency cost due to large numbers of tree nodes in lower levels of the tree.
+When $n$ is small relative to $m$ and the number of leaves $k$ and levels $\ell$ are small, then these barriers do not pose a problem until $p$ is very large.
+However, if the input matrix short and fat (i.e., has many samples with few features), then the bandwidth cost can hinder performance for smaller $p$.
+Likewise, if $k$ is large or the tree is lopsided, then achieving scalability for very small problems is more difficult.
+We also note that in the case of sparse $\M{A}$, it becomes more difficult to hide communication behind the cheaper matrix multiplications, and other costs may become more dominant.
+
+One approach for reducing the bandwidth cost of Rank-2 NMF is to choose a more balanced data distribution over a 2D grid, as proposed by Kannan et al. \cite{KBP16}.
+This will reduce the communicated data and achieve a local data matrix that is more square, which can improve local matrix multiplication performance.
+The downside of this approach is that it requires a redistribution of the data for each split, but if many NMF iterations are required, then the single upfront redistribution cost may be amortized.
+
+Another approach to alleviate the rising latency costs of lower levels of the tree is to parallelize across nodes of the tree.
+This will result in fewer processors working on any given node, reducing the synchronization time among them, and it will allow small, latency-bound problems to be solved simultaneously.
+Prioritizing the sequence of node splits is more difficult in this case, but modifying the stopping criterion for splitting to use a score threshold instead of a target number of leaves will allow truly independent computation.
+
+In the future, we also plan to compare performance of \Cref{alg:hiernmf2} with flat NMF algorithms and employ the Divide-and-Conquer NMF technique \cite{DKDP17} of seeding an iterative flat NMF algorithm with the feature vectors of the leaf nodes.
+The parallel technique proposed here can be combined with the existing PLANC library \cite{EH+19-TR} to obtain faster overall convergence for very large datasets.

### 2020-07-16 latex/tex/experiments.tex
drafted conclusion

--- 2020-07-16 latex/tex/experiments.tex
@@
 On Summit, we linked this version of Armadillo with OpenBLAS (version 0.3.9) and IBM's Spectrum MPI (version 10.3.1.2-20200121).
+
+\GB{specify that $\imath=\jmath=100$ in all experiments, also what do we set $k$ to be?
+fixed number for each dataset?
+any other experimental parameters we should specify?
+could also specify these at the beginning of \Cref{sec:perf}}
 
@@
 \GB{add only speedup plot for entire tree (at least all complete levels) from 1 core to 42 cores}
-
-\hyper{} is small compared to the other datasets, so it can easily fit on one compute node.
-Also, its small number of 191 rows doesn't allow for parallelizing beyond that number of MPI processes.
-So, this dataset was used for a single-node scaling experiment on Summit from 1 to 42 cores.
-Because the hierarchical clustering algorithm is memory bandwidth bound, there should be limited speedup on one node due to it being essentially shared memory 
-parallelism. Figure \LM{FIXME: DC REF} shows that there is enough speedup for it to be worth parallelizing such a small problem. In this experiment, the
-processes were distributed across both sockets so that a minimum amount of shared memory between processes was used.
 
@@
 \subsubsection{Hierarchical Clustering Strong Scaling}
+\label{sec:hiernmf2scaling}
 

### 2020-07-16 latex/tex/experiments.tex
level scaling start

--- 2020-07-16 latex/tex/experiments.tex
@@
 
+To show the level scaling in a particular tree, we only considered complete levels as the analysis from Equation \ref{eq:levelcost} only considers
+complete levels. From Equation \ref{eq:levelcost}, the high order computational term (represented by MatMul) is constant per level, the lower order 
+computational term (represented by NNLS) grows like $O(2^\ell)$, and the latency cost grows similarly like $O(2^\ell)$.
+
+From Figure \ref{fig:seqlevelbreakdown}, the MatMul cost decreases slightly per level, which
+
 \begin{itemize}

### 2020-07-16 latex/tex/analysis.tex
small update to analysis

@@
+% !TEX root = ../paper.tex
+
 \subsubsection{Analysis}
@@
 There are four communication collectives each iteration, and each involves all $p$ processors.
-The two all-reduce collectives to compute the Gram matrices of the factor matrices involve $2\times 2$ matrices and incur a communication cost of $(\beta + \alpha) \cdot O(\log p)$.
+The two all-reduce collectives to compute the Gram matrices of the factor matrices involve $2\times 2$ matrices and incur a communication cost of $(\gamma+\beta + \alpha) \cdot O(\log p)$.
 The reduce-scatter and all-gather collectives involve $n\times 2$ matrices (the size of $\M{H}$) and require $\beta \cdot O(n) + \alpha \cdot O(\log p)$ in communication cost (we ignore the computation cost of the reduce-scatter because it is typically dominated by the bandwidth cost).

### 2020-07-16 latex/tex/algorithm.tex
completed algorithm and prelims

@@
 
-In this section, we consider the options for parallelizing Hierarchical Rank-2 NMF Clustering (\Cref{alg:hiernmf2}) and provide an analysis for the approach we take.
+In this section, we consider the options for parallelizing Hierarchical Rank-2 NMF Clustering (\Cref{alg:hiernmf2}) and provide an analysis for our approach.
 The running time of an algorithm is data dependent because not only does each Rank-2 NMF computation require a variable number of iterations, but also the shape of the tree can vary from a balanced binary tree with $O(\log k)$ levels to a tall, flat tree with $O(k)$ levels.

### 2020-07-16 latex/tex/introduction.tex

@@
 We also show the limits of strong scalability when scaling to large numbers of clusters (leaf nodes), as the execution time shifts to becoming interprocessor bandwidth bound and eventually latency bound.
-The image classification data set requires 800 GB of memory across multiple nodes to process, and in scaling from 10 nodes to 80 nodes of the Summit supercomputer (see \Cref{sec:summit}), we demonstrate parallel speedups of 7.1$\times$ for a single Rank-2 NMF and ZZ$\times$ for a complete hierarchical clustering.
-
-\GB{we should have perhaps compared against SmallK and also against flat NMF...}
+The image classification data set requires 800 GB of memory across multiple nodes to process, and in scaling from 10 nodes to 80 nodes of the Summit supercomputer (see \Cref{sec:summit}), we demonstrate parallel speedups of 7.1$\times$ for a single Rank-2 NMF and 5.9$\times$ for a complete hierarchical clustering.
 

### 2020-07-16 latex/tex/experiments.tex
edited section 4.1

@@
 
-H2NMF uses the Armadillo library (version 9.900.1) for all matrix operations. 
-In Armadillo, sparse matrices are stored in Compressed Sparse Column (CSC) format and dense matrices are stored in column-major ordering.
+Our implementation builds on the PLANC open-source library and uses the Armadillo library (version 9.900.1) for all matrix operations. 
+%In Armadillo, sparse matrices are stored in Compressed Sparse Column (CSC) format and dense matrices are stored in column-major ordering.
 On Summit, we linked this version of Armadillo with OpenBLAS (version 0.3.9) and IBM's Spectrum MPI (version 10.3.1.2-20200121).

### 2020-07-16 latex/tex/experiments.tex
levels section

--- 2020-07-16 latex/tex/experiments.tex
@@
 
-From Figure \ref{fig:seqlevelbreakdown}, the MatMul cost decreases slightly per level, which
+From Figure \ref{fig:seqlevelbreakdown}, the MatMul cost decreases slightly per level, which may be explained by BLAS optimizations occurring in the
+local matrix multiply, especially as each node's subproblem becomes smaller and smaller. The NNLS grows exponentially, which is exactly in line with theory,
+and none of the levels seem to become bandwidth or latency bound for this case.
+
+As we increase the nodes to 40 in the same sequential case, the story completely changes.
+Figure \ref{fig:parallellevelbreakdown} shows the MatMul cost is almost precisely the same, which is line with the theory.
+The NNLS cost grows exponentially as before and becomes dominating at the lower levels. The latency is also appearing as the AllReduce cost, which is latency bound,
+starts to grow exponentially from level 3 onwards.
+
+In Figure \ref{fig:rwparallellevelbreakdown}, note that this levels experiment is from the image dataset, which is considerably larger than the synthetic case but is of the same size.
+Although all is relatively the same as in Figure \ref{fig:parallellevelbreakdown}, matrix multiply still dominates at the lower levels and the problem is computationally bound
+at each level, which is indiciative of the strong speedup as seen in the previous section.
 

### 2020-07-16 latex/tex/experiments.tex
edited dataset descriptions

--- 2020-07-16 latex/tex/experiments.tex
@@
 \subsection{Datasets}
-\begin{itemize}
-	\item \textit{Hyperspectral Imaging}:
-	For Hyperspectral imaging, we chose the Hyperspectral digital imagery collection experiment (HYDICE) image of the Washington DC Mall. We will refer
+
+\paragraph{\emph{Hyperspectral Imaging}}
+
+	We use the Hyperspectral Digital Imagery Collection Experiment (HYDICE) image of the Washington DC Mall. We will refer
 	to this dataset as \hyper{}\cite{DC-HYDICE}.
@@
 	\hyper{} is formatted into a 3-way tensor representing two spatial dimensions of pixels and one dimension of spectral bands. So, a slice along
-	the spectral band dimnesion would be the full \hyper{} image in that spectral band. For H2NMF, these tensors are flattened so that the rows represent the
-	$191$ spectral bands and the columns represent the $392960$ pixels.
+	the spectral band dimension would be the full \hyper{} image in that spectral band. For hierarchical clustering, these tensors are flattened so that the rows represent the
+	$191$ spectral bands and the columns represent the $392960$ pixels.   The data set is approximately 600 MB in size.
 
-	Larger hyperspectral datasets tend to not grow larger than around 200 spectral bands and have millions or billions of pixels. So, scaling is 
-	limited for this dataset given our data distribution scheme.
-	\item \textit{Image Classification}:
-	Image Classification datasets contain tens of thousands of images of equal size. The dataset we chose is the SIIM-ISIC Melanoma classification
-	dataset, which we will refer to as \image{}\cite{SIIM-ISIC}. \image{} consists of $33126$ RGB training images equally sized at $1024 \times 1024$. Unlike with hyperspectral imaging, the resulting
-	matrix used in H2NMF consists of image pixels along the rows and individual images along the columns. So, the resulting sized matrix is
-	$3145728 \times 33126$, which is approximately 800GB in size.
+	%Larger hyperspectral datasets tend to not grow larger than around 200 spectral bands and have millions or billions of pixels. So, scaling is 
+	%limited for this dataset given our data distribution scheme.
 
-	Given its size, \image{} can only scale from $10$ Summit nodes.
-	\item \textit{Synthetic Dataset}:
+\paragraph{\emph{Image Classification}}
+
+	The SIIM-ISIC Melanoma classification
+	dataset, which we will refer to as \image{}\cite{SIIM-ISIC}, consists of $33126$ RGB training images equally sized at $1024 \times 1024$. Unlike with hyperspectral imaging, the resulting
+	matrix used in hierarchical clustering consists of image pixels along the rows and individual images along the columns. So, the resulting sized matrix is
+	$3145728 \times 33126$, which is approximately 800 GB in size.
+
+	Given its size, \image{} requries $10$ Summit nodes to perform hierarchical clustering.
+
+\paragraph{\emph{Synthetic Dataset}}
+
 	Our synthetic dataset has the same aspect ratio of \image{} but consists of fewer rows and columns by a factor of $3$. The resulting matrix is $1048576 \times 11042$. 
-	We chose a smaller size dataset to fit on a single node.
-\end{itemize}
+	We choose the smaller size in order to fit on a single node for scaling experiments.
 

### 2020-07-16 latex/tex/experiments.tex
performance preamble updates

--- 2020-07-16 latex/tex/experiments.tex
@@
 \label{sec:perf}
-For all hierarchical clustering experiments in this section, the number of tree leaf nodes $k$ was set at $100$, the tolerance
-for the power iteration was set at $0$, and only complete levels were considered for analysis purposes for both level and strong scaling plots.
+For all hierarchical clustering experiments in this section, the number of tree leaf nodes $k$ was set at $100$, 
+the number of NMF iterations was set to $100$,
+the power iteration was allowed to stop iterating after convergence,
+and only complete levels were considered for analysis purposes for both level and strong scaling plots.
 

### 2020-07-16 latex/tex/conclusion.tex

@@
 When $n$ is small relative to $m$ and the number of leaves $k$ and levels $\ell$ are small, then these barriers do not pose a problem until $p$ is very large.
-However, if the input matrix short and fat (i.e., has many samples with few features), then the bandwidth cost can hinder performance for smaller $p$.
+However, if the input matrix is short and fat (i.e., has many samples with few features), then the bandwidth cost can hinder performance for smaller $p$.
 Likewise, if $k$ is large or the tree is lopsided, then achieving scalability for very small problems is more difficult.

### 2020-07-16 latex/tex/experiments.tex
editing DC perf section

--- 2020-07-16 latex/tex/experiments.tex
@@
 
-\GB{add only speedup plot for entire tree (at least all complete levels) from 1 core to 42 cores}
-
 \hyper{} is small compared to the other datasets, so it can easily fit on one compute node.
@@
 So, this dataset was used for a single-node scaling experiment on Summit from 1 to 42 cores.
-Because the hierarchical clustering algorithm is memory bandwidth bound, there should be limited speedup on one node due to it being essentially shared memory 
-parallelism. Figure \LM{FIXME: DC REF} shows that there is enough speedup for it to be worth parallelizing such a small problem. In this experiment, the
-processes were distributed across both sockets so that a minimum amount of shared memory between processes was used.
+Because Rank-2 NMF is memory bandwidth bound, we expect limited speedup on one node due to the memory bandwidth not scaling linearly with the number of cores. 
+\Cref{fig:dcspeedup} shows that there is enough speedup (14$\times$ on 42 cores) for it to be worth parallelizing such a small problem, but perfect scaling requires more memory bandwidth. 
+In this experiment, the processes were distributed across both sockets so that an even number of cores on each socket are used.
 
-\begin{itemize}
-	\item emphasize we start with a small data set that fits on one node and has only 191 rows
-	\item using 1 node (2 sockets), because we are memory bandwidth bound we expect limited speedup but we can save time with parallelism even for this small a data set
-\end{itemize}
+\begin{figure}
+\begin{center}
+\includegraphics[height=2in, width=\columnwidth]{plots/dc_hierarchical_speedup.pdf}
+\caption{Strong Scaling Speedup for Hierarchical Clustering on \hyper{}}
+\label{fig:dcspeedup}
+\end{center}
+\end{figure}
 

### 2020-07-16 latex/tex/experiments.tex

@@
 \Cref{fig:synrank2speedup,fig:rwrank2speedup} show performance relative to the smallest number of compute nodes required to store data and factor matrices.
-The synthetic data is chosen to fit on a single node, while the \image{} data requires 10 compute nodes to store the input matrix as well as temporary and output matrices.
-For these data sets, we observe nearly perfect strong scaling, with 47$\times$ speedup on 50 compute nodes (over 1 compute node) for synthetic data and 7.1$\times$ speedup on 80 compute nodes (over 10 compute nodes) for \image{} data.
+For these data sets, we observe nearly perfect strong scaling, with 42$\times$ speedup on 40 compute nodes (over 1 compute node) for synthetic data and 7.1$\times$ speedup on 80 compute nodes (over 10 compute nodes) for \image{} data.
 
@@
 
-\GB{Do these breakdowns include the score function?  actually this may explain why all-reduce costs more, it also has $O(n)$ data for score...}
-
-\GB{4 plots: 2 data sets with speedup/breakdown, note that we include time for complete levels, ignoring nodes in incomplete levels for easier interpretation}
-
-From equation \ref{eq:treecost}, we expect to see perfect strong scaling in a computationally bound hierarchical clustering problem with fixed frontier node $k=100$ count.
-The problem should become latency bound with its $\log p$ cost as we scale.
+From \Cref{eq:treecost}, we expect to see perfect strong scaling in a computationally bound hierarchical clustering problem with target cluster count $k=100$.
+As $k$ is large, we expect the latency cost of small problems deep in the tree to limit scalability.
 
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hierarchical_speedup.pdf}
-\caption{Strong Scaling Speedup for Hierarchical Clustering. The total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
+\caption{Strong Scaling Speedup for Hierarchical Clustering on Synthetic data}
 \label{fig:synhierspeedup}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hier_strongscaling.pdf}
-\caption{Relative Time Breakdown for Hierarchical Clustering. The plot uses the total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
+\caption{Relative Time Breakdown for Hierarchical Clustering on Synthetic data}
 \label{fig:synhierstrongscaling}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_hierarchical_speedup.pdf}
-\caption{Strong Scaling Speedup for Hierarchical Clustering on \image{} Data. The total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
+\caption{Strong Scaling Speedup for Hierarchical Clustering on \image{} Data}
 \label{fig:rwhierspeedup}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_hier_strongscaling.pdf}
-\caption{Relative Time Breakdown for Hierarchical Clustering on \image{}. The plot uses the total time taken for complete Hierarchical Clustering \cref{alg:hiernmf2}}
+\caption{Relative Time Breakdown for Hierarchical Clustering on \image{} Data}
 \label{fig:rwhierstrongscaling}

### 2020-07-16 latex/tex/experiments.tex
figure captions consistency

@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_speedup.pdf}
-\caption{Strong Scaling Speedup for Rank-2 NMF on synthetic data}
+\caption{Strong Scaling Speedup for Rank-2 NMF on Synthetic Data}
 \label{fig:synrank2speedup}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_strongscaling.pdf}
-\caption{Relative Time Breakdown for Rank-2 NMF on synthetic data}
+\caption{Relative Time Breakdown for Rank-2 NMF on Synthetic Data}
 \label{fig:synrank2strongscaling}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_speedup.pdf}
-\caption{Strong Scaling Speedup for Rank-2 NMF on \image{} data}
+\caption{Strong Scaling Speedup for Rank-2 NMF on \image{} Data}
 \label{fig:rwrank2speedup}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_strongscaling.pdf}
-\caption{Relative Time Breakdown for Rank-2 NMF on \image{} data}
+\caption{Relative Time Breakdown for Rank-2 NMF on \image{} Data}
 \label{fig:rwrank2strongscaling}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hierarchical_speedup.pdf}
-\caption{Strong Scaling Speedup for Hierarchical Clustering on Synthetic data}
+\caption{Strong Scaling Speedup for Hierarchical Clustering on Synthetic Data}
 \label{fig:synhierspeedup}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hier_strongscaling.pdf}
-\caption{Relative Time Breakdown for Hierarchical Clustering on Synthetic data}
+\caption{Relative Time Breakdown for Hierarchical Clustering on Synthetic Data}
 \label{fig:synhierstrongscaling}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_sequential_level_breakdown.pdf}
-\caption{Sequential Breakdown plot for levels 0 to 5. y-axis is absolute wall clock time in seconds}
+\caption{Levels Breakdown Timings for 1 Compute Node on Synthetic Data}
 \label{fig:seqlevelbreakdown}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_parallel_level_breakdown.pdf}
-\caption{Parallel Breakdown plot for levels 0 to 5 on 40 nodes. y-axis is absolute wall clock time in seconds}
+\caption{Levels Breakdown Timings for 40 Compute Nodes on Synthetic Data}
 \label{fig:parallellevelbreakdown}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_parallel_level_breakdown.pdf}
-\caption{Parallel Breakdown plot for levels 0 to 5 on 80 nodes. y-axis is absolute wall clock time in seconds}
+\caption{Levels Breakdown Timings for 80 Compute Nodes on \image{} Data}
 \label{fig:rwparallellevelbreakdown}

### 2020-07-16 latex/tex/experiments.tex
edited hier perf section

--- 2020-07-16 latex/tex/experiments.tex
@@
 
-Figure \ref{fig:synhierstrongscaling} shows the limits of scaling as the hierarchical cost is becoming saturated with communication.
-The AllReduce segment would consistently grow if the problem was becoming latency bound, which is not the case.
-So, the problem is becoming bandwidth bound as seen through the ReduceScatter and AllGather segments. 
-However, computation still makes up more than 60\% of the total cost.
+\Cref{fig:synhierspeedup} demonstrates the scalability of the synthetic data set on up to 40 nodes, and we observe a 15$\times$ speedup compared to 1 node.
+\Cref{fig:synhierstrongscaling} shows the relative time breakdown and explains the limitation on scaling.
+On 40 nodes, computation still takes 60\% of the total time, but the all-gather and reduce-scatter costs have grown in relative time because they do not scale with $p$.
+Because all-reduce involves only a constant amount of data and its time remains relatively small, we conclude the communication is bandwidth bound at this scale.
 
-With the much larger \image{} dataset, it's possible to scale much further as seen in Figure \ref{fig:rwhierspeedup} without being overwhelmed by the communication cost.
-From Figure \ref{fig:fig:rwhierstrongscaling}, the communication cost doesn't make up more than 20\% of the total cost even scaling to 80 compute nodes.
+With the larger \image{} dataset, it's possible to scale much further as seen in \Cref{fig:rwhierspeedup}, where we observe a 5.9$\times$ speedup of 80 compute nodes compared to 10.
+From \Cref{fig:fig:rwhierstrongscaling}, we see that the communication cost constitutes less than 20\% of the total time even at 80 compute nodes.
 
-Note that in order to compare hierarchical clusters between differing compute nodes, we limited the output to include only complete levels in the tree.
+We note that the speedup of the overall hierarchical clustering algorithm is not as high as for a single Rank-2 NMF (measured at the root node).
+This is due to inefficiencies in the lower levels of the tree, as we explore in the next section.
 
-\begin{itemize}
-	\item theory (\Cref{eq:treecost}) says...
-	\item Fig 7 shows limit of strong scaling for synthetic data (time increases at 50 nodes compared to 40), Fig 8 gives breakdown and explains
-	\begin{itemize}
-		\item drastic performance dropoff may be artifact of all-gather, which should take about the same time as reduce-scatter, but speedup from 30 to 40 nodes is also slight
-		\item by 50 nodes, about 50\% of time in communication; still mostly dominated by bandwidth cost because ag and rs dominate ar
-		\item note the difference between the scaling of single rank-2 and entire tree: poor scaling must be due to lower levels, which we explore in the next section
-	\end{itemize}
-	\item Fig 9 shows reasonable scaling for entire tree -- speedup drops from 7x to 6x over 80 nodes; Fig 10 shows that MM is still 70\% of run time on 80 nodes, which allows for near-linear scaling
-\end{itemize}
+%Note that in order to compare hierarchical clusters between differing compute nodes, we limited the output to include only complete levels in the tree.
 

### 2020-07-16 latex/paper.tex
tightening bib
edited level scaling section

@@
 \title{Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization
-\thanks{This material is based upon work supported by the National Science Foundation under Grant No. OAC-1642385 and OAC-1642410.
-This manuscript has been co-authored by UT-Battelle, LLC under Contract No. DE-AC05-00OR22725 with the U.S. Department of
-Energy. This project was partially funded by the Laboratory Director's Research and Development fund. This research used resources
-of the Oak Ridge Leadership Computing Facility at the Oak Ridge National Laboratory, which is supported by the Office of Science of
-the U.S. Department of Energy.}
+%Add this back in for camera ready
+%\thanks{This material is based upon work supported by the National Science Foundation under Grant No. OAC-1642385 and OAC-1642410.
+%This manuscript has been co-authored by UT-Battelle, LLC under Contract No. DE-AC05-00OR22725 with the U.S. Department of
+%Energy. This project was partially funded by the Laboratory Director's Research and Development fund. This research used resources
+%of the Oak Ridge Leadership Computing Facility at the Oak Ridge National Laboratory, which is supported by the Office of Science of
+%the U.S. Department of Energy.}
 }
@@
 
-\input{tex/ack.tex}
+%Add this back in for camera ready
+%\input{tex/ack.tex}
 
+\bibliographystyle{plainurl}
 \bibliography{paper}
-\bibliographystyle{plainurl}
 

### 2020-07-16 latex/tex/experiments.tex
tightening bib
edited level scaling section

@@
 All the experiments in this section were conducted on Summit. Summit is a supercomputer created by IBM for the Oak Ridge National Laboratory. 
-There are approximately 4,600 nodes on Summit. Each node contains two IBM POWER9 processors on separate sockets with dual NVLINK bricks to facilitate a transfer rate of 
-25 GB/s between the processors. Each node contains 512 GB of DDR4 memory for use by both processors. Each POWER9 processor utilizes 22 IBM SIMD Multi-Cores (SMCs), 
-although one of these SMCs on each processor is dedicated to memory transfer and is therefore not available for computation. 
+There are approximately 4,600 nodes on Summit. 
+Each node contains two IBM POWER9 processors on separate sockets with 512 GB of DDR4 memory.
+Each POWER9 processor utilizes 22 IBM SIMD Multi-Cores (SMCs), although one of these SMCs on each processor is dedicated to memory transfer and is therefore not available for computation. 
 For node scaling experiments, all 42 available SMCs were utilized in each node so that every node computed with 42 separate MPI processes.
@@
 
-Our implementation builds on the PLANC open-source library and uses the Armadillo library (version 9.900.1) for all matrix operations. 
-%In Armadillo, sparse matrices are stored in Compressed Sparse Column (CSC) format and dense matrices are stored in column-major ordering.
+Our implementation builds on the PLANC open-source library \cite{EH+19-TR} and uses the Armadillo library (version 9.900.1) for all matrix operations. 
 On Summit, we linked this version of Armadillo with OpenBLAS (version 0.3.9) and IBM's Spectrum MPI (version 10.3.1.2-20200121).
@@
 With the larger \image{} dataset, it's possible to scale much further as seen in \Cref{fig:rwhierspeedup}, where we observe a 5.9$\times$ speedup of 80 compute nodes compared to 10.
-From \Cref{fig:fig:rwhierstrongscaling}, we see that the communication cost constitutes less than 20\% of the total time even at 80 compute nodes.
+From \Cref{fig:rwhierstrongscaling}, we see that the communication cost constitutes less than 20\% of the total time even at 80 compute nodes.
 
@@
 This is due to inefficiencies in the lower levels of the tree, as we explore in the next section.
-
-%Note that in order to compare hierarchical clusters between differing compute nodes, we limited the output to include only complete levels in the tree.
 
@@
 
-\GB{3 plots: for synthetic level breakdown for 1 node and for 40 nodes; for realworld just for 80 nodes}
-
-To show the level scaling in a particular tree, we only considered complete levels as the analysis from Equation \ref{eq:levelcost} only considers
-complete levels. From Equation \ref{eq:levelcost}, the high order computational term (represented by MatMul) is constant per level, the lower order 
+To compare execution time across levels of a particular tree, we consider only complete levels. 
+From \Cref{eq:levelcost}, the dominant computational term (due to MatMul) is constant per level, the lower order 
 computational term (represented by NNLS) grows like $O(2^\ell)$, and the latency cost grows similarly like $O(2^\ell)$.
 
-From Figure \ref{fig:seqlevelbreakdown}, the MatMul cost decreases slightly per level, which may be explained by BLAS optimizations occurring in the
-local matrix multiply, especially as each node's subproblem becomes smaller and smaller. The NNLS grows exponentially, which is exactly in line with theory,
-and none of the levels seem to become bandwidth or latency bound for this case.
+\Cref{fig:seqlevelbreakdown} show absolute time across levels for the synthetic data set on 1 node.
+The MatMul cost decreases slightly per level, which may be explained by cache effects in the local matrix multiply, as each node's subproblem decreases in size. 
+The NNLS grows exponentially, as expected, and communication is negligible.
 
-As we increase the nodes to 40 in the same sequential case, the story completely changes.
-Figure \ref{fig:parallellevelbreakdown} shows the MatMul cost is almost precisely the same, which is line with the theory.
-The NNLS cost grows exponentially as before and becomes dominating at the lower levels. The latency is also appearing as the AllReduce cost, which is latency bound,
-starts to grow exponentially from level 3 onwards.
+\Cref{fig:parallellevelbreakdown} shows the level breakdown for the synthetic data on 40 nodes, where we see different behavior.
+MatMul cost is again constant across levels and the NNLS cost becomes dominating at lower levels suggesting it does not scale as well as MatMul. 
+We also see all-reduce time becoming significant as communication time increases, indicating that the nodes at lower levels are becoming more latency bound.
+Thus, we see that the poorer scaling at the lower levels of the tree is the main reason the overall hierarchical clustering algorithm does not scale as well as the single Rank-2 NMF at the root node.
 
-In Figure \ref{fig:rwparallellevelbreakdown}, note that this levels experiment is from the image dataset, which is considerably larger than the synthetic case but is of the same size.
-Although all is relatively the same as in Figure \ref{fig:parallellevelbreakdown}, matrix multiply still dominates at the lower levels and the problem is computationally bound
-at each level, which is indiciative of the strong speedup as seen in the previous section.
-
-\begin{itemize}
-	\item theory says per-level cost is \Cref{eq:levelcost}
-\end{itemize}
+%In Figure \ref{fig:rwparallellevelbreakdown}, note that this levels experiment is from the image dataset, which is considerably larger than the synthetic case but is of the same size.
+%Although all is relatively the same as in Figure \ref{fig:parallellevelbreakdown}, matrix multiply still dominates at the lower levels and the problem is computationally bound
+%at each level, which is indiciative of the strong speedup as seen in the previous section.
 
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/realworld_parallel_level_breakdown.pdf}
-\caption{Levels Breakdown Timings for 80 Compute Nodes on \image{} Data}
-\label{fig:rwparallellevelbreakdown}
-\end{center}
-\end{figure}
+%\begin{figure}
+%\begin{center}
+%\includegraphics[height=2in, width=\columnwidth]{plots/realworld_parallel_level_breakdown.pdf}
+%\caption{Levels Breakdown Timings for 80 Compute Nodes on \image{} Data}
+%\label{fig:rwparallellevelbreakdown}
+%\end{center}
+%\end{figure}

### 2020-07-16 latex/tex/introduction.tex

@@
 An NMF with $k=2$ determines a row of $\M{H}$ of length 2 for each sample that determines its part membership.
+\LM{delete previous sentence?}
 By recursively bipartitioning the samples in each part, the method determines a binary tree such that each sample belongs to a unique leaf and internal nodes of the tree determine higher-level clusters of the samples.
+\LM{By recursively bipartioning the samples in each node, the method determines a binary tree such that all leaves contain unique samples and the structure of the tree determines hierarchical clusters}
 A single latent feature vector for each node can also be used for interpreting clusters.
+\LM{A single W vector for each node can also be used for cluster interpretation.}
 Similarly, Du et al.
@@
 We discuss the hierarchical method in more detail in \Cref{sec:prelim} and \Cref{sec:seqalgs}.
+\LM{Get rid of this entirely since it isn't really discussed further in 2 and 3?}
 
@@
 The algorithm proceeds to split nodes until a target number of leaves is obtained, until no node is worth splitting, or some other stopping criterion is met.
+\LM{Should we have text explaining it or could we add these labels to the figure itself?}
 

### 2020-07-16 latex/tex/experiments.tex
tightening figs
removed merge text

@@
 \includegraphics[height=2in, width=\columnwidth]{plots/dc_hierarchical_speedup.pdf}
-\caption{Strong Scaling Speedup for Hierarchical Clustering on \hyper{}}
+\caption{Strong Scaling for Hierarchical Clustering on \hyper{}}
 \label{fig:dcspeedup}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_speedup.pdf}
-\caption{Strong Scaling Speedup for Rank-2 NMF on Synthetic Data}
+\caption{Strong Scaling for Rank-2 NMF on Synthetic}
 \label{fig:synrank2speedup}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_strongscaling.pdf}
-\caption{Relative Time Breakdown for Rank-2 NMF on Synthetic Data}
+\caption{Time Breakdown for Rank-2 NMF on Synthetic}
 \label{fig:synrank2strongscaling}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_speedup.pdf}
-\caption{Strong Scaling Speedup for Rank-2 NMF on \image{} Data}
+\caption{Strong Scaling for Rank-2 NMF on \image{}}
 \label{fig:rwrank2speedup}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_strongscaling.pdf}
-\caption{Relative Time Breakdown for Rank-2 NMF on \image{} Data}
+\caption{Time Breakdown for Rank-2 NMF on \image{}}
 \label{fig:rwrank2strongscaling}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hierarchical_speedup.pdf}
-\caption{Strong Scaling Speedup for Hierarchical Clustering on Synthetic Data}
+\caption{Strong Scaling for Hierarchical Clustering on Synthetic}
 \label{fig:synhierspeedup}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hier_strongscaling.pdf}
-\caption{Relative Time Breakdown for Hierarchical Clustering on Synthetic Data}
+\caption{Time Breakdown for Hierarchical Clustering on Synthetic}
 \label{fig:synhierstrongscaling}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_hierarchical_speedup.pdf}
-\caption{Strong Scaling Speedup for Hierarchical Clustering on \image{} Data}
+\caption{Strong Scaling for Hierarchical Clustering on \image{}}
 \label{fig:rwhierspeedup}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_hier_strongscaling.pdf}
-\caption{Relative Time Breakdown for Hierarchical Clustering on \image{} Data}
+\caption{Time Breakdown for Hierarchical Clustering on \image{}}
 \label{fig:rwhierstrongscaling}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_sequential_level_breakdown.pdf}
-\caption{Levels Breakdown Timings for 1 Compute Node on Synthetic Data}
+\caption{Levels Breakdown Timings for 1 Compute Node on Synthetic}
 \label{fig:seqlevelbreakdown}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_parallel_level_breakdown.pdf}
-\caption{Levels Breakdown Timings for 40 Compute Nodes on Synthetic Data}
+\caption{Levels Breakdown Timings for 40 Compute Nodes on Synthetic}
 \label{fig:parallellevelbreakdown}
@@
 %\includegraphics[height=2in, width=\columnwidth]{plots/realworld_parallel_level_breakdown.pdf}
-%\caption{Levels Breakdown Timings for 80 Compute Nodes on \image{} Data}
+%\caption{Levels Breakdown Timings for 80 Compute Nodes on \image{}}
 %\label{fig:rwparallellevelbreakdown}

### 2020-07-16 latex/tex/introduction.tex
removed unnecessary sentences

--- 2020-07-16 latex/tex/introduction.tex
@@
 
-Hierarchical clustering is the process of recursively breaking a group into two similar groups until every sample is a cluster by itself.
-That is., we consider all the data points as a single cluster and in each iteration, the data points in the current cluster are split into two separate clusters till there is only one sample in a cluster.
+Hierarchical clustering is the process of recursively paritioning a group of samples. 
 While standard NMF is interpreted as a flat clustering, it can also be extended for hierarchical clustering.
 Kuang and Park \cite{KP13} propose a method that uses rank-2 NMF to recursively bipartition the samples.
-An NMF with $k=2$ determines a row of $\M{H}$ of length 2 for each sample that determines its part membership.
-By recursively bipartitioning the samples in each part, the method determines a binary tree such that each sample belongs to a unique leaf and internal nodes of the tree determine higher-level clusters of the samples.
-A single latent feature vector for each node can also be used for interpreting clusters.
-Similarly, Du et al.
-\cite{DKDP17} show that the latent features vectors of the leaves can also be used to initialize a $\M{W}$ within an algorithm for flat NMF, yielding fast convergence and overall speedup compared to initializing the algorithm with random data.
+The method determines a binary tree such that all leaves contain unique samples and the structure of the tree determines hierarchical clusters.A single $\M{W}$ vector for each node can also be used for cluster interpretation.
 We discuss the hierarchical method in more detail in \Cref{sec:prelim} and \Cref{sec:seqalgs}.

### 2020-07-16 latex/tex/abstract.tex
little edits

@@
+% !TEX root = ../paper.tex
+
 \begin{abstract}
@@
 We demonstrate the scalability of our parallel algorithm in terms of data size (up to 800 GB) and number of processors (up to 80 nodes of the Summit supercomputer), applying the hierarchical clustering approach to hyperspectral imaging and image classification data.
-Our algorithm for Rank-2 NMF scales perfectly on up to 1000s of cores and the entire hierarchical clustering method achieves 5.9x speedup scaling from 10 to 80 nodes on an 800 GB dataset.
+Our algorithm for Rank-2 NMF scales perfectly on up to 1000s of cores and the entire hierarchical clustering method achieves 5.9x speedup scaling from 10 to 80 nodes on the 800 GB dataset.
 \end{abstract}

### 2020-07-16 latex/tex/experiments.tex
little edits
briging down to 10 pages

--- 2020-07-16 latex/tex/experiments.tex
@@
 \begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/dc_hierarchical_speedup.pdf}
+\includegraphics[height=1.5in, width=\columnwidth]{plots/dc_hierarchical_speedup.pdf}
 \caption{Strong Scaling Speedup for Hierarchical Clustering on \hyper{}}
@@
 \begin{figure}
+\begin{subfigure}{1.5in}
 \begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_speedup.pdf}
-\caption{Strong Scaling Speedup for Rank-2 NMF on Synthetic Data}
+\includegraphics[height=1.4in, width=1.4in]{plots/synthetic_rank2_speedup.pdf}
+\caption{Synthetic Data}
 \label{fig:synrank2speedup}
 \end{center}
+\end{subfigure}
+\begin{subfigure}{1.5in}
+\begin{center}
+\includegraphics[height=1.4in, width=1.4in]{plots/realworld_rank2_speedup.pdf}
+\caption{\image{} Data}
+\label{fig:rwrank2speedup}
+\end{center}
+\end{subfigure}
+\caption{Strong Scaling Speedup for Rank-2 NMF }
+\label{fig:rank2speedup}
 \end{figure}
@@
 \begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_strongscaling.pdf}
+\includegraphics[height=1.5in, width=\columnwidth]{plots/synthetic_rank2_strongscaling.pdf}
 \caption{Relative Time Breakdown for Rank-2 NMF on Synthetic Data}
@@
 \begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_speedup.pdf}
-\caption{Strong Scaling Speedup for Rank-2 NMF on \image{} Data}
-\label{fig:rwrank2speedup}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_strongscaling.pdf}
+\includegraphics[height=1.5in, width=\columnwidth]{plots/realworld_rank2_strongscaling.pdf}
 \caption{Relative Time Breakdown for Rank-2 NMF on \image{} Data}
@@
 \begin{figure}
+\begin{subfigure}{1.5in}
 \begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hierarchical_speedup.pdf}
-\caption{Strong Scaling Speedup for Hierarchical Clustering on Synthetic Data}
+\includegraphics[height=1.4in, width=1.4in]{plots/synthetic_hierarchical_speedup.pdf}
+\caption{Synthetic Data}
 \label{fig:synhierspeedup}
 \end{center}
+\end{subfigure}
+\begin{subfigure}{1.5in}
+\begin{center}
+\includegraphics[height=1.4in, width=1.4in]{plots/realworld_hierarchical_speedup.pdf}
+\caption{\image{} Data}
+\label{fig:rwhierspeedup}
+\end{center}
+\end{subfigure}
+\caption{Strong Scaling Speedup for Hierarchical Clustering}
+\label{fig:rank2speedup}
 \end{figure}
@@
 \begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hier_strongscaling.pdf}
+\includegraphics[height=1.5in, width=\columnwidth]{plots/synthetic_hier_strongscaling.pdf}
 \caption{Relative Time Breakdown for Hierarchical Clustering on Synthetic Data}
@@
 \begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/realworld_hierarchical_speedup.pdf}
-\caption{Strong Scaling Speedup for Hierarchical Clustering on \image{} Data}
-\label{fig:rwhierspeedup}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/realworld_hier_strongscaling.pdf}
+\includegraphics[height=1.5in, width=\columnwidth]{plots/realworld_hier_strongscaling.pdf}
 \caption{Relative Time Breakdown for Hierarchical Clustering on \image{} Data}
@@
 \begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_sequential_level_breakdown.pdf}
+\includegraphics[height=1.5in, width=\columnwidth]{plots/synthetic_sequential_level_breakdown.pdf}
 \caption{Levels Breakdown Timings for 1 Compute Node on Synthetic Data}
@@
 \begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_parallel_level_breakdown.pdf}
+\includegraphics[height=1.5in, width=\columnwidth]{plots/synthetic_parallel_level_breakdown.pdf}
 \caption{Levels Breakdown Timings for 40 Compute Nodes on Synthetic Data}
@@
 %\begin{center}
-%\includegraphics[height=2in, width=\columnwidth]{plots/realworld_parallel_level_breakdown.pdf}
+%\includegraphics[height=1.5in, width=\columnwidth]{plots/realworld_parallel_level_breakdown.pdf}
 %\caption{Levels Breakdown Timings for 80 Compute Nodes on \image{} Data}

### 2020-07-16 latex/tex/experiments.tex

--- 2020-07-16 latex/tex/experiments.tex
@@
 Each POWER9 processor utilizes 22 IBM SIMD Multi-Cores (SMCs), although one of these SMCs on each processor is dedicated to memory transfer and is therefore not available for computation. 
-For node scaling experiments, all 42 available SMCs were utilized in each node so that every node computed with 42 separate MPI processes.
+For node scaling experimeheight=2innts, all 42 available SMCs were utilized in each node so that every node computed with 42 separate MPI processes.
 Additionally, every node also supports six NVIDIA Volta V100 accelerators but these were unused by our algorithm.
@@
 	to this dataset as \hyper{}\cite{DC-HYDICE}.
-
 	\hyper{} is formatted into a 3-way tensor representing two spatial dimensions of pixels and one dimension of spectral bands. So, a slice along
@@
 	$3145728 \times 33126$, which is approximately 800 GB in size. 
-
 	Given its size, \image{} requries $10$ Summit nodes to perform hierarchical clustering.
@@
 \begin{figure}
+\begin{subfigure}{1.5in}
 \begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_speedup.pdf}
-\caption{Strong Scaling for Rank-2 NMF on Synthetic}
+\includegraphics[height=1.4in, width=1.4in]{plots/synthetic_rank2_speedup.pdf}
+\caption{Synthetic Data}
 \label{fig:synrank2speedup}
 \end{center}
+\end{subfigure}
+\begin{subfigure}{1.5in}
+\begin{center}
+\includegraphics[height=1.4in, width=1.4in]{plots/realworld_rank2_speedup.pdf}
+\caption{\image{} Data}
+\label{fig:rwrank2speedup}
+\end{center}
+\end{subfigure}
+\caption{Strong Scaling Speedup for Rank-2 NMF }
+\label{fig:rank2speedup}
 \end{figure}
@@
 \begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_speedup.pdf}
-\caption{Strong Scaling for Rank-2 NMF on \image{}}
-\label{fig:rwrank2speedup}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_strongscaling.pdf}
-\caption{Time Breakdown for Rank-2 NMF on \image{}}
+\caption{Time Breakdown for Rank-2 NMF on \image{} }
 \label{fig:rwrank2strongscaling}
@@
 \begin{figure}
+\begin{subfigure}{1.5in}
 \begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hierarchical_speedup.pdf}
-\caption{Strong Scaling for Hierarchical Clustering on Synthetic}
+\includegraphics[height=1.4in, width=1.4in]{plots/synthetic_hierarchical_speedup.pdf}
+\caption{Synthetic Data}
 \label{fig:synhierspeedup}
 \end{center}
+\end{subfigure}
+\begin{subfigure}{1.5in}
+\begin{center}
+\includegraphics[height=1.4in, width=1.4in]{plots/realworld_hierarchical_speedup.pdf}
+\caption{\image{} Data}
+\label{fig:rwhierspeedup}
+\end{center}
+\end{subfigure}
+\caption{Strong Scaling Speedup for Hierarchical Clustering}
+\label{fig:rank2speedup}
 \end{figure}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hier_strongscaling.pdf}
-\caption{Time Breakdown for Hierarchical Clustering on Synthetic}
+\caption{Relative Time Breakdown for Hierarchical Clustering on Synthetic Data}
 \label{fig:synhierstrongscaling}
@@
 \begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/realworld_hierarchical_speedup.pdf}
-\caption{Strong Scaling for Hierarchical Clustering on \image{}}
-\label{fig:rwhierspeedup}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_hier_strongscaling.pdf}
-\caption{Time Breakdown for Hierarchical Clustering on \image{}}
+\caption{Relative Time Breakdown for Hierarchical Clustering on \image{}}
 \label{fig:rwhierstrongscaling}

### 2020-07-16 latex/tex/experiments.tex
centering speedup plots
figure caption renaming

@@
 \includegraphics[height=2in, width=\columnwidth]{plots/dc_hierarchical_speedup.pdf}
-\caption{Strong Scaling for Hierarchical Clustering on \hyper{}}
+\caption{Strong Scaling for Clustering on \hyper{}}
 \label{fig:dcspeedup}
@@
 \end{subfigure}
-\caption{Strong Scaling Speedup for Rank-2 NMF }
+\caption{Strong Scaling Speedup for Rank-2 NMF}
 \label{fig:rank2speedup}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_strongscaling.pdf}
-\caption{Time Breakdown for Rank-2 NMF on \image{} }
+\caption{Time Breakdown for Rank-2 NMF on \image{}}
 \label{fig:rwrank2strongscaling}
@@
 \end{subfigure}
-\caption{Strong Scaling Speedup for Hierarchical Clustering}
+\caption{Strong Scaling Speedup for Clustering}
 \label{fig:rank2speedup}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_hier_strongscaling.pdf}
-\caption{Relative Time Breakdown for Hierarchical Clustering on Synthetic Data}
+\caption{Time Breakdown for Clustering on Synthetic}
 \label{fig:synhierstrongscaling}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/realworld_hier_strongscaling.pdf}
-\caption{Relative Time Breakdown for Hierarchical Clustering on \image{}}
+\caption{Time Breakdown for Clustering on \image{}}
 \label{fig:rwhierstrongscaling}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_sequential_level_breakdown.pdf}
-\caption{Levels Breakdown Timings for 1 Compute Node on Synthetic}
+\caption{Level Timings for 1 Compute Node on Synthetic}
 \label{fig:seqlevelbreakdown}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_parallel_level_breakdown.pdf}
-\caption{Levels Breakdown Timings for 40 Compute Nodes on Synthetic}
+\caption{Level Timings for 40 Compute Nodes on Synthetic}
 \label{fig:parallellevelbreakdown}

### 2020-07-16 latex/tex/introduction.tex

@@
 If the algorithm continues, it chooses to split the leaf node that provides the greatest benefit to the overall tree, which can be quantified as a node's ``score'' in various ways.
-The algorithm proceeds to split nodes until a target number of leaves is obtained, until no node is worth splitting, or some other stopping criterion is met.
-\LM{Should we have text explaining it or could we add these labels to the figure itself?}
 
@@
 \input{fig/dc}
-\caption{Hierarchical Clustering of DC Mall Hyperspectral Image}
+\caption{Hierarchical Clustering of DC Mall HSI}
 \label{fig:dc}

### 2020-07-16 latex/tex/experiments.tex
to 10 pages

@@
 \begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/dc_hierarchical_speedup.pdf}
+\includegraphics[height=1.4in, width=1.4in]{plots/dc_hierarchical_speedup.pdf}
 \caption{Strong Scaling for Clustering on \hyper{}}
@@
 \begin{figure}
+\centering
 \begin{subfigure}{1.5in}
@@
 
-From \Cref{eq:treecost}, we expect to see perfect strong scaling in a computationally bound hierarchical clustering problem with target cluster count $k=100$.
+From \Cref{eq:treecost}, we expect to see perfect strong scaling in a computationally bound clustering problem with target cluster count $k=100$.
 As $k$ is large, we expect the latency cost of small problems deep in the tree to limit scalability.

### 2020-07-16 latex/tex/experiments.tex
shortened captions

--- 2020-07-16 latex/tex/experiments.tex
@@
 \begin{center}
-\includegraphics[height=2in, width=\columnwidth]{plots/dc_hierarchical_speedup.pdf}
+\includegraphics[height=1.5in, width=.7\columnwidth]{plots/dc_hierarchical_speedup.pdf}
 \caption{Strong Scaling for Clustering on \hyper{}}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_sequential_level_breakdown.pdf}
-\caption{Level Timings for 1 Compute Node on Synthetic}
+\caption{Level Times for 1 Compute Node on Synthetic}
 \label{fig:seqlevelbreakdown}
@@
 \includegraphics[height=2in, width=\columnwidth]{plots/synthetic_parallel_level_breakdown.pdf}
-\caption{Level Timings for 40 Compute Nodes on Synthetic}
+\caption{Level Times for 40 Compute Nodes on Synthetic}
 \label{fig:parallellevelbreakdown}

### 2020-11-13 latex/paper.tex
briging down to 10 pages
added back in funding acks
adding description of rank scaling experiment, still need to add figure

--- 2020-07-16 latex/paper.tex
@@
-\documentclass[conference,compsoc]{IEEEtran}
-\IEEEoverridecommandlockouts
+\documentclass[10pt, conference]{IEEEtran}
+%\IEEEoverridecommandlockouts
 % The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
@@
 \usepackage{ourmacros}
+\usepackage{subcaption}
+
+\usepackage[subtle,tracking=normal,leading=tight,leadingfraction=0.98, indent=tight, wordspacing=normal, bibliography=tight]{savetrees}
+
 \def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
     T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+
+\usepackage{paralist}
+\renewenvironment{itemize}[1]{\begin{compactitem}#1}{\end{compactitem}}
 
@@
 
-\title{Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization
-\thanks{This material is based upon work supported by the National Science Foundation under Grant No. OAC-1642385 and OAC-1642410.
-This manuscript has been co-authored by UT-Battelle, LLC under Contract No. DE-AC05-00OR22725 with the U.S. Department of
-Energy. This project was partially funded by the Laboratory Director's Research and Development fund. This research used resources
-of the Oak Ridge Leadership Computing Facility at the Oak Ridge National Laboratory, which is supported by the Office of Science of
-the U.S. Department of Energy.}
-}
+\title{Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization}
 
@@
 
-%\begin{IEEEkeywords}
-%component, formatting, style, styling, insert
-%\end{IEEEkeywords}
+\begin{IEEEkeywords}
+low-rank approximation, distributed-memory parallel algorithms, scalable clustering
+\end{IEEEkeywords}
 
@@
 
+%Add this back in for camera ready
 \input{tex/ack.tex}
 
+\bibliographystyle{plainurl}
 \bibliography{paper}
-\bibliographystyle{plainurl}
 

### 2020-11-13 latex/tex/experiments.tex
adding description of rank scaling experiment, still need to add figure

--- 2020-07-16 latex/tex/experiments.tex
@@
 Each POWER9 processor utilizes 22 IBM SIMD Multi-Cores (SMCs), although one of these SMCs on each processor is dedicated to memory transfer and is therefore not available for computation. 
-For node scaling experimeheight=2innts, all 42 available SMCs were utilized in each node so that every node computed with 42 separate MPI processes.
+For node scaling experiments, all 42 available SMCs were utilized in each node so that every node computed with 42 separate MPI processes.
 Additionally, every node also supports six NVIDIA Volta V100 accelerators but these were unused by our algorithm.
@@
 %\end{figure}
+
+\subsubsection{Rank Scaling}
+
+To confirm the slow growth in running time of the hierarchical algorithm in terms of the number of clusters $k$, we perform rank scaling experiments for \hyper{} and synthetic data. 
+Assuming a balanced tree and relatively small $k$, \Cref{eq:treecost} shows that the dominant computational cost is proportional to $\log k$, while flat NMF algorithms have a dominant cost that is linear in $k$ \cite{KBP17}.
+\Cref{fig:rankscaling} shows the raw time for various values of $k$, confirming that running time for HierNMF grows more slowly in $k$ than a flat NMF algorithm from PLANC \cite{EH+19-TR} with the same number of columns and processor grid.
+
+\begin{figure}
+\caption{\GB{To be added...}}
+\label{fig:rankscaling}
+\end{figure}

### 2020-11-14 latex/tex/algorithm.tex
tightening itemize
tightening back
fixed bug in algorithm pointed out by reviewer
updated Jupyter notebook for new plots, tighted back to 10 pages

@@
+% !TEX root = ../paper.tex
+
 \section{Algorithms}
@@
 where $\V{w}_1$ and $\V{W}_2$ are the two columns of $\M{W}$ and $\V{a}_i$ is the $i$ column of $\M{A}$.
-We note that there are four possibilities of solutions, as each of the two scalar variables may be positive or zero.
+We note that there are four possibilities of solutions, as each of the two variables may be positive or zero.
 
@@
 	\item if $\M{W}$ and $\M{A}$ are both nonnegative, then the candidate solution with two zero values is never (uniquely) optimal and can be discarded, and
-	\item if the solution to the unconstrained problem does not admit two positive values, the better of the two remaining solutions can be determined by comparing $\V{a}_{j}^\Tra\V{w}_1 / \|\V{w}_1\|$ and $\V{a}_{j}^\Tra\V{w}_2 / \|\V{w}_2\|$.
+	\item if the unconstrained problem does not admit a positive solution, the better of the two remaining solutions can be determined by comparing $\V{a}_{j}^\Tra\V{w}_1 / \|\V{w}_1\|$ and $\V{a}_{j}^\Tra\V{w}_2 / \|\V{w}_2\|$.
 \end{itemize}
@@
 A Rank-2 NMF computes factor matrices $\M{W}$ and $\M{H}$ to approximate $\M{A}$, the values of $\M{H}$ are used to determine child membership of each column (either red or blue), and the corresponding column of the $\M{W}$ matrix represents the part's feature weighting.
-The 1D distribution is depicted for 3 processors to show that the splitting requires no interprocessor redistribution as the children are evenly distributed identically to the parent.}
+The 1D distribution is depicted for 3 processors to show that splitting requires no interprocessor redistribution as children are evenly distributed identically to the parent.}
 \label{fig:split}
@@
 			\State $\M[\hat]{G}_W = \M[\hat]{W}^\Tra \M[\hat]{W}$
-			\State $\M{G}_W = \textsc{All-Reduce}(\M[\hat]{G})$
+			\State $\M{G}_W = \textsc{All-Reduce}(\M[\hat]{G}_W)$
 			\State $\M[\hat]{B} = \M[\hat]{A}^\Tra \M[\hat]{W}$ %\hfill \Comment{local matrix multiplication}

### 2020-11-14 latex/tex/experiments.tex
updated Jupyter notebook for new plots, tighted back to 10 pages

@@
 \begin{center}
-\includegraphics[height=1.4in, width=1.4in]{plots/dc_hierarchical_speedup.pdf}
+\includegraphics[height=1.4in, width=1.5in]{plots/dc_hierarchical_speedup.pdf}
 \caption{Strong Scaling for Clustering on \hyper{}}
@@
 The theory (\Cref{eq:r2nmfcost}) suggests that perfect strong scaling is possible as long as the execution time is dominated by local computation.
-Both the matrix multiplications and NNLS solves scale linearly with $1/p$ (we expect MatMul to dominate), but the bandwidth cost is independent of $p$ and the latency cost increases slightly with $p$.
+Both the matrix multiplications and NNLS solves scale linearly with $1/p$ (we expect MatMul to dominate), but the bandwidth cost is independent of $p$ and latency increases slightly with $p$.
 
@@
 \begin{center}
-\includegraphics[height=1.4in, width=1.4in]{plots/synthetic_rank2_speedup.pdf}
+\includegraphics[height=1.4in, width=1.5in]{plots/synthetic_rank2_speedup.pdf}
 \caption{Synthetic Data}
@@
 \begin{center}
-\includegraphics[height=1.4in, width=1.4in]{plots/realworld_rank2_speedup.pdf}
+\includegraphics[height=1.4in, width=1.5in]{plots/realworld_rank2_speedup.pdf}
 \caption{\image{} Data}
@@
 \begin{center}
-\includegraphics[height=1.4in, width=1.4in]{plots/synthetic_hierarchical_speedup.pdf}
+\includegraphics[height=1.4in, width=1.5in]{plots/synthetic_hierarchical_speedup.pdf}
 \caption{Synthetic Data}
@@
 \begin{center}
-\includegraphics[height=1.4in, width=1.4in]{plots/realworld_hierarchical_speedup.pdf}
+\includegraphics[height=1.4in, width=1.5in]{plots/realworld_hierarchical_speedup.pdf}
 \caption{\image{} Data}
@@
 To confirm the slow growth in running time of the hierarchical algorithm in terms of the number of clusters $k$, we perform rank scaling experiments for \hyper{} and synthetic data. 
-Assuming a balanced tree and relatively small $k$, \Cref{eq:treecost} shows that the dominant computational cost is proportional to $\log k$, while flat NMF algorithms have a dominant cost that is linear in $k$ \cite{KBP17}.
-\Cref{fig:rankscaling} shows the raw time for various values of $k$, confirming that running time for HierNMF grows more slowly in $k$ than a flat NMF algorithm from PLANC \cite{EH+19-TR} with the same number of columns and processor grid.
+Assuming a balanced tree and relatively small $k$, \Cref{eq:treecost} shows that the dominant computational cost is proportional to $\log k$, while a flat NMF algorithm has a dominant cost that is linear in $k$ \cite{KBP17}.
+\Cref{fig:rankscaling} shows the raw time for various values of $k$, confirming that running time for HierNMF grows more slowly in $k$ than a flat NMF algorithm (based on Block Principal Pivoting) from PLANC \cite{EH+19-TR} with the same number of columns and processor grid.
+We see that for sufficiently large $k$, the hierarchical algorithm outperforms flat NMF and it scales much better with $k$.
 
 \begin{figure}
-\caption{\GB{To be added...}}
+\centering
+\begin{subfigure}{1.5in}
+\begin{center}
+\includegraphics[height=1.4in, width=1.5in]{plots/dc_rankscaling.pdf}
+\caption{\hyper{} Data}
+\label{fig:dcrankscaling}
+\end{center}
+\end{subfigure}
+\begin{subfigure}{1.5in}
+\begin{center}
+\includegraphics[height=1.4in, width=1.5in]{plots/synthetic_rankscaling.pdf}
+\caption{Synthetic Data}
+\label{fig:synrankscaling}
+\end{center}
+\end{subfigure}
+\caption{Rank Scaling for Hierarchical and Flat NMF}
 \label{fig:rankscaling}

### 2020-11-15 latex/slides/slides.tex
slides draft

\documentclass{beamer}

\usepackage{cite,doi}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithm,algpseudocode}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{cleveref}
\usepackage{tikz}
\usepackage{mathtools}
\usepackage{../ourmacros}
\usepackage{subcaption}
\usepackage{import}
\usepackage{adjustbox}

\newcommand{\hyper}{DC-HYDICE}
\newcommand{\image}{SIIM-ISIC}

\title{Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization}

\author{
    Lawton Manning\inst{1}
    \and Grey Ballard\inst{1}\\
    \and Ramakrishnan Kannan\inst{2}
    \and Haesun Park\inst{3}
}

\institute{
    \inst{1}%
    Wake Forest University
    \and
    \inst{2}%
    Oak Ridge National Laboratory
    \and
    \inst{3}%
    Georgia Institute of Technology
}

\date{
    HiPC 2020
}

\begin{document}

\frame{\titlepage}

\begin{frame}{Nonnegative Matrix Factorization (NMF)}
    \begin{itemize}
        \item $\M{A} \approx \M{W}\M{H}^\Tra$
        \item approximate $\M{A}$ (features $\times$ samples) into $\M{W}$ (features $\times k$) and $\M{H}$ (samples $\times k$)
        \item nonnegativity gives interpretability of $\M{W}$ and $\M{H}$ as clusters and cluster membership, respectively
        \item applications
            \begin{itemize}
                \item text document clustering using bag of words or TF-IDF matrices
                \item hyperspectral imaging segmentation
            \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{Hierarchical NMF}
    \begin{itemize}
        \item repeatedly use NMF with $k = 2$ to create a hierarchical tree of clusters
        \item application: hyperspectral imaging
    \end{itemize}
    \begin{adjustbox}{max totalsize={.7\textwidth}{.6\textheight},center}
    \import{..}{fig/dc}
    \end{adjustbox}
\end{frame}

\begin{frame}{Solving NMF}
    \begin{itemize}
        \item constrained optimization: $$\min_{\M{W},\M{H}\geq \M{0}} \|\M{A} - \M{W}\M{H}^\Tra\|_2$$
        \item alternating update
        \begin{itemize}
            \item fix $\M{H}$ and solve the NNLS for $\M{W}$ exactly
            \item alternate and repeat until convergence
        \end{itemize}
        \item NNLS Solvers
        \begin{itemize}
            \item Block Principal Pivoting (BPP)
            \item ??? Other Method ???
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{Rank-2 NMF (R2NMF)}
    \begin{itemize}
        \item when $k = 2$, BPP can be solved quickly as the size of the active set is 4
        \item Active Sets for R2NMF (row by row)
        \begin{itemize}
            \item both columns nonnegative
            \item only left column nonnegative
            \item only right column nonnegative
            \item both columns negative
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{Parallel R2NMF}
\end{frame}

\begin{frame}{Splitting with R2NMF}
    \begin{itemize}
        \item use columns of $\M{H}^\Tra$ to split the columns of $\M{A}$ into two submatrices
        \item assign the corresponding columns of $\M{W}$ to the submatrices
        \item repeat NMF and split on submatrices
    \end{itemize}
    \begin{adjustbox}{max totalsize={.7\textwidth}{.6\textheight},center}
        \import{..}{fig/split}
    \end{adjustbox}
\end{frame}

\begin{frame}{HierNMF}
    \begin{itemize}
        \item call R2NMF on root node to split into two children
        \item use the power iteration to select children that least approximate rank-1 matrices
        \item repeat until maximum number of nodes is reached or all child matrices are rank-1
    \end{itemize}
    \begin{adjustbox}{max totalsize={.7\textwidth}{.6\textheight},center}
        \import{..}{fig/tree}
    \end{adjustbox}
\end{frame}

\begin{frame}{Parallel HierNMF}
    FIXME: Need data distribution figure
\end{frame}

\begin{frame}{Data Sets}
    \begin{itemize}
        \item DC-HYDICE
        \begin{itemize}
            \item Hyperspectral Digital Imagery Collection Experiment (HYDICE) of the National Mall in Washington, DC
        \end{itemize}
        \item SIIM-ISIC
        \begin{itemize}
            \item Society for Imaging Informatics in Medicine - International Skin Imaging Colloboration image classification of melanoma images
        \end{itemize}
        \item Synthetic Image classification
        \begin{itemize}
            \item smaller image classification dataset which has the same aspect ratio as SIIM-ISIC but small enough to fit in memory on one machine
        \end{itemize}
    \end{itemize}
\end{frame}
\end{document}

### 2020-11-15 latex/tex/ack.tex

@@
+%!TEX root = ../paper.tex
+
 \section*{Acknowledgment}
 
+%\small
+This material is based upon work supported by the National Science Foundation under Grant No. OAC-1642385 and OAC-1642410.
+This manuscript has been co-authored by UT-Battelle, LLC under Contract No. DE-AC05-00OR22725 with the U.S. Department of
+Energy. This project was partially funded by the Laboratory Director's Research and Development fund. This research used resources
+of the Oak Ridge Leadership Computing Facility at the Oak Ridge National Laboratory, which is supported by the Office of Science of
+the U.S. Department of Energy.
+
 The authors would like to thank Simin Ma for contributions to the algorithmic analysis and John Farrell for his contributions to the implementation of the parallel algorithm.
+%\normalsize

### 2020-11-15 latex/tex/conclusion.tex

@@
 This will reduce the communicated data and achieve a local data matrix that is more square, which can improve local matrix multiplication performance.
-The downside of this approach is that it requires a redistribution of the data for each split, but if many NMF iterations are required, then the single upfront redistribution cost may be amortized.
+The downside of this approach is requiring a redistribution of the data for each split, but if many NMF iterations are required, then the single upfront cost may be amortized.
 

### 2020-11-16 latex/slides/slides.tex
added theme and metadata to slides

@@
 \newcommand{\image}{SIIM-ISIC}
+
+\definecolor{wfugold}{rgb}{0.6196078,0.494117647,0.21960784}
+\newcommand{\red}[1]{\textcolor{red}{#1}}
+\newcommand{\blue}[1]{\textcolor{blue}{#1}}
+\newcommand{\multiplycolor}{red}
+\newcommand{\zero}{}
+\newcommand{\cred}[1]{\textcolor{red}{#1}}
+\newcommand{\cblue}[1]{\textcolor{blue}{#1}}
+\newcommand{\cgold}[1]{\textcolor{wfugold}{#1}}
+\newcommand{\email}[1]{\href{mailto:#1}{\texttt{#1}}}
 
@@
 \date{
+    27th IEEE International Conference on High Performance Computing, Data, \& Analytics (HiPC 2020)
+}
+
+\usetheme{Warsaw}
+\usecolortheme[named=wfugold]{structure}
+%Madrid
+%\usecolortheme{dolphin}
+%\useinnertheme{rounded}
+%\usefonttheme{serif}
+\setbeamertemplate{navigation symbols}{} % gets rid of navigation bars
+\setbeamertemplate{footline}
+{
+  \hbox{%
+  \begin{beamercolorbox}[wd=.33\paperwidth,ht=2.25ex,dp=1ex,left]{author in head/foot}%
+    \usebeamerfont{author in head/foot}
+    Manning, Ballard, Kannan, Park
+  \end{beamercolorbox}%
+  \begin{beamercolorbox}[wd=.34\paperwidth,ht=2.25ex,dp=1ex,center]{title in head/foot}%
+    \usebeamerfont{title in head/foot}
     HiPC 2020
+  \end{beamercolorbox}%
+  \begin{beamercolorbox}[wd=.33\paperwidth,ht=2.25ex,dp=1ex,right]{date in head/foot}%
+    \usebeamerfont{date in head/foot}
+    \insertframenumber{} \hspace*{2ex} 
+  \end{beamercolorbox}}%
 }
@@
 \end{frame}
+
+\begin{frame}
+\frametitle{Author Contact Information}
+
+For questions, please contact
+\begin{itemize}
+	\item Lawton Manning: \email{mannlg15@wfu.edu}
+	\item Grey Ballard: \email{ballard@wfu.edu}
+	\item Ramakrishnan Kannan: \email{kannanr@ornl.gov}
+	\item Haesun Park: \email{hpark@cc.gatech.edu}
+\end{itemize}
+
+\end{frame}
+
 \end{document}

### 2020-11-17 latex/slides/slides.tex
finished draft

@@
 \usepackage{algorithm,algpseudocode}
-\usepackage{graphicx}
+\usepackage{stackengine,graphicx}
 \usepackage{textcomp}
@@
 \usepackage{tikz}
+\usetikzlibrary{3d}
+\usetikzlibrary{patterns}
+\usetikzlibrary{calc}
+\usetikzlibrary{arrows}
+\usetikzlibrary{matrix}
+\usetikzlibrary{positioning}
+\usetikzlibrary{decorations.pathreplacing}
 \usepackage{mathtools}
-\usepackage{../ourmacros}
 \usepackage{subcaption}
 \usepackage{import}
+\usepackage{bm}
+\usepackage[mathscr]{eucal}
+\usepackage{amsbsy}
+\usepackage[section]{placeins}
 \usepackage{adjustbox}
+\stackMath
+
+\usepackage{../ourmacros}
+
+\newcommand{\T}[1]{\boldsymbol{\mathscr{#1}}}
 
@@
 
-\title{Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization}
+\newlength\matfield
+\newlength\tmplength
+\def\matscale{1.}
+\newcommand{\dimbox}[3]{%
+  \setlength\matfield{\matscale\baselineskip}%
+  \setbox0=\hbox{\vphantom{X}\smash{#3}}%
+  \setlength{\tmplength}{#1\matfield-\ht0-\dp0}%
+  \fboxrule=1pt\fboxsep=-\fboxrule\relax%
+  \fbox{\makebox[#2\matfield]{\addstackgap[.5\tmplength]{\box0}}}%
+}
+\newcommand{\raiserows}[2]{%
+   \setlength\matfield{\matscale\baselineskip}%
+   \raisebox{#1\matfield}{#2}%
+}
+\newcommand{\matbox}[5]{
+  \stackunder{\dimbox{#1}{#2}{$#5$}}{\scriptstyle(#3\times #4)}%
+}
+
+\title{Parallel Algorithms for Low-Rank Approximations of Matrices and Tensors}
 
 \author{
-    Lawton Manning\inst{1}
-    \and Grey Ballard\inst{1}\\
-    \and Ramakrishnan Kannan\inst{2}
-    \and Haesun Park\inst{3}
-}
-
-\institute{
-    \inst{1}%
-    Wake Forest University
-    \and
-    \inst{2}%
-    Oak Ridge National Laboratory
-    \and
-    \inst{3}%
-    Georgia Institute of Technology
-}
-
-\date{
-    27th IEEE International Conference on High Performance Computing, Data, \& Analytics (HiPC 2020)
+    Lawton Manning
 }
@@
     \usebeamerfont{author in head/foot}
-    Manning, Ballard, Kannan, Park
+    Manning
   \end{beamercolorbox}%
@@
     \usebeamerfont{title in head/foot}
-    HiPC 2020
+    Thesis Proposal
   \end{beamercolorbox}%
@@
 
+\begin{frame}{Introduction}
+    \begin{itemize}
+        \item Parallel Hierarchical Clustering using Rank-Two Nonnegative Matrix Factorization
+        \item Tensor Train Rounding Using Gram Matrices
+    \end{itemize}
+\end{frame}
+
 \begin{frame}{Nonnegative Matrix Factorization (NMF)}
+    \begin{adjustbox}{max totalsize={.4\textwidth}{.5\textheight},center}
+    \import{..}{fig/nmf}
+    \end{adjustbox}
     \begin{itemize}
-        \item $\M{A} \approx \M{W}\M{H}^\Tra$
         \item approximate $\M{A}$ (features $\times$ samples) into $\M{W}$ (features $\times k$) and $\M{H}$ (samples $\times k$)
@@
                 \item text document clustering using bag of words or TF-IDF matrices
-                \item hyperspectral imaging segmentation
+                \item hyperspectral image segmentation
             \end{itemize}
@@
         \item constrained optimization: $$\min_{\M{W},\M{H}\geq \M{0}} \|\M{A} - \M{W}\M{H}^\Tra\|_2$$
-        \item alternating update
+        \item Alternating Nonnegatively Constrained Least Squares (ANLS)
         \begin{itemize}
-            \item fix $\M{H}$ and solve the NNLS for $\M{W}$ exactly
+            \item fix $\M{H}$ and solve the NLS for $\M{W}$ exactly
             \item alternate and repeat until convergence
         \end{itemize}
-        \item NNLS Solvers
+
+        \item Block Principal Pivoting (BPP)
         \begin{itemize}
-            \item Block Principal Pivoting (BPP)
-            \item ??? Other Method ???
+            \item solve unconstrained system
+            \item $\M{W}^\Tra\M{W}$ and $\M{W}^\Tra\M{A}$ for $\M{H}$ and likewise for $\M{W}$
+            \item account for negative solutions with active-set method (quadratic in $k$)
         \end{itemize}
@@
 
-\begin{frame}{Rank-2 NMF (R2NMF)}
+\begin{frame}{Rank-2 NMF (R2NMF) NLS}
     \begin{itemize}
-        \item when $k = 2$, BPP can be solved quickly as the size of the active set is 4
+        \item when $k = 2$, the NLS can be solved quickly as the size of the active set is 4
         \item Active Sets for R2NMF (row by row)
@@
 \begin{frame}{Parallel R2NMF}
+    \begin{itemize}
+        \item distribute rows of $\M{A}$ onto separate processors
+        \item distribute rows of $\M{W}$ and $\M{H}$ evenly
+        \begin{itemize}
+            \item $\M{W}^\Tra\M{A}$ and $\M{W}^\Tra\M{W}$ for $\M{H}$
+            \item $(\M{A}\M{H})^\Tra$ and $\M{H}^\Tra\M{H}$ for $\M{W}$
+        \end{itemize}
+    \end{itemize}
+        \import{..}{fig/r2nmf}
 \end{frame}
@@
     \begin{itemize}
-        \item use columns of $\M{H}^\Tra$ to split the columns of $\M{A}$ into two submatrices
+        \item use columns of $\M{H}^\Tra$ to partition the columns of $\M{A}$ into clusters
+        \item assign the cluster columns into two submatrices
         \item assign the corresponding columns of $\M{W}$ to the submatrices
@@
     \begin{itemize}
-        \item call R2NMF on root node to split into two children
-        \item use the power iteration to select children that least approximate rank-1 matrices
+        \item split root node into two children with R2NMF
+        \item power iteration $S\left(\cal N\right) = \sigma_1^2\left(\cal L\right) +
+        \sigma_1^2\left(\cal R\right) -
+        \sigma_1^2\left(\cal N\right)$
+        \begin{itemize}
+            \item heuristic for choosing next node to split
+            \item requires node to have precomputed children
+        \end{itemize}
         \item repeat until maximum number of nodes is reached or all child matrices are rank-1
@@
     \end{adjustbox}
-\end{frame}
-
-\begin{frame}{Parallel HierNMF}
-    FIXME: Need data distribution figure
 \end{frame}
@@
 
+\begin{frame}{R2NMF Strong Scaling}
+    \centering
+    \begin{columns}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\textwidth]{../plots/synthetic_rank2_speedup.pdf}
+            \caption{Synthetic  Data}
+            \end{figure}
+        \end{column}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\textwidth]{../plots/realworld_rank2_speedup.pdf}
+            \caption{\image{} Data}
+            \end{figure}
+        \end{column}
+    \end{columns}
+\end{frame}
+
+\begin{frame}{R2NMF Scaling Time Breakdowns}
+    \centering
+    \begin{columns}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\textwidth]{../plots/synthetic_rank2_strongscaling.pdf}
+            \caption{Synthetic  Data}
+            \end{figure}
+        \end{column}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\textwidth]{../plots/realworld_rank2_strongscaling.pdf}
+            \caption{\image{} Data}
+            \end{figure}
+        \end{column}
+    \end{columns}
+\end{frame}
+
+\begin{frame}{HierNMF Strong Scaling}
+    \centering
+    \begin{columns}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\textwidth]{../plots/synthetic_hierarchical_speedup.pdf}
+            \caption{Synthetic  Data}
+            \end{figure}
+        \end{column}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\textwidth]{../plots/realworld_hierarchical_speedup.pdf}
+            \caption{\image{} Data}
+            \end{figure}
+        \end{column}
+    \end{columns}
+\end{frame}
+
+\begin{frame}{HierNMF Scaling Time Breakdowns}
+    \centering
+    \begin{columns}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\textwidth]{../plots/synthetic_hier_strongscaling.pdf}
+            \caption{Synthetic  Data}
+            \end{figure}
+        \end{column}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\textwidth]{../plots/realworld_hier_strongscaling.pdf}
+            \caption{\image{} Data}
+            \end{figure}
+        \end{column}
+    \end{columns}
+\end{frame}
+
+\begin{frame}{Tree Level Times on Synthetic Data}
+    \centering
+    \begin{columns}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\textwidth]{../plots/synthetic_sequential_level_breakdown.pdf}
+            \caption{1 Compute Node}
+            \end{figure}
+        \end{column}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\textwidth]{../plots/synthetic_parallel_level_breakdown.pdf}
+            \caption{40 Compute Nodes}
+            \end{figure}
+        \end{column}
+    \end{columns}
+\end{frame}
+
+\begin{frame}{Rank Scaling for Hierarchical and Flat NMF}
+    \centering
+    \begin{columns}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\textwidth]{../plots/dc_rankscaling.pdf}
+            \caption{\hyper{}  Data}
+            \end{figure}
+        \end{column}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\textwidth]{../plots/synthetic_rankscaling.pdf}
+            \caption{Synthetic Data}
+            \end{figure}
+        \end{column}
+    \end{columns}
+\end{frame}
+
 \begin{frame}

### 2020-11-17 latex/slides/slides.tex
making pass through slides

@@
 
+\newcommand{\GB}[1]{\textcolor{red}{\textbf{GB}: #1}}
 \newcommand{\hyper}{DC-HYDICE}
@@
 
-\title{Parallel Algorithms for Low-Rank Approximations of Matrices and Tensors}
+\title{Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization}
 
 \author{
-    Lawton Manning
+    Lawton Manning\inst{1}
+    \and Grey Ballard\inst{1}\\
+    \and Ramakrishnan Kannan\inst{2}
+    \and Haesun Park\inst{3}
+}
+
+\institute{
+    \inst{1}%
+    Wake Forest University
+    \and
+    \inst{2}%
+    Oak Ridge National Laboratory
+    \and
+    \inst{3}%
+    Georgia Institute of Technology
+}
+
+\date{
+    27th IEEE International Conference on High Performance Computing, Data, \& Analytics (HiPC 2020)
 }
@@
     \usebeamerfont{author in head/foot}
-    Manning
+    Manning, Ballard, Kannan, Park
   \end{beamercolorbox}%
@@
     \usebeamerfont{title in head/foot}
-    Thesis Proposal
+    HiPC 2020
   \end{beamercolorbox}%
@@
 
-\begin{frame}{Introduction}
+\begin{frame}{Summary}
+
     \begin{itemize}
-        \item Parallel Hierarchical Clustering using Rank-Two Nonnegative Matrix Factorization
-        \item Tensor Train Rounding Using Gram Matrices
+        \item Nonnegative data can be hierarchically clustered using recursive bipartitioning based on Rank-2 Nonnegative Matrix Factorization (R2-NMF)
+        \vfill
+        \item Scalable parallelization requires efficient iterative algorithm for R2-NMF as well as data distribution conducive to data-dependent hierarchy structure
+        \vfill
+        \item Our approach scales well to 1000s of cores and is able to cluster 800GB image dataset
     \end{itemize}
+
 \end{frame}
 
-\begin{frame}{Nonnegative Matrix Factorization (NMF)}
+\begin{frame}{Nonnegative Matrix Factorization (NMF) for Clustering}
     \begin{adjustbox}{max totalsize={.4\textwidth}{.5\textheight},center}
@@
     \begin{itemize}
-        \item approximate $\M{A}$ (features $\times$ samples) into $\M{W}$ (features $\times k$) and $\M{H}$ (samples $\times k$)
+        \item approximate $\M{A}$ (features $\times$ samples) into $\M{W}$ (features $\times$ clusters) and $\M{H}$ (samples $\times$ clusters)
+        \vfill
         \item nonnegativity gives interpretability of $\M{W}$ and $\M{H}$ as clusters and cluster membership, respectively
-        \item applications
-            \begin{itemize}
-                \item text document clustering using bag of words or TF-IDF matrices
-                \item hyperspectral image segmentation
-            \end{itemize}
+        \vfill
+        \item applications in text document topic modeling and hyperspectral image segmentation, for example
     \end{itemize}
@@
 \begin{frame}{Hierarchical NMF}
+
 \begin{itemize}
-        \item repeatedly use NMF with $k = 2$ to create a hierarchical tree of clusters
+	\item repeatedly use NMF with $k = 2$ to bipartition nodes in order to create a hierarchical tree of clusters
 	\item application: hyperspectral imaging
 \end{itemize}
-    \begin{adjustbox}{max totalsize={.7\textwidth}{.6\textheight},center}
+
+\begin{columns}
+\begin{column}{.4\textwidth}
+\GB{put bigger picture of DC mall (node 0) here, add pointers to leaves of tree for particular material (like grass or water)}
+\end{column}
+\begin{column}{.6\textwidth}
+	\begin{adjustbox}{width=\textwidth}
     		\import{..}{fig/dc}
    	\end{adjustbox}
+\end{column}
+\end{columns}
+
 \end{frame}
@@
     \begin{itemize}
-        \item constrained optimization: $$\min_{\M{W},\M{H}\geq \M{0}} \|\M{A} - \M{W}\M{H}^\Tra\|_2$$
-        \item Alternating Nonnegatively Constrained Least Squares (ANLS)
+        \item NMF constrained optimization problem: $$\min_{\M{W},\M{H}\geq \M{0}} \|\M{A} - \M{W}\M{H}^\Tra\|_2$$
+
+        \vfill
+
+        \item Alternating Nonnegative Least Squares (ANLS)
         \begin{itemize}
@@
 
-        \item Block Principal Pivoting (BPP)
+        \vfill
+
+        \item General rank case
         \begin{itemize}
-            \item solve unconstrained system
-            \item $\M{W}^\Tra\M{W}$ and $\M{W}^\Tra\M{A}$ for $\M{H}$ and likewise for $\M{W}$
-            \item account for negative solutions with active-set method (quadratic in $k$)
+            \item can solve each row exactly using iterative active-set method
+            \item determine which values are positive and which are zero
+        \end{itemize}
+
+        \vfill
+
+         \item Rank-2 case
+        \begin{itemize}
+            \item 4 possibilities: both positive, one or other is zero, both zeroes 
+            \item quickly solve each and choose the optimal solution
         \end{itemize}
@@
 
-\begin{frame}{Rank-2 NMF (R2NMF) NLS}
+\begin{frame}{Parallel R2-NMF}
     \begin{itemize}
-        \item when $k = 2$, the NLS can be solved quickly as the size of the active set is 4
-        \item Active Sets for R2NMF (row by row)
+        \item Use row distribution of $\M{A}$ and row distributions for $\M{W}$ and $\M{H}$ 
+        \item Computational bottlenecks are matrix multiplications
         \begin{itemize}
-            \item both columns nonnegative
-            \item only left column nonnegative
-            \item only right column nonnegative
-            \item both columns negative
-        \end{itemize}
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{Parallel R2NMF}
-    \begin{itemize}
-        \item distribute rows of $\M{A}$ onto separate processors
-        \item distribute rows of $\M{W}$ and $\M{H}$ evenly
-        \begin{itemize}
-            \item $\M{W}^\Tra\M{A}$ and $\M{W}^\Tra\M{W}$ for $\M{H}$
-            \item $(\M{A}\M{H})^\Tra$ and $\M{H}^\Tra\M{H}$ for $\M{W}$
+            \item compute $\M{W}^\Tra\M{A}$ using reduce-scatter for $\M{H}$
+            \item compute $\M{A}\M{H}$ using all-gather for $\M{W}$
+            \item compute $\M{W}^\Tra\M{W}$ and $\M{H}^\Tra\M{H}$ using all-reduce
         \end{itemize}
@@
 
-\begin{frame}{Splitting with R2NMF}
+\begin{frame}{Bipartitioning with R2-NMF}
     \begin{itemize}
-        \item use columns of $\M{H}^\Tra$ to partition the columns of $\M{A}$ into clusters
-        \item assign the cluster columns into two submatrices
-        \item assign the corresponding columns of $\M{W}$ to the submatrices
-        \item repeat NMF and split on submatrices
+        \item Use columns of $\M{H}^\Tra$ to bipartition the columns of $\M{A}$
+        \item Assign the cluster columns into two submatrices
+        \item Assign columns of $\M{W}$ as feature signatures of clusters
     \end{itemize}
@@
 
-\begin{frame}{HierNMF}
+\begin{frame}{Hierarchical Clusterings using R2-NMF (HierNMF2)}
     \begin{itemize}
-        \item split root node into two children with R2NMF
-        \item power iteration $S\left(\cal N\right) = \sigma_1^2\left(\cal L\right) +
-        \sigma_1^2\left(\cal R\right) -
-        \sigma_1^2\left(\cal N\right)$
+        \item Split root node into two children with R2-NMF
+        \item Choose next node to split based on scoring metric
         \begin{itemize}
-            \item heuristic for choosing next node to split
-            \item requires node to have precomputed children
+        		\item computing score requires pre-computing R2-NMF
 	\end{itemize}
-        \item repeat until maximum number of nodes is reached or all child matrices are rank-1
+        \item Repeat until maximum number of frontier nodes is reached
     \end{itemize}
@@
 
-\begin{frame}{Data Sets}
+\begin{frame}{Experimental Data Sets}
     \begin{itemize}
@@
         \end{itemize}
+        \vfill
         \item SIIM-ISIC
@@
         \end{itemize}
+        \vfill
         \item Synthetic Image classification
         \begin{itemize}
-            \item smaller image classification dataset which has the same aspect ratio as SIIM-ISIC but small enough to fit in memory on one machine
+            \item smaller image classification dataset which has the same aspect ratio as SIIM-ISIC but fits in one node's memory
         \end{itemize}
     \end{itemize}
+    \vfill
+    \begin{center}
+    \scriptsize
+    	All experiments performed on Summit at Oak Ridge Leadership Computing Facility
+    \end{center}
 \end{frame}

### 2020-11-18 latex/slides/slides.tex
finished pass on experimental slides

@@
     \begin{itemize}
-        \item DC-HYDICE
+        \item DC-HYDICE (191 $\times$ 392{,}960, 600MB)
         \begin{itemize}
@@
         \vfill
-        \item SIIM-ISIC
+        \item SIIM-ISIC (3{,}145{,}728 $\times$ 33{,}126, 800GB)
         \begin{itemize}
@@
         \vfill
-        \item Synthetic Image classification
+        \item Synthetic (1{,}048{,}576 $\times$ 11{,}042, 86GB)
         \begin{itemize}
-            \item smaller image classification dataset which has the same aspect ratio as SIIM-ISIC but fits in one node's memory
+            \item smaller synthetic image classification dataset which has the same aspect ratio as SIIM-ISIC but fits in one node's memory
         \end{itemize}
@@
 
-\begin{frame}{R2NMF Strong Scaling}
+\newcommand{\figscal}{.75\textwidth}
+
+\begin{frame}{R2-NMF Strong Scaling and Time Breakdown}
     \centering
@@
             \begin{figure}
-            \includegraphics[width=\textwidth]{../plots/synthetic_rank2_speedup.pdf}
+            \includegraphics[width=\figscal]{../plots/synthetic_rank2_speedup.pdf}
+            \caption{Synthetic Scaling}
+            \end{figure}
+        \end{column}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\figscal]{../plots/realworld_rank2_speedup.pdf}
+            \caption{\image{} Scaling}
+            \end{figure}
+        \end{column}
+    \end{columns}
+\vspace{-.5cm}
+    \begin{columns}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\figscal]{../plots/synthetic_rank2_strongscaling.pdf}
+            \caption{Synthetic Breakdown}
+            \end{figure}
+        \end{column}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\figscal]{../plots/realworld_rank2_strongscaling.pdf}
+            \caption{\image{} Breakdown}
+            \end{figure}
+        \end{column}
+    \end{columns}
+
+\begin{itemize}
+\small
+    \item Nearly perfect strong scaling because time dominated by local matrix multiplication
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{Hierarchical Clustering Scaling and Breakdown}
+    \centering
+    \begin{columns}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\figscal]{../plots/synthetic_hierarchical_speedup.pdf}
             \caption{Synthetic  Data}
@@
             \begin{figure}
-            \includegraphics[width=\textwidth]{../plots/realworld_rank2_speedup.pdf}
+            \includegraphics[width=\figscal]{../plots/realworld_hierarchical_speedup.pdf}
             \caption{\image{} Data}
@@
     \end{columns}
-\end{frame}
-
-\begin{frame}{R2NMF Scaling Time Breakdowns}
-    \centering
+\vspace{-.5cm}
     \begin{columns}
@@
             \begin{figure}
-            \includegraphics[width=\textwidth]{../plots/synthetic_rank2_strongscaling.pdf}
+            \includegraphics[width=\figscal]{../plots/synthetic_hier_strongscaling.pdf}
             \caption{Synthetic  Data}
@@
             \begin{figure}
-            \includegraphics[width=\textwidth]{../plots/realworld_rank2_strongscaling.pdf}
+            \includegraphics[width=\figscal]{../plots/realworld_hier_strongscaling.pdf}
             \caption{\image{} Data}
@@
     \end{columns}
-\end{frame}
 
-\begin{frame}{HierNMF Strong Scaling}
-    \centering
-    \begin{columns}
-        \begin{column}{0.5\textwidth}
-            \begin{figure}
-            \includegraphics[width=\textwidth]{../plots/synthetic_hierarchical_speedup.pdf}
-            \caption{Synthetic  Data}
-            \end{figure}
-        \end{column}
-        \begin{column}{0.5\textwidth}
-            \begin{figure}
-            \includegraphics[width=\textwidth]{../plots/realworld_hierarchical_speedup.pdf}
-            \caption{\image{} Data}
-            \end{figure}
-        \end{column}
-    \end{columns}
-\end{frame}
+\begin{itemize}
+\small
+    \item Scaling hindered by NNLS computations and communication of small subproblems at leaves of hierarchy
+\end{itemize}
 
-\begin{frame}{HierNMF Scaling Time Breakdowns}
-    \centering
-    \begin{columns}
-        \begin{column}{0.5\textwidth}
-            \begin{figure}
-            \includegraphics[width=\textwidth]{../plots/synthetic_hier_strongscaling.pdf}
-            \caption{Synthetic  Data}
-            \end{figure}
-        \end{column}
-        \begin{column}{0.5\textwidth}
-            \begin{figure}
-            \includegraphics[width=\textwidth]{../plots/realworld_hier_strongscaling.pdf}
-            \caption{\image{} Data}
-            \end{figure}
-        \end{column}
-    \end{columns}
 \end{frame}
@@
     \end{columns}
+
+\vfill
+
+\begin{itemize}
+    \item Evidence that scaling is hindered by leaves of tree
+    \begin{itemize}
+    	\item MatMul costs are constant over levels of balanced tree 
+	\item NNLS and communication costs grow over levels
+    \end{itemize}
+\end{itemize}
+
 \end{frame}
 
-\begin{frame}{Rank Scaling for Hierarchical and Flat NMF}
-    \centering
-    \begin{columns}
-        \begin{column}{0.5\textwidth}
-            \begin{figure}
-            \includegraphics[width=\textwidth]{../plots/dc_rankscaling.pdf}
-            \caption{\hyper{}  Data}
-            \end{figure}
-        \end{column}
-        \begin{column}{0.5\textwidth}
-            \begin{figure}
-            \includegraphics[width=\textwidth]{../plots/synthetic_rankscaling.pdf}
-            \caption{Synthetic Data}
-            \end{figure}
-        \end{column}
-    \end{columns}
+\begin{frame}{Conclusions and Future Directions}
+
+\begin{itemize}
+	\item Scalability is possible when MatMul dominates, particularly when input data is tall and skinny and tree is balanced
+	\vfill
+	\item 2D distribution of matrix $\M{A}$ could improve scalability but requires redistribution of $\M{A}$ for subtrees
+	\vfill
+	\item Parallelization across tree could improve scalability but requires careful load balancing and decentralizing split orders
+	\vfill
+	\item Hierarchical clustering can also be used to initialize flat NMF, particularly when $k$ is large, for Divide-and-Conquer NMF
+\end{itemize}
+
 \end{frame}

### 2020-11-18 latex/fig/dc.tex
labels to HYDICE tree

@@
             {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/1.png}};
+            \node[draw, text=orange, fill=white] at (-3,-2) {vegetation};
             \draw[thick,->] (-3.4,-1.2) .. controls (-4.4,-1.2) .. (-4.4,-2.4);
@@
                 {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/4.png}};
+                \node[draw, text=orange, fill=white] at (-4.4,-4) {grass};
             \node[below=0.1] at (3.south) {\textbf{3}};
@@
                 {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/3.png}};
+                \node[draw, text=orange, fill=white] at (-1.6,-4) {trees};
                 \draw[thick,->] (-2,-3.2) .. controls (-2.5,-3.2) .. (-2.5,-4.4);
@@
             {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/2.png}};
+            \node[draw, text=orange, fill=white] at (3,-2) {non-vegetation};
             \draw[thick,->] (2.6,-1.2) .. controls (1.6,-1.2) .. (1.6,-2.4);
@@
                 {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/6.png}};
+                \node[draw, text=orange, fill=white] at (1.6,-4) {buildings};
                 \draw[thick,->] (1.2,-3.2) .. controls (0.7,-3.2) .. (0.7,-4.4);
@@
                 {\includegraphics[width=0.8cm,height=3.2cm]{data/DC/5.png}};
+                \node[draw, text=orange, fill=white] at (4.4,-4) {water/roads};
             \node[below=0.1] at (6.south) {\textbf{6}};

### 2020-11-18 latex/slides/slides.tex
original HYDICE picture
Lawton/Grey pass over slides

@@
 }
+\usepackage{caption}
+\captionsetup[figure]{labelformat=empty}%
 
@@
     \begin{itemize}
-        \item Nonnegative data can be hierarchically clustered using recursive bipartitioning based on Rank-2 Nonnegative Matrix Factorization (R2-NMF)
+        \item Nonnegative data can be hierarchically clustered using recursive bipartitioning based on Rank-2 Nonnegative Matrix Factorization (NMF)
         \vfill
-        \item Scalable parallelization requires efficient iterative algorithm for R2-NMF as well as data distribution conducive to data-dependent hierarchy structure
+        \item Scalable parallelization requires efficient iterative algorithm for Rank-2 NMF as well as data distribution conducive to data-dependent hierarchy structure
         \vfill
         \item Our approach scales well to 1000s of cores and is able to cluster 800GB image dataset
+        \begin{itemize}
+        		\item scaling from 10 to 80 nodes, achieves $7\times$ parallel speedup for Rank-2 NMF and $6\times$ speedup for hierarchical clustering
+        \end{itemize}
     \end{itemize}
@@
     \begin{itemize}
-        \item approximate $\M{A}$ (features $\times$ samples) into $\M{W}$ (features $\times$ clusters) and $\M{H}$ (samples $\times$ clusters)
+        \item Approximate $\M{A}$ (features $\times$ samples) into $\M{W}$ (features $\times$ clusters) and $\M{H}$ (samples $\times$ clusters)
         \vfill
-        \item nonnegativity gives interpretability of $\M{W}$ and $\M{H}$ as clusters and cluster membership, respectively
+        \item Nonnegativity gives interpretability of $\M{W}$ and $\M{H}$ as clusters and cluster membership, respectively
         \vfill
-        \item applications in text document topic modeling and hyperspectral image segmentation, for example
+        \item Applications in text document topic modeling and hyperspectral image segmentation, for example
     \end{itemize}
@@
 \begin{itemize}
-	\item repeatedly use NMF with $k = 2$ to bipartition nodes in order to create a hierarchical tree of clusters
-	\item application: hyperspectral imaging
+	\item Repeatedly use NMF with $k = 2$ to bipartition nodes in order to create a hierarchical tree of clusters
+	\item Application data: hyperspectral image of DC National Mall
 \end{itemize}
@@
 \begin{column}{.4\textwidth}
-\GB{put bigger picture of DC mall (node 0) here, add pointers to leaves of tree for particular material (like grass or water)}
+    \centering
+    \includegraphics[height=0.6\textheight]{../data/DC/0.png}
 \end{column}
@@
 	\begin{adjustbox}{width=\textwidth}
-    		\import{..}{fig/dc}
+    		\import{..}{fig/dc_labels}
    	\end{adjustbox}
@@
 
-        \item Alternating Nonnegative Least Squares (ANLS)
+        \item Algorithm: Alternating Nonnegative Least Squares (ANLS)
         \begin{itemize}
@@
             \item 4 possibilities: both positive, one or other is zero, both zeroes 
-            \item quickly solve each and choose the optimal solution
+            \item quickly solve all possibilities and choose the optimal solution
         \end{itemize}
@@
 
-\begin{frame}{Parallel R2-NMF}
+\begin{frame}{Parallel Rank-2 NMF}
     \begin{itemize}
@@
 
-\begin{frame}{Bipartitioning with R2-NMF}
+\begin{frame}{Bipartitioning with Rank-2 NMF}
     \begin{itemize}
@@
 
-\begin{frame}{Hierarchical Clusterings using R2-NMF (HierNMF2)}
+\begin{frame}{Hierarchical Clusterings using Rank-2 NMF (HierNMF2)}
     \begin{itemize}
-        \item Split root node into two children with R2-NMF
+        \item Split root node into two children with Rank-2 NMF
         \item Choose next node to split based on scoring metric
         \begin{itemize}
-        		\item computing score requires pre-computing R2-NMF
+        		\item computing score requires pre-computing Rank-2 NMF
 	\end{itemize}
@@
         \begin{itemize}
-            \item smaller synthetic image classification dataset which has the same aspect ratio as SIIM-ISIC but fits in one node's memory
+            \item smaller synthetic dataset which has the same aspect ratio as SIIM-ISIC but fits in one compute node's memory
         \end{itemize}
@@
     \vfill
-    \begin{center}
     \scriptsize
     	All experiments performed on Summit at Oak Ridge Leadership Computing Facility
-    \end{center}
+	\begin{itemize}
+		\item each compute node has 21 available cores on each of 2 sockets
+	\end{itemize}
+
 \end{frame}
@@
 
-\begin{frame}{R2-NMF Strong Scaling and Time Breakdown}
+\begin{frame}{Rank-2 NMF Strong Scaling and Time Breakdown}
     \centering
@@
 \small
-    \item Nearly perfect strong scaling because time dominated by local matrix multiplication
+    \item Nearly perfect strong scaling because time dominated by local matrix multiplication (MatMul)
 \end{itemize}
@@
 \small
-    \item Scaling hindered by NNLS computations and communication of small subproblems at leaves of hierarchy
+    \item Reasonable scaling, though limited by NNLS computations and communication of small subproblems at leaves of hierarchy
 \end{itemize}
@@
 
+\begin{frame}{Comparison with Flat NMF}
+    \centering
+    \begin{columns}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\textwidth]{../plots/dc_rankscaling.pdf}
+            \caption{DC-HYDICE}
+            \end{figure}
+        \end{column}
+        \begin{column}{0.5\textwidth}
+            \begin{figure}
+            \includegraphics[width=\textwidth]{../plots/synthetic_rankscaling.pdf}
+            \caption{Synthetic}
+            \end{figure}
+        \end{column}
+    \end{columns}
+
+\vfill
+
+\begin{itemize}
+    \item Hierarchical clustering scales better than flat NMF with number of clusters $k$, potentially useful for flat clustering
+\end{itemize}
+
+\end{frame}
+
 \begin{frame}{Conclusions and Future Directions}
@@
 	\vfill
+	\item Possibilities for improving scalability
+	\begin{itemize}
 		\item 2D distribution of matrix $\M{A}$ could improve scalability but requires redistribution of $\M{A}$ for subtrees
-	\vfill
 		\item Parallelization across tree could improve scalability but requires careful load balancing and decentralizing split orders
+	\end{itemize}
 	\vfill
